### 2022-07-26 manu-draft.Rmd
title and abstract update

---
title: "Addressing climate change and development pressures in an urban estuary through habitat restoration planning"
shorttitle: "Tampa Bay watershed habitat restoration planning"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
    number_sections: false
    pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
always_allow_html: true
bibliography: refs.bib
author: 
  - Gary E Raulerson:
      email: gmangrove@gmail.com
      institute: [TBEP]
      correspondence: false
  - Douglas E Robison:
      email: DRobison@esassoc.com 
      institute: [ESA]
      correspondence: false
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Maya C Burke:
      email: mburke@tbep.org
      institute: [TBEP]
      corrspondence: false
  - Justin Saarinen:
      email: justinsaarinen@gmail.com 
      institute: [ESA]
      correspondence: false
  - Christine Sciarrino:
      email: CSciarrino@esassoc.com 
      institute: [ESA]
      correspondence: false
  - Edward T Sherwood:
      email: esherwood@tbep.org 
      institute: [TBEP]
      correspondence: false
  - David A Tomasko:
      email: dave@sarasotabay.org
      institute: [SBEP]
      correspondence: false
institute:
  - TBEP: Tampa Bay Estuary Program, St. Petersburg, Florida 33701 USA
  - ESA: Environmental Science Associates, Tampa, Florida 33609 USA
  - SBEP: Sarasota Bay Estuary Program, Sarasota, Florida 34236 USA
link-citations: true
---

# Abstract

Native habitats in Florida face dual pressures at the land-sea interface from urban development and sea-level rise associated with climate change. To address these pressures, restoration practitioners require robust tools that identify reasonable goals given historical land use trends, current status of the habitat mosaic, and anticipated future impacts from coastal stressors. A target-setting approach for native habitats was created for the Tampa Bay watershed that identifies current restoration opportunities and establishes short (2030) and long-term (2050) goals. The approach was informed through a three-decade habitat change analysis and over forty years of habitat restoration experience in the region. Restoration goals were defined based on what is possible today and the projected needs for the future, rather than attempting to replicate past ecological conditions. The new paradigm also accounts for the expected impacts of sea-level rise, climate change, and watershed development - stressors which are pervasive in Florida. The resulting habitat goals are spatially explicit with maps that identify remaining restoration and conservation opportunities, while also providing an approach for the entire watershed that targets subtidal, intertidal, and coastal uplands. This approach represents a general framework for addressing competing interests in planning that could be applied in other coastal settings where sustainable urbanization practices are desired to co-exist with natural environments.  Methods for repeatable analyses are also available using an open source workflow to update progress over time and for adoption by others.

### 2022-07-27 manu-draft.Rmd
added JS edit

@@
 The resulting habitat goals are spatially explicit with maps that identify remaining restoration and conservation opportunities, while also providing an approach for the entire watershed that targets subtidal, intertidal, and coastal uplands.
-This approach represents a general framework for addressing competing interests in planning that could be applied in other coastal settings where sustainable urbanization practices are desired to co-exist with natural environments.
+This approach represents a general framework to support coastal planning decisions that need to address competing interests and could be applied in other coastal settings where sustainable urbanization practices need to co-exist with natural environments.
 Methods for repeatable analyses are also available using an open source workflow to update progress over time and for adoption by others.

### 2022-07-28 manu-draft.qmd
back to qmd

--- 2022-07-27 manu-draft.Rmd
@@
 ---
+format: 
+  docx:
+    reference-doc: my_styles.docx
 title: "Addressing climate change and development pressures in an urban estuary through habitat restoration planning"
 shorttitle: "Tampa Bay watershed habitat restoration planning"
-output: 
-  bookdown::word_document2:
-    reference_docx: my_styles.docx
-    number_sections: false
-    pandoc_args:
-      - --lua-filter=templates/scholarly-metadata.lua
-      - --lua-filter=templates/author-info-blocks.lua
-always_allow_html: true
 bibliography: refs.bib
 author: 
-  - Gary E Raulerson:
+  - name: Gary E Raulerson
     email: gmangrove@gmail.com
-      institute: [TBEP]
-      correspondence: false
-  - Douglas E Robison:
+    affiliations: 
+      - ref: tbep
+    attributes:
+      corresponding: false
+  - name: Douglas E Robison
     email: DRobison@esassoc.com 
-      institute: [ESA]
-      correspondence: false
-  - Marcus W Beck:
+    affilations: 
+      - ref: esa
+    attributes:
+      corresponding: false
+  - name: Marcus W Beck
     email: mbeck@tbep.org
-      institute: [TBEP]
-      correspondence: true
-  - Maya C Burke:
+    affilations: 
+      - ref: tbep
+    attributes:
+      corresponding: true
+  - name: Maya C Burke
     email: mburke@tbep.org
-      institute: [TBEP]
-      corrspondence: false
-  - Justin Saarinen:
+    affilations: 
+      - ref: tbep
+    attributes:
+      corresponding: false
+  - name: Justin Saarinen
     email: justinsaarinen@gmail.com 
-      institute: [ESA]
-      correspondence: false
-  - Christine Sciarrino:
+    affilations: 
+      - ref: esa
+    attributes:
+      corresponding: false
+  - name: Christine Sciarrino
     email: CSciarrino@esassoc.com 
-      institute: [ESA]
-      correspondence: false
-  - Edward T Sherwood:
+    affilations: 
+      - ref: esa
+    attributes:
+      corresponding: false
+  - name: Edward T Sherwood
     email: esherwood@tbep.org 
-      institute: [TBEP]
-      correspondence: false
-  - David A Tomasko:
+    affilations: 
+      - ref: tbep
+    attributes:
+      corresponding: false
+  - name: David A Tomasko
     email: dave@sarasotabay.org
-      institute: [SBEP]
-      correspondence: false
-institute:
-  - TBEP: Tampa Bay Estuary Program, St. Petersburg, Florida 33701 USA
-  - ESA: Environmental Science Associates, Tampa, Florida 33609 USA
-  - SBEP: Sarasota Bay Estuary Program, Sarasota, Florida 34236 USA
-link-citations: true
+    affilations: 
+      - ref: sbep
+    attributes:
+      corresponding: false
+affiliations:
+  - id: tbep
+    name: Tampa Bay Estuary Program
+    city: St. Petersburg
+    state: Florida 
+    postal-code: 33701 
+    country: USA
+  - id: esa
+    name: Environmental Science Associates
+    city: Tampa
+    state: Florida
+    postal-code: 33609 
+    country: USA
+  - id: sbep
+    name: Sarasota Bay Estuary Program
+    city: Sarasota
+    state: Florida
+    postal-code: 34236
+    country: USA
 ---

### 2022-07-28 manu-draft.tex
back to qmd

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Addressing climate change and development pressures in an urban estuary through habitat restoration planning},
  pdfauthor={Gary E Raulerson; Douglas E Robison; Marcus W Beck; Maya C Burke; Justin Saarinen; Christine Sciarrino; Edward T Sherwood; David A Tomasko},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Addressing climate change and development pressures in an urban
estuary through habitat restoration planning}
\author{Gary E Raulerson \and Douglas E Robison \and Marcus W
Beck \and Maya C Burke \and Justin Saarinen \and Christine
Sciarrino \and Edward T Sherwood \and David A Tomasko}
\date{}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Native habitats in Florida face dual pressures at the land-sea interface
from urban development and sea-level rise associated with climate
change. To address these pressures, restoration practitioners require
robust tools that identify reasonable goals given historical land use
trends, current status of the habitat mosaic, and anticipated future
impacts from coastal stressors. A target-setting approach for native
habitats was created for the Tampa Bay watershed that identifies current
restoration opportunities and establishes short (2030) and long-term
(2050) goals. The approach was informed through a three-decade habitat
change analysis and over forty years of habitat restoration experience
in the region. Restoration goals were defined based on what is possible
today and the projected needs for the future, rather than attempting to
replicate past ecological conditions. The new paradigm also accounts for
the expected impacts of sea-level rise, climate change, and watershed
development - stressors which are pervasive in Florida. The resulting
habitat goals are spatially explicit with maps that identify remaining
restoration and conservation opportunities, while also providing an
approach for the entire watershed that targets subtidal, intertidal, and
coastal uplands. This approach represents a general framework to support
coastal planning decisions that need to address competing interests and
could be applied in other coastal settings where sustainable
urbanization practices need to co-exist with natural environments.
Methods for repeatable analyses are also available using an open source
workflow to update progress over time and for adoption by others.

\end{document}

### 2022-07-29 manu-draft.qmd
lua filters work with qmd, full affiliations working

--- 2022-07-27 manu-draft.Rmd
@@
 ---
+format: 
+  docx:
+    reference-doc: my_styles.docx
 title: "Addressing climate change and development pressures in an urban estuary through habitat restoration planning"
 shorttitle: "Tampa Bay watershed habitat restoration planning"
-output: 
-  bookdown::word_document2:
-    reference_docx: my_styles.docx
-    number_sections: false
-    pandoc_args:
-      - --lua-filter=templates/scholarly-metadata.lua
-      - --lua-filter=templates/author-info-blocks.lua
-always_allow_html: true
 bibliography: refs.bib
@@
   - SBEP: Sarasota Bay Estuary Program, Sarasota, Florida 34236 USA
-link-citations: true
+
+filters: 
+  - templates/scholarly-metadata.lua
+  - templates/author-info-blocks.lua
 ---

### 2022-08-04 manu-draft.qmd
intro w/ formatted references for FEE

@@
 bibliography: refs.bib
+csl: frontiers-in-ecology-and-evolution.csl
 author: 
@@
 
-# Abstract
+## Abstract
 
@@
 Methods for repeatable analyses are also available using an open source workflow to update progress over time and for adoption by others.
+
+## Introduction
+
+The health of estuarine systems, coastal habitats, and associated fauna and flora are inextricably linked to land uses and management throughout the watershed [@Yoskowitz15] These habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18].
+The Tampa Bay Estuary Program (TBEP) is one of 28 programs administered by the US Environmental Protection Agency (USEPA) under the National Estuary Program (NEP).
+In recognition of threats to habitats from development and climate change stressors, the TBEP and partners recently created the third iteration of a plan to establish targets and goals for habitat restoration within the Tampa Bay watershed.
+As an NEP, the program has guided regional environmental restoration initiatives for the estuary since 1991.
+The methodologies used in the creation of a 2020 Habitat Master Plan Update for Tampa Bay [@Robison20] are highly transferable to both coastal and non-coastal systems.
+
+## Development, climate change, and coastal squeeze
+
+Development causes multiple perturbations within any watershed.
+The accumulated impacts of construction and associated infrastructure remove or substantially modify existing habitats, and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14].
+Given the degree to which the Tampa Bay watershed has been urbanized, the synergistic effects of continued coastal development and future climate change are primary concerns for maintenance of estuarine and coastal habitat health.
+Observed and potential adverse effects of climate change and sea level rise on marine and estuarine ecosystems are well-documented [@Scavia02; @Spalding07; @Titus09].
+With regard to estuarine habitats, the primary concerns are that sea level rise is now occurring at such a rapid rate that the landward migration of tidal wetlands in response cannot keep pace; or that the upland slope has already been lost to urban development and hardening, leaving no place for tidal wetland migration [@Titus09].
+Geological, physical and chemical changes could include alterations in sediment deposition and erosion patterns, micro-topography, and water quality [@Arnell15; @Whitehead09].
+
+Related changes in habitats in response to climate change (sea level rise and warming) include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].
+Species dependent upon these habitats will be forced to change use patterns or adapt to the new conditions [@Iwamura13].
+For example, while black needle rush (*Juncus roemerianus*) can tolerate a wide salinity range [@Eleuterius79; @Stout84], the largest remaining *J.
+roemerianus* marshes in Tampa Bay are located in the lower-salinity reaches of tidal rivers and creeks.
+The greatest extents of those marshes occur in river systems where their upstream extent is constrained by impoundments for public water supplies.
+Spatial restriction in these hydrologically truncated rivers may make these marshes particularly vulnerable to "pinching out", as upstream migration in response to sea level rise will be cut off by anthropogenic barriers.
+Similarly, landward migration of salt barrens (high marsh areas in Tampa Bay) in response to sea level rise will be restricted by the filling and hardening of coastal uplands associated with existing or future urban development.
+
+## Need for a paradigm shift
+
+The synergistic effects of development and climate change diminish the available space for future restoration in urbanizing estuaries, thereby impacting the variety of ecosystem services provided by these habitats and the wildlife they support [@Enwright15].
+Given projected habitat losses without intervention efforts and the limited resources (including time, land, funding, and labor force) available, it is important to appropriately and realistically site restoration projects to increase the likelihood of success.
+To achieve this objective, a new restoration approach on a broad watershed scale will be implemented.
+
+Previously, a "retrospective" approach to setting habitat protection and restoration targets in Tampa Bay was employed [@Lewis96; @Robison10; @Cicchetti11; @Russell15].
+Under this paradigm, priority was given to restoration activities focused on habitat types, important for a suite of ten estuarine faunal guilds, that were disproportionately lost or degraded compared to a benchmark period.
+Primary criticisms of this approach included a lack of consideration for future sea level rise and other climate change factors [@Yoskowitz15], use of expanded and different habitats outside the Tampa Bay watershed by the faunal guilds [@Robison10], lack of attention to upland or freshwater wetland habitats, and little recognition of other stressors such as land development trends or actual available space for restoration efforts.
+
+Past approaches for guiding restoration planning have been successfully used in other contexts, but they do not fully balance competing needs.
+For example, an integrated watershed approach [@EPA96] has been utilized since the early 1990s to diagnose and manage water quantity and quality problems that have contributed to seagrass restoration in the system.
+Additionally, the habitat mosaic approach [@Henningsen05] of including multiple habitat types within restoration projects is recognized as necessary in Tampa Bay [@Hughes11] and elsewhere to allow for ecosystem state changes in response to different environmental pressures [@Duarte09; @Palmer09].
+
+Adaptive management [@Holling78; @Gregory06] components have been increasingly used to address challenges of sea level rise, climate change, and development stressors, including monitoring to identify critical restoration decision points and needed intervention with contingency plans.
+Rising sea levels and temperatures and altered rainfall patterns are causing observable changes to habitats on a global scale [@Cavanaugh14; @Garner15; @Yoskowitz15], including within Tampa Bay [@Price17], and those changes are expected to become more pronounced over the next several decades [@Sheehan16; @Nerem18].
+
+## Updating the approach
+
+The new approach integrates the whole watershed, addresses historical changes, focuses on trajectories that have occurred during more contemporary time periods, and considers both current and future stressors -- particularly land development and sea level rise.
+There is relatively consistent extent and distribution data for most Tampa Bay habitats of interest (1988 to 2018), representing a time period when federal, state and local regulations were in effect and regional impacts from climate change are documented [@Raabe12; @Cavanaugh14].
+This approach establishes a broader framework that guides both watershed-level habitat master planning and site-level restoration design activities and incorporates applicable elements of the other habitat restoration paradigms discussed above [@Palmer09].
+Our broader framework for guiding restoration activities includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in acreage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories.
+
+With regard to habitat restoration projects, the design approach must envision not only what is possible today, but also what the coastal landscape will look like in 50 years and beyond.
+Design features should continue to use the historical "habitat mosaic" approach, but should also include coastal upland features that accommodate tidal inundation and the landward advance of emergent tidal wetlands [@Enwright16].
+
+## References

### 2022-08-04 manu-draft.qmd
all original text added and references cited with cls

@@
 
+## Methods
+
+### Study area
+
+Tampa Bay is a large open water estuary (open water area approximately 983 km^2^) on the west-central coast of Florida (Fig 1).
+Its watershed encompasses another 5,872 km^2^, for a total combined area of approximately 6,855 km^2^.
+It is subtropical, and within the current (2020) ecotone for mangrove and salt marsh habitats.
+The upstream watershed includes multiple habitats, including pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands.
+The watershed is heavily developed with an estimated (2019) population of 3.3 million people in the four counties that comprise most of the watershed [@Rayer20]*.* Numerous anthropogenic changes have been made to the natural systems within and surrounding Tampa Bay, including direct removal of habitat (including dredge and fill of bay bottom), alteration of hydrology, and destruction and fragmentation of habitat from development.
+
+### Habitats of Tampa Bay
+
+Addressing the full suite of habitats in a watershed is now recognized as critical for large-scale restoration planning efforts [@Palmer09; @Lamb18]**.** The major habitat types of Tampa Bay can be described and organized relative to tidal influence and location in the watershed.
+Subtidal habitats include those that are submerged all or most of the time; emergent tidal wetlands include those that are submerged during high tides but exposed during low tides; and supratidal habitats include those that occur above the high tide line.
+
+Habitats generally described as 'subtidal' include hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13].
+Mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15] are classified as emergent tidal wetlands.
+For the purposes of this planning effort, supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* As discussed below, uplands are sub-divided into coastal and non-coastal uplands, based on location relative to the 5-foot contour.
+
+### Approach
+
+Recommended targets were created stepwise with geospatial analyses: 1) quantifying habitat status and historical trends; 2) synthesizing historic habitat restoration efforts; and 3) identifying and defining remaining 'opportunities' for restoration that integrated results from the first two analyses.
+Final development of restoration targets and goals integrated strata and opportunity area information with an analysis of past restoration implementation to use previous completed projects as a guide to a feasible amount of restoration that could be conducted by partners.
+This approach will ensure that the estuarine-dependent species and faunal guilds throughout the watershed, as defined in the original approach, will continue to be supported.
+
+### Habitat status and trends
+
+For the majority of subtidal, intertidal and supratidal habitats, primary data derived from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD) were utilized.
+However, to address data gaps for some habitats, results from special studies were integrated with the primary data sources.
+These included hard bottom [@Kaufman17; @CSA19], dredged holes [@Griffen05; @Raulerson19] and oyster habitat [@OKeife06]*.*
+
+The source data used to estimate the most current coverage of seagrasses, tidal flats, and oysters was the *Seagrass in 2018* geospatial database [@SWFWMD19].
+The bi-annual seagrass monitoring program was initiated in 1988 under SWFWMD's Surface Water Improvement and Management program [@Sherwood17; @Tomasko20]*.* SWFWMD has estimated oyster bed coverage as part of this program since 2014.
+
+The source data used to estimate and map trends in development, emergent tidal wetlands, freshwater wetlands, and native upland habitats was the SWFWMD Land Use Land Cover (LULC) series geospatial database [@SWFWMD18].
+This comprehensive database classifies the land use and cover types (natural and developed) pursuant to the Florida Land Use Cover and Forms Classification System [@FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually.
+While the photointerpretation of specific freshwater wetland types is often very difficult, it is possible to accurately distinguish forested wetlands from non-forested wetlands.
+Therefore, for this analysis, all applicable FLUCCS codes representing the suite of natural freshwater wetlands were combined within those two classifications.
+Similarly, within the target- and goal-setting exercise, uplands were combined in one classification.
+These classifications were reported for each mapping exercise conducted every 2-3 years from the start of the program 1990 through 2017, the most recent year with available data.
+
+To quantify the extent of tidal creeks, GIS data [@JEI16; @JEI20] was clipped to the Florida Department of Environmental Protection stream segments that were classified as estuarine.
+
+### Restoration database
+
+We quantified past restoration efforts in each of the major habitat types to guide decisions on future targets and goals.
+Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years were compiled, reviewed, and consolidated into a single, consistent geospatial database.
+Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act (GPRA) reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the TBEP.
+Primary data collected included project name, year, description, lead partner, size (area or length), and latitude and longitude.
+Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and documenting the knowledge of local professionals.
+Living shoreline projects, including seawall enhancements and oyster reef modules, were inventoried separately.
+
+### Creating and combining opportunity layers
+
+Three strata, distinct geographic breakpoints in the watershed where different habitat management and restoration activities could take place, were created to establish the process for target and goal-setting within this framework.
+The first is the coastal stratum, which extends from the local Mean Lower Low Water elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation by 2070.
+The coastal stratum is the zone where emergent tidal wetland restoration would be conducted and includes low lying coastal uplands which serve as important tidal wetland buffers that will be critically important in the future as lands reserved to accommodate ideal wetland migration in response to sea level rise.
+The river floodplain stratum includes all hydrologically contiguous forested and non-forested wetlands within the river and stream corridors of the Tampa Bay watershed.
+Floodplain corridors provide vital watershed functions including fish and wildlife habitat and migratory pathways, floodwater attenuation and storage, erosion control, and delivery of complex organic matter to the estuarine food web.
+Finally, the upland stratum encompasses those areas outside of the coastal and river floodplain strata, including native upland habitats as well as hydrologically isolated wetlands.
+These habitats provide important aquifer recharge and wildlife habitat functions.
+
+Opportunity areas, defined here as locations where habitat protection and restoration activities are possible, and where they should best be focused to attain defined targets, were also analyzed.
+The definition and mapping of opportunity areas is necessary to quantify the "restoration potential" for a particular habitat type, which is a measure of what is actually possible under current and future projected conditions.
+The most appropriate opportunity areas are generally not developed and located on existing public lands or areas identified for acquisition.
+
+The Land Use Land Cover 2017 geospatial database (Table 1) was used as the baseline for cataloguing existing and opportunity areas for intertidal and supratidal habitats in the Tampa Bay watershed.
+All FLUCCS classification codes were placed into one of three categories.
+First, native habitats cover the full range of natural plant communities and other habitats that are endemic to the Tampa Bay watershed, and were further grouped into three major habitat types (tidal wetlands, freshwater wetlands, and uplands).
+Second, restorable habitats include existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through the restoration of more natural hydrology, soils strata, and/or topography.
+Third, existing development includes developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration activities.
+
+Layers for existing public lands and parcels targeted for acquisition were compiled by combining data from the Florida Natural Areas Inventory (Table 1), consulting staff from various federal, state and local entities, and inventorying conservation and drainage easements data.
+
+Compared to vegetation communities, soil characteristics typically change slowly (e.g., decades to centuries) in response to hydrologic impacts, unless physically disturbed [@Osland12; @Stockmann14].
+Therefore, soils distributions can be used to generally represent historical habitat distribution, and can be used to provide generalized restoration guidelines (e.g., tidal wetlands, freshwater wetlands, and native uplands).
+@Ries14 created a soils suitability analysis for wetland mitigation and restoration sing data from the USDA Web Soil Survey (Table 1) and classified all soils in the Tampa Bay soils into one of three categories (xeric, mesic, and hydric).
+The mesic and hydric categories were combined to represent wetland restoration potential, while the xeric category was used to represent upland restoration potential.
+
+A distinction was made between tidal and freshwater wetland restoration potential by intersecting the combined mesic and hydric soils polygons with the coastal stratum.
+Mesic or hydric soils that occur below the 5-foot contour were classified as having tidal wetland restoration potential, while mesic or hydric soils occurring above the 5-foot contour were classified as having freshwater wetland restoration potential.
+
+Within the Tampa Bay region, lands adjacent to waters with average salinity values greater than 18 psu are considered most appropriate for higher salinity mangrove/salt barren restoration, while lands adjacent to waters with average salinity values less than 18 psu are considered most appropriate for lower salinity salt marsh (*Juncus* spp.) restoration [@Eleuterius79].
+To estimate the relative restoration potential of mangrove/salt barrens and salt marshes, a regional long-term water quality data set was used to create salinity isohalines, which was then binned into two salinity categories: greater and less than an annual mean of 18 psu (Table 1).
+
+All acquired and created GIS layers (Table 1) were converted to 10m x 10m raster data sets and delimited using a previously established watershed boundary.
+The reservation, restorable, conservation, acquisition, soils, and salinity isohaline layers were then merged to create a comprehensive set of maps and matrices of all possible combinations of opportunity areas.
+Within an ArcGIS Pro 2.x GIS environment, the SWFWMD 2017 FLUCCS and 2018 submerged data was classified and extracted to the study area into described categories.
+The dataset was then spatially assigned to existing and proposed restoration areas, reservation areas, soil type, and salinity level.
+
+## Results
+
+### Habitat status and trends
+
+#### Subtidal habitats
+
+While no trend information is available, the best cumulative estimate of natural hard bottom extent in Tampa Bay is 171 ha (Table 2).
+Oyster bars covered 69 ha (Table 2) in 2018, and a 30% (6 ha) increase reflected since mapping began in 2014 (Table 3a) probably represents improved ground-truthing and photointerpretation of oyster bar signatures from aerial photography.
+Twelve artificial reefs in Tampa Bay are managed by Hillsborough, Manatee, and Pinellas Counties.
+Surface area estimates were not available for the Manatee and Pinellas County reefs, but assuming an average size of 4.2 ha, based on the Hillsborough County reefs, the total coverage of artificial reefs in Tampa Bay is estimated to be approximately 67 ha (Table 2).
+
+Total seagrass coverage has increased by 7,027 ha (75%) during the 30-year period of record (Table 3a), and the most current (2018) estimate of total seagrass meadow coverage in Tampa Bay is 16,452 ha.
+The 2018 coverage of tidal flats and sand other than beaches in Tampa Bay was 6,564 ha (Table 2).
+A decrease of 4,496 ha (55%) during the 30-year period of record (Table 3a) is associated with the expansion of seagrass to previously non-vegetated bottom area.
+
+#### Intertidal habitats
+
+Between 1990 and 2017, the suite of emergent tidal wetlands (mangroves, salt barrens, and salt marshes) experienced a net gain of 725 ha (10%, Table 3b).
+The current estimate of the extent of mangrove forests in Tampa Bay is 6,192 ha.
+Mangrove forest coverage increased by 684 ha (12%).
+Salt marshes in Tampa Bay cover 1,844 ha, and coverage increased by 30 ha (2%).
+However, from 1990 to 2017, it is estimated that a net area of 219 ha of salt marshes converted from salt marsh to mangrove habitat (<https://tbep-tech.github.io/landuse/>).
+The 2017 estimate of the extent of salt barrens in Tampa Bay is 201 ha, and coverage has increased by 14 ha (7%) during the 27-year period of record (Table 3b).
+Based on GIS data from Janicki Environmental and Mote Marine Laboratory [@JEI16; @JEI20], the extent of tidal creek habitat in the Tampa Bay watershed is approximately 623 km (Table 2).
+No trend analysis is available.
+
+#### Supratidal habitats
+
+As determined in the 2017 land use/land cover update, the most current estimate of the extent of non-coastal native upland habitats in the Tampa Bay watershed is 56,899 ha, and the extent of "coastal uplands" (defined as below the 5-foot contour) in the Tampa Bay watershed is 1,465 ha (Table 2).
+Over the 27-year period of record, the suite of native upland habitats has experienced a net loss of 37,051 ha (39%, Table 3b).
+
+The 2017 extent of freshwater wetlands in the Tampa Bay watershed was 88,917 ha (Table 2).
+Of this total, forested freshwater wetlands comprised 61,565 ha (69%), while non-forested freshwater wetlands comprised 27,351 ha (31%).
+From 1990-2017, the suite of freshwater wetlands experienced a net gain of 2,444 ha (3%, Table 3b).
+There has been a 5,335 ha (24%) increase in vegetated non-forested freshwater wetlands since 1990, while forested freshwater wetlands have decreased by 2,891 ha (4%).
+
+### Habitat restoration
+
+A total of 460 projects were documented between 1971 and 2019 (Table 4), addressing the full range of habitat types, including Estuarine (n=228), Freshwater (n=53), Uplands (n=119), and Mixed (n=60).
+A total of 1,978 ha have been restored, and 12,930 ha and 42.8 km of linear projects were enhanced during the time period.
+Forty lead partners were documented as responsible for the projects, although some of these lead partners are departments within the same overall agency.
+Eighty-nine living shoreline projects, seawall enhancements, and oyster reef module installations along shorelines were inventoried, with a linear footprint of 18.2 km.
+
+### Habitat restoration opportunities
+
+In 2017, 1,555 km^2^ of land (26.5%) in the Tampa Bay watershed above the MLLW line was classified as natural and 2,144 km^2^ (37%) was considered restorable.
+Developed areas in the Tampa Bay watershed encompassed 2,172 km^2^ (36.5%) of the 5,872 km^2^ watershed area.
+Between 2014 and 2017, the developed footprint increased by seven percent (7%, Fig 2).
+
+The Tampa Bay watershed included a total of 1,260 km^2^ of existing conservation lands, either publicly owned or in conservation easements.
+However, excluding subtidal areas owned by the State of Florida results in a total of 816 km^2^ of conservation lands, about 13.9% of the watershed area occurring above the MLLW line.
+(Fig 3).
+Proposed conservation lands in the Tampa Bay watershed total 1,254 km^2^.
+The mapped proposed conservation lands generally link eastward and provide wildlife connectivity to the larger-scale Florida Wildlife Corridor (Fig 3).
+
+Reservation lands, from the MLLW line landward to elevation 5-feet (NAVD 88), currently (2017) include 4,622 ha proposed for conservation in the Tampa Bay watershed, with 3,303 ha of native habitats and 1,319 ha of restorable habitats (Fig 1).
+
+Xeric soils were roughly aggregated in an east-west band through the middle of the watershed (Fig 4). Approximately 2,107 km2 was classified as xeric, while 3,760 km2 was classified as mesic/hydric.
+
+Integration of all of the opportunity layers provides a summary of the available restoration for the TBEP habitats of interest (Table 5).
+The "Native Habitats" columns show the total current extent as well as the portion of the current extent occurring on existing conservation lands and proposed conservation lands, respectively.
+The "Restorable Habitats" columns show the total restoration opportunity as well as the portion of the total restoration opportunity on existing and proposed conservation lands.
+The majority of restoration opportunities on existing conservation lands are for native uplands and freshwater wetlands (Table 5).
+However, there are approximately 627 ha of emergent tidal wetland restoration opportunities on existing conservation lands, including 530 ha applicable to higher salinity mangrove forests and salt barrens, and 16 ha applicable to lower salinity salt marsh (e.g., *Juncus roemerianus*) restoration and creation.
+
+The best estimates of total restoration opportunities for urban shorelines and tidal tributaries are provided by the Tampa Bay Living Shoreline Suitability Model (LSSM) prepared by the Florida Fish and Wildlife Conservation Commission [@Boland18].
+There is 2,566 km of shoreline in Tampa Bay, and approximately 33% is recommended for living shoreline enhancement.
+
+### Establishment of goals and targets
+
+Recommended targets were based on habitat status and trends, habitat restoration data, identified restoration opportunities, and current and anticipated trends in development, available funding, and regulations [@Hilderbrand05; @Hobbs07].
+
+The targets and goals (Table 6) identify where the 2017 CCMP Bay Habitat goals and strategies and the 2021-2025 Strategic Plan [@Burke20] thriving habitats and abundant wildlife programmatic priorities can be implemented.
+However, it also recognizes that the identified habitat protection and restoration areas will change over time, and will be revisited on a 10-year recurring cycle.
+A 30-year planning horizon (2050) is also identified based upon sea level rise projections developed specifically for Tampa Bay [@Burke19].
+The coastal stratum (from the existing mean low water line to the approximate 5-foot contour) is projected to directly experience the effects of sea level rise by 2050, and is the primary focus area for coastal habitat protection and restoration activities.
+Land acquisition or protection (through conservation easements or other mechanisms) will be needed to ensure completion of targets and goals for both salt marsh and upland habitats (Fig 3).
+
+Targets that maintain current coverage ("hold-the line strategies") were identified for habitats that appear to currently be sustained at acceptable levels.
+Evolving information such as an Oyster Habitat Suitability Index [@Boswell12] and ongoing mapping exercises will be used to identify optimal locations to conduct restoration activities that help achieve targets and goals.
+Coordination with the establishment of state-mandated minimum flows [@Munson07] will be necessary for restoration and maintenance of low-salinity salt marsh habitats that will experience higher salinities and rapid transition to mangroves under existing sea level rise scenarios [@Sherwood14; @Geselbracht15].
+
+## Discussion
+
+Because of multiple stressors such as encroaching development and climate change, habitat protection and restoration priorities should be tempered and "reality tested" by what is actually possible today, and what is possible in the future.
+Many native and potentially restorable habitats are limited, and there will always be restrictions on the financial resources that can be dedicated to public conservation land acquisition and habitat restoration activities.
+
+This replicable method for setting restoration targets and goals provides a systematic attempt to identify habitat protection and restoration targets that are based on what is actually achievable within those limitations.
+It focuses on existing opportunities for all habitat types, and what is realistically possible in the future, rather than attempting to mimic previous ecological conditions.
+
+### Habitat trends
+
+When viewed as a whole, the most significant and meaningful trends in the TBEP habitats of interest over the periods of record examined include: 1) the 75 percent gain in seagrasses since 1988; 2) the slight gains in emergent tidal wetlands (10%) and freshwater wetlands (2%) since 1990; and 3) the 39% loss in native upland habitats since 1990.
+
+The intertidal zone in Tampa Bay is currently experiencing dynamic change, driven by sea level rise and climate change, whereby mangrove forests are outcompeting salt marshes and salt barrens for the available niche space.
+Without increasing the total area of the intertidal zone, restoring a greater coverage of mangroves would reduce the niche space available for salt marshes and salt barrens.
+This phenomenon has been observed throughout the Gulf of Mexico, and has been attributed to both climate change (e.g., fewer freeze events) and sea level rise [@Comeaux12].
+
+The observed gains in wetlands (Table 3b) are likely a reflection of: 1) the effectiveness of state and federal wetland regulatory programs; and 2) the cumulative gains resulting from, primarily, publicly-funded habitat restoration projects (Table 4).
+Minor gains in some emergent tidal wetlands (e.g., salt barrens) may also be a reflection of the landward expansion of the complex suite of these habitats associated with climate change and sea level rise.
+Gains in vegetated non-forested freshwater wetlands are related to the clearing of forested wetlands followed by the creation of herbaceous mitigation areas and stormwater systems.
+
+The decrease in native uplands (Table 3b) is the result of continued development in the Tampa Bay watershed, combined with the lack of regulatory protection of native uplands.
+Attaining the target and goal will require concerted restoration of native upland habitats on existing conservation lands, as well as new conservation lands to offset the continued loss of these habitats to development, and amendments to existing planning, zoning and land development policies or regulations.
+While federal and state regulations related to listed species management impart some protection to certain rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats, such as pine flatwoods, are left largely unprotected.
+Unless local governments in the Tampa Bay watershed improve local protections for native uplands, such as strengthening language within comprehensive plans and development ordinances, this trend will likely continue.
+
+### Habitat restoration
+
+Increases observed in tidal and freshwater wetlands are primarily due to publicly-funded habitat restoration projects, state and federal wetland regulatory programs, and to a lesser extent, regulatory mitigation.
+While restoration activities date to 1971, few projects were completed prior to 1990, and from 1990-2010, an annual mean of 68 ha of habitat was restored.
+Over the past decade (2010-2019), this rate of restoration project completion has increased to over 81 ha/yr.
+Assuming that funding levels remain in the same range as the past decade, this annual mean can be used to set reasonable limits on restoration potential and targets.
+
+While existing development areas are not considered feasible for major habitat restoration activities at this time, there are many opportunities to enhance and restore habitat functions and improve coastal resilience in urbanized locations.
+Examples include the construction of living shorelines, placement of submerged habitat modules along developed urban shorelines and seawalls, and creation of backyard habitats.
+Tidal tributary restoration could also entail improvements including removal of salinity barriers and filling of dredged channel sections with low dissolved oxygen.
+
+Four major types of disturbed sites around the Tampa Bay coastline have been identified as priority estuarine habitat restoration sites by TBEP stakeholders over the past two decades, including dredged holes, filled and spoil disposal areas, abandoned aquaculture ponds; and coastal borrow pits and stormwater ponds.
+Substantial opportunities also exist for upland restoration on reclaimed mined lands within the watershed.
+
+There is a general consensus among restoration practitioners and natural resource managers that habitat restoration and management is most cost-effective on publicly-owned conservation lands and provides long-term benefits.
+Furthermore, given current development trends in the Tampa Bay watershed, public acquisition of remaining critical lands (e.g., coastal uplands, river floodplain wetlands) is a high priority, and some restoration targets (e.g., salt marshes) will not be feasible without additional public acquisition or public-private partnerships [@Holl01; @Benson18].
+Therefore, varied approaches to leverage resources, including traditional grants, partner funding, and use of volunteers for habitat restoration are recommended to maximize the potential for successful target and goal achievement.
+
+### Rolling easements, mitigation, and restoration consortium
+
+As discussed, land acquisition for coastal habitat restoration must prioritize adjacent low-lying coastal uplands to serve as buffers to accommodate future landward migration of tidal wetlands in response to sea level rise.
+Where public acquisition is not possible, other conservation mechanisms need to be explored.
+Coastal setbacks, buffers, or public easements are traditionally used to restrict development within a given distance from the shoreline.
+A rolling easement is a dynamic mechanism that "rolls" landward as sea levels rise and cause tidal encroachments onto low-lying coastal uplands [@Titus09].
+The application of rolling easements in Tampa Bay could disincentivize more intense urban development (e.g., discourage up-zoning) of low-lying coastal uplands that may be currently in less intense agricultural or recreational (e.g., golf courses) land uses.
+Under a rolling easement, landowners would be able to maintain current economic uses, while "reserving" such lands to accommodate tidal wetland migration with advancing sea level rise.
+
+Wetland impacts and associated compensatory mitigation projects authorized under wetland regulatory programs have historically been conducted independent of watershed-level planning and monitoring processes.
+This disconnect has contributed to fragmented implementation and inconsistent compliance monitoring of mitigation projects, as well as historically poor documentation of wetland losses and gains in the Tampa Bay watershed.
+However, if properly focused and comprehensively coordinated, compensatory mitigation activities could significantly contribute to the attainment of wetland habitat restoration goals and targets for the Tampa Bay estuarine system and its contributing watershed.
+
+TBEP will form a public-private partnership to provide the framework for the development of a coordinated approach linking regulatory (compensatory mitigation) and resource management (publicly funded habitat enhancement, restoration, and establishment) programs in the watershed.
+The creation of the Habitat Management Consortium is expected to provide benefits such as optimizing and improving the cost-effectiveness of habitat protection, restoration and mitigation activities in the watershed.
+
+### Conclusion
+
+The establishment of targets and goals considering climate change, development, land availability, and past achievements, expands the restoration palette to a more comprehensive list of habitats within the system.
+If successfully implemented, the 2050 goals would total over 4,000 ha of habitat restoration throughout the Tampa Bay watershed.
+Land acquisition will be an important component of successful completion of different targets and goals, including the salt marsh restoration target for the first ten years.
+Land acquisition will also provide new opportunities for outdoor access, given that these projects often have a public recreation component.
+
+Our new approach will continue to engage multiple partner agencies, non-governmental organizations, and private citizens in the successful implementation of the restoration plan.
+Emphases will include recognition of land types particularly vulnerable to climate change or development stressors and these needs will be communicated to restoration partners.
+Consistent education, targeted funding opportunities, and reporting will also ensure that these newly established targets and goals lead to successful restoration projects, land acquisition, and enhanced ecosystem services.
+
 ## References

### 2022-08-04 R/dat_copy.R
copying data and R funcs from hmpu-workflow, scoping out tables/figs

@@ -1,0 +1,15 @@
+# get data files from hmpu-workflow repo (local only)
+
+library(here)
+
+
+# get data ------------------------------------------------------------------------------------
+
+fls <- list.files(here('../hmpu-workflow/data'), full.names = T, pattern = '\\.RData$|\\.csv$')
+
+file.copy(fls, to = here('data/'), overwrite = T)
+
+
+#  get R functions ----------------------------------------------------------------------------
+
+file.copy(here('../hmpu-workflow/R/funcs.R'), here('R/'))

### 2022-08-04 R/funcs.R

@@ -1,0 +1,1031 @@
+# extract legend from existing ggplot object
+g_legend <- function(a.gplot){
+  tmp <- ggplot_gtable(ggplot_build(a.gplot))
+  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
+  legend <- tmp$grobs[[leg]]
+  return(legend)
+}
+
+# estimate lulc area in acres for available HMPU targets 
+#
+# lulcin in lulc rdata file for a given year
+# coastal is coastal stratum from strats
+# flucss is lookup table
+# sumout logical if summary of acreage returned, otherwise
+lulc_est <- function(lulcin, coastal, fluccs, sumout = T){
+
+  # FLUCCS codes to remove, these are all subtidal and irrelevant for lulc
+  # in order: bays and estuaries, major bodies of water, gulf of mexico, tidal flats, oyster bars, sand other than beaches submerged
+  # patchy seagrass, continuous seagrass, attached algae, hardbottom x 6
+  cds <- c(5400, 5700, 5720, 6510, 6540, 7210, 9113, 9116, 9121, 9510, 9511, 9512, 9513, 9514, 9515)
+
+  # lulc area, all categories
+  # remove open water and subtidal habitats (open water area changes between layers and subtidal not consistently collected)
+  out <- lulcin %>%
+    filter(!FLUCCSCODE %in% cds) %>% 
+    add_coast_up(coastal, fluccs) 
+  
+  if(!sumout)
+    return(out)
+  
+  out <- out %>% 
+    mutate(
+      Acres = st_area(.),
+      Acres = set_units(Acres, acres), 
+      Acres = as.numeric(Acres)
+    ) %>% 
+    st_set_geometry(NULL) %>%
+    group_by(HMPU_TARGETS) %>%
+    summarise(Acres = sum(Acres), .groups = 'drop') %>% 
+    arrange(HMPU_TARGETS)
+  
+  return(out)
+  
+}
+
+# estimate subtidal area in acres for available HMPU targets 
+#
+# subtin in seagrass rdata file for a given year
+# flucss is lookup table
+subt_est <- function(subtin, fluccs, sumout = TRUE){
+  
+  # subtidal area, all categories
+  out <- subtin %>%
+    mutate(
[…]

### 2022-08-04 manu-draft.qmd
copying data and R funcs from hmpu-workflow, scoping out tables/figs
added gis workflow figure

@@
   - templates/author-info-blocks.lua
+
+execute: 
+  echo: false
+  warning: false
 ---
+
+```{r, results = 'hide'}
+library(knitr)
+library(here)
+```
 
@@
 ### Creating and combining opportunity layers
+
+Figure \@ref(fig:gisworkflow).
 
@@
 
+## Figures
+
+```{r gisworkflow}
+#| fig.cap: 'Spatial analysis workflow used to identify opportunity areas for habitat conservation and restoration in the Tampa Bay watershed.
+Workflows are divided into binning of land use/land cover categories into relevant habitat types and spatial overlay of datasets to identify the opportunity areas.'
+knitr::include_graphics(here('figs/gisworkflow.png'))
+```
+
+Figure 2: Land use change as Sankey
+
+Figure 3: Subtidal habitat change as Sankey
+
+Figure 4: Map of opportunity assessment
+
+Figure 5: Map of restoration potential
+
+## Tables
+
+Table 1: Land use change table
+
+Table 2: Subtidal change table
+
+Table 3: Summary of the opportunity assessment analysis
+
+Table 4: Targets
+
 ## References

### 2022-08-05 manu-draft.qmd
adding alluvial plots

--- 2022-08-04 manu-draft.qmd
@@
 
+Figure @fig-gisworkflow
+
 Three strata, distinct geographic breakpoints in the watershed where different habitat management and restoration activities could take place, were created to establish the process for target and goal-setting within this framework.
@@
 
+Figure @fig-subtchgdatalluv
+
 While no trend information is available, the best cumulative estimate of natural hard bottom extent in Tampa Bay is 171 ha (Table 2).
@@
 #### Intertidal habitats
+
+Figure @fig-chgdatalluv
 
@@
 
+## Figures
+
+![Spatial analysis workflow used to identify opportunity areas for habitat conservation and restoration in the Tampa Bay watershed.
+Workflows are divided into binning of land use/land cover categories into relevant habitat types and spatial overlay of datasets to identify the opportunity areas.](figs/gisworkflow.png){#fig-gisworkflow fig-align="center"}
+
+![Change analysis of habitat categories in the subtidal strata of the Tampa Bay watershed.
+The left column shows relative areas in 1988 and the right column shows relative areas in 2018 for each habitat category.
+The grey lines show the proportional change of each habitat category between the years.](figs/subtchgdatalluv.png){#fig-subtchgdatalluv fig-align="center"}
+
+![Change analysis of habitat categories in the intertidal and supratidal strata of the Tampa Bay watershed.
+The left column shows relative areas in 1990 and the right column shows relative areas in 2017 for each habitat category.
+The grey lines show the proportional change of each habitat category between the years.](figs/chgdatalluv.png){#fig-chgdatalluv fig-align="center"}
+
+Figure 4: Map of opportunity assessment
+
+Figure 5: Map of restoration potential
+
+## Tables
+
+Table 1: Land use change table
+
+Table 2: Subtidal change table
+
+Table 3: Summary of the opportunity assessment analysis
+
+Table 4: Targets
+
 ## References

### 2022-08-05 R/dat_copy.R

@@ -2,7 +2,6 @@
 
 library(here)
 
-
 # get data ------------------------------------------------------------------------------------
 
 fls <- list.files(here('../hmpu-workflow/data'), full.names = T, pattern = '\\.RData$|\\.csv$')
@@ -12,4 +11,4 @@ file.copy(fls, to = here('data/'), overwrite = T)
 
 #  get R functions ----------------------------------------------------------------------------
 
-file.copy(here('../hmpu-workflow/R/funcs.R'), here('R/'))
+file.copy(here('../hmpu-workflow/R/funcs.R'), here('R/'), overwrite = T)

### 2022-08-05 R/dat_proc.R

@@ -1,0 +1,185 @@
+library(sf)
+library(tidyverse)
+library(here)
+library(units)
+
+source(here('R/funcs.R'))
+
+# LULC change analysis ----------------------------------------------------
+
+data(coastal)
+
+# get final year
+maxyr <- 2017
+
+load(file = here(paste0('data/lulc', maxyr, '.RData')))
+maxdat <- maxyr %>%
+  paste0('lulc', .) %>%
+  get %>%
+  lulc_est(coastal, fluccs, sumout = F) %>%
+  rename(Category = HMPU_TARGETS) %>%
+  st_union(by_feature = TRUE) %>%
+  mutate(
+    Category = paste0(Category, ', ', maxyr)
+  )
+
+# year
+yr <- 1990
+
+cat(yr, 'importing\n')
+print(Sys.time() - strt)
+
+# current year, add coastal stratum
+load(file = here(paste0('data/lulc', yr, '.RData')))
+a <- yr %>%
+  paste0('lulc', .) %>%
+  get %>%
+  lulc_est(coastal, fluccs, sumout = F) %>%
+  rename(Category = HMPU_TARGETS) %>%
+  st_union(by_feature = TRUE) %>%
+  mutate(Category = paste0(Category, ', ', yr))
+b <- maxdat
+
+cat('\tintersecting...\n')
+
+# so intersect doesnt complain about attributes
+st_agr(a) = "constant"
+st_agr(b) = "constant"
+
+# some clean up stuff for slivers
+a <- a %>%
+  st_set_precision(1e5) %>%
+  st_make_valid() %>%
+  st_buffer(dist = 0)
+b <- b %>%
+  st_set_precision(1e5) %>%
+  st_make_valid() %>%
+  st_buffer(dist = 0)
+aunion <- a %>%
+  st_union %>%
+  st_set_precision(1e5) %>%
+  st_make_valid() %>%
+  st_buffer(dist = 0)
+bunion <- b %>%
+  st_union %>%
+  st_set_precision(1e5) %>%
+  st_make_valid() %>%
+  st_buffer(dist = 0)
+
+# get full union
+op1 <- st_difference(a, bunion)
+op2 <- st_difference(b, aunion) %>%
+  rename(Category.1 = Category)
+op3 <- st_intersection(a, b)
+
+chgdatmanu <- bind_rows(op1, op2, op3) %>%
+  mutate(
[…]

### 2022-08-05 R/figs.R

@@ -1,0 +1,21 @@
+library(networkD3)
+library(dplyr)
+library(webshot)
+library(here)
+
+source(here('R/funcs.R'))
+
+# sankey --------------------------------------------------------------------------------------
+
+fluccs <- read.csv(here('data/FLUCCShabsclass.csv'))
+data(chgdatmanu)
+data(subtchgdatmanu)
+
+a1 <- alluvout2(chgdatmanu, fluccs, mrg = 230, height = 700, width = 725, fontSize = 16)
+a2 <- alluvout2(subtchgdatmanu, fluccs, mrg = 130, height = 700, width = 750, fontSize = 16)
+
+saveNetwork(a1, here("figs/chgdatalluv.html"))
+saveNetwork(a2, here("figs/subtchgdatalluv.html"))
+
+# go to chrome and use svgcrowbar to svg, then inkscape svg to png
+# webshot screws up the labels

### 2022-08-05 R/funcs.R

@@ -6,7 +6,7 @@ g_legend <- function(a.gplot){
   return(legend)
 }
 
-# estimate lulc area in acres for available HMPU targets 
+# estimate lulc area in acres for available HMPU targets
 #
 # lulcin in lulc rdata file for a given year
 # coastal is coastal stratum from strats
@@ -22,45 +22,13 @@ lulc_est <- function(lulcin, coastal, fluccs, sumout = T){
   # lulc area, all categories
   # remove open water and subtidal habitats (open water area changes between layers and subtidal not consistently collected)
   out <- lulcin %>%
-    filter(!FLUCCSCODE %in% cds) %>% 
-    add_coast_up(coastal, fluccs) 
-  
-  if(!sumout)
-    return(out)
-  
-  out <- out %>% 
-    mutate(
-      Acres = st_area(.),
-      Acres = set_units(Acres, acres), 
-      Acres = as.numeric(Acres)
-    ) %>% 
-    st_set_geometry(NULL) %>%
-    group_by(HMPU_TARGETS) %>%
-    summarise(Acres = sum(Acres), .groups = 'drop') %>% 
-    arrange(HMPU_TARGETS)
-  
-  return(out)
-  
-}
+    filter(!FLUCCSCODE %in% cds) %>%
+    add_coast_up(coastal, fluccs)
 
-# estimate subtidal area in acres for available HMPU targets 
-#
-# subtin in seagrass rdata file for a given year
-# flucss is lookup table
-subt_est <- function(subtin, fluccs, sumout = TRUE){
-  
-  # subtidal area, all categories
-  out <- subtin %>%
-    mutate(
-      FLUCCSCODE = as.integer(FLUCCSCODE)
-    ) %>% 
-    left_join(fluccs, by = 'FLUCCSCODE') %>% 
-    select(HMPU_TARGETS)
-  
   if(!sumout)
     return(out)
-    
-  out <- out %>% 
+
+  out <- out %>%
     mutate(
       Acres = st_area(.),
       Acres = set_units(Acres, acres),
@@ -68,41 +36,73 @@ subt_est <- function(subtin, fluccs, sumout = TRUE){
     ) %>%
     st_set_geometry(NULL) %>%
     group_by(HMPU_TARGETS) %>%
[…]

### 2022-08-22 manu-draft.qmd
updated figures done

@@
 
-Figure @fig-gisworkflow
+@fig-gisworkflow
 
@@
 
-Figure @fig-subtchgdatalluv
+@fig-subtchgdatalluv
 
@@
 
-Figure @fig-chgdatalluv
+@fig-chgdatalluv
 
@@
 #### Supratidal habitats
+
+@fig-chgdatalluv
 
@@
 
+@fig-oppmap
+
 In 2017, 1,555 km^2^ of land (26.5%) in the Tampa Bay watershed above the MLLW line was classified as natural and 2,144 km^2^ (37%) was considered restorable.
@@
 Xeric soils were roughly aggregated in an east-west band through the middle of the watershed (Fig 4). Approximately 2,107 km2 was classified as xeric, while 3,760 km2 was classified as mesic/hydric.
+
+@fig-restmap
 
@@
 
-![Spatial analysis workflow used to identify opportunity areas for habitat conservation and restoration in the Tampa Bay watershed.
+![Spatial analysis workflow used to identify opportunity areas and restoration potential in the Tampa Bay watershed.
 Workflows are divided into binning of land use/land cover categories into relevant habitat types and spatial overlay of datasets to identify the opportunity areas.](figs/gisworkflow.png){#fig-gisworkflow fig-align="center"}
@@
 
-Figure 4: Map of opportunity assessment
+![Opportunity areas for habitat conservation and reservation in the Tampa Bay watershed.
+Green indicates existing conservation, blue indicates proposed conservation, and pink indicates reservations opportunities.
+Each category is also grouped into native and restorable habitats.](figs/oppmap.png){#fig-oppmap fig-align="center"}
 
-Figure 5: Map of restoration potential
+![Habitat restoration potential in the Tampa Bay watershed.
+Areas are identified as those where habitat restoration could target the four identified categories as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands.
+Categories are based on the opportunity areas and soil types.](figs/restmap.png){#fig-restmap fig-align="center"}
 

### 2022-08-22 R/figs.R

@@ -1,9 +1,18 @@
 library(networkD3)
 library(dplyr)
+library(tidyr)
 library(webshot)
 library(here)
+library(sf)
+library(ggmap)
+library(ggspatial)
 
 source(here('R/funcs.R'))
+
+
+# gis workflow --------------------------------------------------------------------------------
+
+# by hand, original at C:\Users\mbeck\Desktop\TBEP\HMPU personal comp
 
 # sankey --------------------------------------------------------------------------------------
 
@@ -19,3 +28,36 @@ saveNetwork(a2, here("figs/subtchgdatalluv.html"))
 
 # go to chrome and use svgcrowbar to svg, then inkscape svg to png
 # webshot screws up the labels
+
+# opportunity map -----------------------------------------------------------------------------
+
+# data copied from commit e6026f831e90cffca8ea906c8fe53c85ceb5746a in hmpu-workflow
+# these are the original HMPU layers in the report
+data("nativersrv")
+data("restorersrv")
+data("nativelyr")
+data("restorelyr")
+data("coastal")
+data("tbshed")
+
+oppdat <- oppdat_fun(nativersrv, restorersrv, nativelyr, restorelyr, coastal)
+
+p <- oppmap_fun(oppdat, tbshed, northloc = 'tr', scaleloc = 'tl',
+                buffdist = 0.04)
+
+png(here('figs/oppmap.png'), height = 5, width = 8, res = 300, units = 'in')
+p
+dev.off()
+
+# restoration potential map -------------------------------------------------------------------
+
+data("restorelyr")
+
+restdat <- restdat_fun(restorelyr)
+
+p <- restmap_fun(restdat, tbshed, northloc = 'tr', scaleloc = 'tl',
+                 buffdist = 0.04)
+
+png(here('figs/restmap.png'), height = 5, width = 8, res = 300, units = 'in')
+p
+dev.off()

### 2022-08-22 R/funcs.R

@@ -571,7 +571,7 @@ oppdat_fun <- function(nativersrv, restorersrv, nativelyr, restorelyr, coastal, 
 # buffdist is buffer for boundary layer
 #
 # requires ggsn, ggmap, ggplot, sf
-oppmap_fun <- function(oppmap, bndry, ttl, northloc = 'tr', scaleloc = 'tl', buffdist = 0.01){
+oppmap_fun <- function(oppmap, bndry, northloc = 'tr', scaleloc = 'tl', buffdist = 0.01){
 
   # colors
   cols <- list(
@@ -636,10 +636,6 @@ oppmap_fun <- function(oppmap, bndry, ttl, northloc = 'tr', scaleloc = 'tl', buf
       axis.ticks = element_line(colour = 'grey'),
       panel.border = element_rect(colour = 'grey', fill = NA)
     ) +
-    labs(
-      title = ttl,
-      caption = 'More info: https://tbep.org/habitat-master-plan-update/'
-      ) +
     annotation_scale(location = scaleloc) +
     annotation_north_arrow(location = northloc, which_north = "true", height = grid::unit(0.75, "cm"),
                            width = grid::unit(0.75, "cm"))
@@ -661,7 +657,7 @@ oppmap_fun <- function(oppmap, bndry, ttl, northloc = 'tr', scaleloc = 'tl', buf
 # stht is scalebar height
 #
 # requires ggsn, ggmap, ggplot, sf
-restmap_fun <- function(restmap, bndry, ttl, northloc = 'topright', scaleloc = 'topleft', stsz = 3, buffdist = 0.01, scldst = 3, stht = 0.02){
+restmap_fun <- function(restmap, bndry, northloc = 'topright', scaleloc = 'topleft', stsz = 3, buffdist = 0.01, scldst = 3, stht = 0.02){
 
   # colors
   cols <- list(
@@ -709,7 +705,7 @@ restmap_fun <- function(restmap, bndry, ttl, northloc = 'topright', scaleloc = '
 
   # plot
   p <- ggmap(bsmap1_transparent) +
-    geom_sf(data = tomap, aes(fill = HMPU_TA), color = NA, inherit.aes = F, alpha = 0.8) +
+    geom_sf(data = tomap, aes(fill = HMPU_TARGETS), color = NA, inherit.aes = F, alpha = 0.8) +
     geom_sf(data = bndry, fill = NA, color = 'black', inherit.aes = F, size = 0.3) +
     scale_fill_manual(values = cols, drop = F) +
     theme(
[…]

### 2022-08-22 manu-draft.qmd
subtidal habitat change table

@@
   warning: false
+  ft.keepnext: false
+  ft.align: left
 ---
@@
 library(here)
+library(flextable)
+library(dplyr)
+library(officer)
+library(officedown)
+
+load(file = here('tabs/subttab.RData'))
 ```
@@
 Surface area estimates were not available for the Manatee and Pinellas County reefs, but assuming an average size of 4.2 ha, based on the Hillsborough County reefs, the total coverage of artificial reefs in Tampa Bay is estimated to be approximately 67 ha (Table 2).
+
+Table 1
 
@@
 
-Table 1: Land use change table
+```{r}
+#| tbl-cap: Change over time in hectares for subtidal habitats in Tampa Bay.
+Columns show years with available data and the final two columns show the change and percent change from 1988 to 2018.
+Oyster bars were not meaningfully quantified prior to 2014.
+subttab
+```
 
-Table 2: Subtidal change table
+Table 2: Land use change table
 

### 2022-08-22 R/dat_copy.R

@@ -8,7 +8,6 @@ fls <- list.files(here('../hmpu-workflow/data'), full.names = T, pattern = '\\.R
 
 file.copy(fls, to = here('data/'), overwrite = T)
 
-
 #  get R functions ----------------------------------------------------------------------------
 
 file.copy(here('../hmpu-workflow/R/funcs.R'), here('R/'), overwrite = T)

### 2022-08-22 R/figs.R

@@ -9,7 +9,6 @@ library(ggspatial)
 
 source(here('R/funcs.R'))
 
-
 # gis workflow --------------------------------------------------------------------------------
 
 # by hand, original at C:\Users\mbeck\Desktop\TBEP\HMPU personal comp

### 2022-08-22 R/tabs.R

@@ -1,0 +1,54 @@
+library(flextable)
+library(tidyverse)
+
+# subtidal habitat change ---------------------------------------------------------------------
+
+data(subtacres)
+
+losscol <- 'red'
+gaincol <- 'green'
+
+# format for the table
+totab <- subtacres %>%
+  filter(HMPU_TARGETS %in% c('Seagrasses', 'Tidal Flats', 'Oyster Bars')) %>%
+  rename(
+    `Habitat Type` = HMPU_TARGETS,
+    yr = name,
+    Hectares = Acres
+  ) %>%
+  filter(yr <= 2018) %>%
+  mutate(
+    Hectares = Hectares / 2.471
+  ) %>%
+  tidyr::pivot_wider(names_from = yr, values_from = Hectares) %>%
+  mutate(
+    `1988 to 2018` = `2018` - `1988`,
+    `% change` = 100 * (`2018` - `1988`) / `1988`,
+    chgcol = ifelse(`% change` < 0, 0, 1)
+  )
+
+thm <- function(x){
+  flextable::colformat_double(x, digits = 0, na_str = '-') %>%
+    flextable::align(align = 'center', part = 'all') %>%
+    flextable::align(align = 'left', j = 1, part = 'all') %>%
+    flextable::border_inner() %>%
+    flextable::width(width = 0.7, j = 1) %>%
+    flextable::width(width = 5.8 / (ncol_keys(.) - 1), j = 2:ncol_keys(.)) %>%
+    flextable::padding(padding = 0, part = 'all') %>%
+    flextable::fontsize(size = 7, part = 'all') %>%
+    flextable::padding(padding = 0, part = 'all') %>%
+    flextable::fontsize(size = 7, part = 'all')
+}
+
+# table
+subttab <- flextable::flextable(totab, col_keys = grep('^chgcol$', names(totab), value = T, invert = T)) %>%
+  flextable::bg(i = ~ `chgcol` == 0, j = '% change', bg = losscol) %>%
+  flextable::bg(i = ~ `chgcol` == 1, j = '% change', bg = gaincol) %>%
+  thm
+
+# subttab
+save(subttab, file = here('tabs/subttab.RData'))
+
+# inter/suptratidal habitat change ------------------------------------------------------------
+
+

### 2022-08-22 manu-draft.qmd
inter and supratidal habitat change analysis table done

@@
 load(file = here('tabs/subttab.RData'))
+load(file = here('tabs/acretab.RData'))
 ```
@@
 
+Table 2
+
 Between 1990 and 2017, the suite of emergent tidal wetlands (mangroves, salt barrens, and salt marshes) experienced a net gain of 725 ha (10%, Table 3b).
@@
 @fig-chgdatalluv
+
+Table 2
 
@@
 
-Table 2: Land use change table
+```{r}
+#| tbl-cap: Change over time in hectares for intertidal and supratidal habitats in Tampa Bay.
+Columns show years with available data and the final two columns show the change and percent change from 1990 to 2017.
+acretab
+```
 

### 2022-08-22 R/tabs.R

@@ -1,5 +1,6 @@
 library(flextable)
 library(tidyverse)
+library(here)
 
 # subtidal habitat change ---------------------------------------------------------------------
 
@@ -51,4 +52,58 @@ save(subttab, file = here('tabs/subttab.RData'))
 
 # inter/suptratidal habitat change ------------------------------------------------------------
 
+data(acres)
 
+losscol <- 'red'
+gaincol <- 'green'
+
+# format for the table
+totab <- acres %>%
+  rename(
+    `Habitat Type` = HMPU_TARGETS,
+    yr = name,
+    Hectares = Acres
+  ) %>%
+  filter(yr <= 2017) %>%
+  mutate(
+    Stratum = case_when(
+      `Habitat Type` %in% c('Mangrove Forests', 'Salt Barrens', 'Salt Marshes') ~ 'Intertidal',
+      T ~ 'Supratidal'
+    ),
+    Hectares = Hectares / 2.471
+  ) %>%
+  tidyr::pivot_wider(names_from = yr, values_from = Hectares) %>%
+  mutate(
+    `1990 to 2017` = `2017` - `1990`,
+    `% change` = 100 * (`2017` - `1990`) / `1990`,
+    chgcol = case_when(
+      `% change` < 0 & `Habitat Type` != 'Developed' ~ 0,
+      `% change` >= 0 & `Habitat Type` != 'Developed' ~ 1,
+      `% change` < 0 & `Habitat Type` == 'Developed' ~ 1,
+      `% change` >= 0 & `Habitat Type` == 'Developed' ~ 0,
+    )
+  ) %>%
+  arrange(Stratum, `Habitat Type`) %>%
+  as_grouped_data(groups = 'Stratum')
+
+thm <- function(x){
+  flextable::colformat_double(x, digits = 0, na_str = '') %>%
+    flextable::align(align = 'center', part = 'all') %>%
+    flextable::align(align = 'left', j = 1, part = 'all') %>%
+    flextable::border_inner() %>%
+    flextable::width(width = 0.7, j = 1) %>%
+    flextable::width(width = 5.8 / (ncol_keys(.) - 1), j = 2:ncol_keys(.)) %>%
+    flextable::padding(padding = 0, part = 'all') %>%
+    flextable::fontsize(size = 7, part = 'all') %>%
+    flextable::padding(padding = 0, part = 'all') %>%
+    flextable::fontsize(size = 7, part = 'all')
+}
+
[…]

### 2022-08-22 manu-draft.qmd
current extent table done

@@
 load(file = here('tabs/acretab.RData'))
+load(file = here('tabs/currtab.RData'))
 ```
@@
 
-Table 3: Summary of the opportunity assessment analysis
+```{r}
+#| tbl-cap: Summary of habitat restoration opportunities in the Tampa Bay watershed.
+Summaries are based on 2017 land use data, 2018 subtidal data, best estimates for habitat types not in existing GIS layers (e.g., living shoreline), and current extent of existing and proposed conservation lands.
+currtab
+```
 

### 2022-08-22 R/funcs.R

@@ -171,9 +171,8 @@ fixgeo <- function(dat){
 # fluccs is fluccs data frame
 # nativelyr is current existing/proposed native sf object
 # restorelyr is current existing/proposed restoration layer
-# cap is chr string for caption
 # crplyr is optional cropping layer
-curex_fun <- function(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strata, nativelyr, restorelyr, cap, crplyr = NULL){
+curex_fun <- function(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strata, nativelyr, restorelyr, crplyr = NULL){
 
   # crop all sf objects by optional crop layer
   if(!is.null(crplyr)){
@@ -247,22 +246,28 @@ curex_fun <- function(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strat
 
   # current summary
   cursum <- bind_rows(lulcsum, subtsum, hardsum, artisum, tidtsum, livssum) %>%
+    rename(
+      Hectares = Acres,
+      Kilometers = Miles
+    ) %>%
     mutate(
+      Hectares = Hectares / 2.471,
+      Kilometers = Kilometers * 1.609,
       unis = case_when(
-        is.na(Acres) ~ 'mi',
-        is.na(Miles) ~ 'ac'
+        is.na(Hectares) ~ 'km',
+        is.na(Kilometers) ~ 'ha'
       ),
       `Current Extent` = case_when(
-        is.na(Acres) ~ Miles,
-        is.na(Miles) ~ Acres
+        is.na(Hectares) ~ Kilometers,
+        is.na(Kilometers) ~ Hectares
       )
     ) %>%
     left_join(strata, ., by = 'HMPU_TARGETS') %>%
     filter(!HMPU_TARGETS %in% 'Total Intertidal') %>%
     mutate(
       unis = case_when(
-        is.na(unis) & HMPU_TARGETS %in% c('Living Shorelines', 'Tidal Tributaries') ~ 'mi',
-        is.na(unis) & !HMPU_TARGETS %in% c('Living Shorelines', 'Tidal Tributaries') ~ 'ac',
+        is.na(unis) & HMPU_TARGETS %in% c('Living Shorelines', 'Tidal Tributaries') ~ 'km',
+        is.na(unis) & !HMPU_TARGETS %in% c('Living Shorelines', 'Tidal Tributaries') ~ 'ha',
         T ~ unis
       ),
[…]

### 2022-08-22 R/tabs.R

@@ -1,6 +1,10 @@
 library(flextable)
 library(tidyverse)
 library(here)
+library(sf)
+library(units)
+
+source(here('R/funcs.R'))
 
 # subtidal habitat change ---------------------------------------------------------------------
 
@@ -35,8 +39,6 @@ thm <- function(x){
     flextable::border_inner() %>%
     flextable::width(width = 0.7, j = 1) %>%
     flextable::width(width = 5.8 / (ncol_keys(.) - 1), j = 2:ncol_keys(.)) %>%
-    flextable::padding(padding = 0, part = 'all') %>%
-    flextable::fontsize(size = 7, part = 'all') %>%
     flextable::padding(padding = 0, part = 'all') %>%
     flextable::fontsize(size = 7, part = 'all')
 }
@@ -94,8 +96,6 @@ thm <- function(x){
     flextable::width(width = 0.7, j = 1) %>%
     flextable::width(width = 5.8 / (ncol_keys(.) - 1), j = 2:ncol_keys(.)) %>%
     flextable::padding(padding = 0, part = 'all') %>%
-    flextable::fontsize(size = 7, part = 'all') %>%
-    flextable::padding(padding = 0, part = 'all') %>%
     flextable::fontsize(size = 7, part = 'all')
 }
 
@@ -107,3 +107,36 @@ acretab <- flextable::flextable(totab, col_keys = grep('^chgcol$', names(totab),
 
 # acretab
 save(acretab, file = here('tabs/acretab.RData'))
+
+# current extent table ------------------------------------------------------------------------
+
+fluccs <- read.csv(here('data', 'FLUCCShabsclass.csv'), stringsAsFactors = F)
+
+lulcfl <- 'lulc2017'
+subtfl <- 'sgdat2018'
+
+# from 01_inputs
+load(here('data', paste0(lulcfl, '.RData')))
+load(here('data', paste0(subtfl, '.RData')))
+lulc <- get(lulcfl)
+subt <- get(subtfl)
+data(hard)
+data(arti)
+data(tidt)
+data(livs)
+data(coastal)
+data(strata)
+
+# data copied from commit e6026f831e90cffca8ea906c8fe53c85ceb5746a in hmpu-workflow
+# these are the original HMPU layers in the report
+data("restorelyr")
+data("nativelyr")
+
[…]

### 2022-08-23 manu-draft.qmd
target table added, tables done

--- 2022-08-22 manu-draft.qmd
@@
 library(dplyr)
-library(officer)
-library(officedown)
 
@@
 load(file = here('tabs/acretab.RData'))
+load(file = here('tabs/currtab.RData'))
+load(file = here('tabs/targtab.RData'))
 ```
@@
 
+\newpage
+
 ## Tables
@@
 
+\newpage
+
 ```{r}
@@
 
-Table 3: Summary of the opportunity assessment analysis
+\newpage
 
-Table 4: Targets
+```{r}
+#| tbl-cap: Summary of habitat restoration opportunities in the Tampa Bay watershed.
+Summaries are based on 2017 land use data, 2018 subtidal data, best estimates for habitat types not in existing GIS layers, and current extent of existing and proposed conservation lands.
+currtab
+```
+
+\newpage
+
+```{r}
+#| tbl-cap: Recommended 2030 targets and 2050 goals for habitat restoration and protection in the Tampa Bay watershed.
+Targets and goals are based on 2017 land use data, 2018 subtidal data, best estimates for habitat types not in existing GIS layers, and current extent of existing and proposed conservation lands.
+targtab
+```
+
+\newpage
 

### 2022-08-23 R/dat_proc.R

@@ -5,6 +5,47 @@ library(units)
 
 source(here('R/funcs.R'))
 
+
+# targets as metric (ha, mi) ------------------------------------------------------------------
+
+trgsmetric <- structure(
+  list(
+    Category = structure(
+      c(1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 3L),
+      .Label = c("Subtidal", "Intertidal", "Supratidal"),
+      class = "factor"
+      ),
+    HMPU_TARGETS = structure(
+      1:15,
+      .Label = c("Hard Bottom", "Artificial Reefs", "Tidal Flats", "Seagrasses", "Oyster Bars", "Living Shorelines", "Total Intertidal", "Mangrove Forests", "Salt Barrens", "Salt Marshes", "Tidal Tributaries", "Coastal Uplands", "Non-Forested Freshwater Wetlands", "Forested Freshwater Wetlands", "Native Uplands"),
+      class = "factor"
+      ),
+    Target2030 = c(171, 67, 6564, 16188, 89, 34, 8641, 6192, 221, 1945, 6, 1525, 27898, 61628, 57082),
+    Target2050 = c(171, 67, 6564, 16188, 191, 90, 9633, 6192, 322, 2208, 29, 1707, 29052, 61810, 57507),
+    rationale = c(
+      "Protect existing hard bottom; continue to identify new hard bottom area using proven mapping techniques",
+      "Protect existing artificial reefs; enhance habitat complexity where feasible; expand reef area to promote fish and wildlife benefits",
+      "Identify and protect existing persistent tidal flats; assess restoration potential of other non-vegetated subtidal areas",
+      "Protect existing seagrasses; establish new HMPU lower limit of 16,188 hectares; assess restoration potential of non-vegetated subtidal areas",
+      "2030: Protect existing oysters + restore 20 hectares; increase target by 20 hectares each out-decade; consider filtration rate to refine long-term goal; an oyster habitat suitability index (HSI) will inform opportunity space",
+      "2030: Construct 1.6 kilometers of LS each year; includes privately owned seawalls; need better definition of opportunity areas; increase target to 2.4 & 3.2 kilometers per year for out decades",
+      "2030: Protect existing intertidal mosaic + restore 405 hectares (based on hydric soils); increase target by 61 hectares each out-decade; includes the mosaic of mangrove, salt barren, and salt marsh habitats",
+      "Protect existing mangrove forests; restore opportunistically within the intertidal mosaic",
[…]

### 2022-08-23 R/funcs.R

@@ -775,9 +775,8 @@ restdat_fun <- function(restorelyr, crplyr = NULL){
 # coastal is coastal stratum sf object
 # fluccs is fluccs data frame
 # restorelyr is current existing/proposed restoration layer
-# trgs is input targets table
-# cap is chr string for caption
-target_fun <- function(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strata, restorelyr, trgs, cap){
+# trgsmetric is input targets table as metric values (ha, km)
+target_fun <- function(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strata, restorelyr, trgsmetric){
 
   # lulc area, all categories
   lulcsum <- lulc_est(lulc, coastal, fluccs)
@@ -833,14 +832,20 @@ target_fun <- function(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, stra
 
   # current summary
   cursum <- bind_rows(lulcsum, intrsum, subtsum, hardsum, artisum, tidtsum, livssum) %>%
+    rename(
+      Hectares = Acres,
+      Kilometers = Miles
+    ) %>%
     mutate(
+      Hectares = Hectares / 2.471,
+      Kilometers = Kilometers * 1.609,
       unis = case_when(
-        is.na(Acres) ~ 'mi',
-        is.na(Miles) ~ 'ac'
+        is.na(Hectares) ~ 'km',
+        is.na(Kilometers) ~ 'ha'
       ),
       `Current Extent` = case_when(
-        is.na(Acres) ~ Miles,
-        is.na(Miles) ~ Acres
+        is.na(Hectares) ~ Kilometers,
+        is.na(Kilometers) ~ Hectares
       )
     ) %>%
     inner_join(strata, by = 'HMPU_TARGETS') %>%
@@ -851,14 +856,14 @@ target_fun <- function(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, stra
 
   restoresum <- restorelyr %>%
     mutate(
-      Acres = st_area(.),
-      Acres = set_units(Acres, acres),
-      Acres = as.numeric(Acres),
+      Hectares = st_area(.),
+      Hectares = set_units(Hectares, hectares),
+      Hectares = as.numeric(Hectares),
       typ = paste('restorable', typ)
     ) %>%
     st_set_geometry(NULL) %>%
[…]

### 2022-08-23 R/tabs.R

@@ -132,11 +132,47 @@ data(strata)
 data("restorelyr")
 data("nativelyr")
 
-curtab <- curex_fun(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strata, nativelyr, restorelyr)
+tab <- curex_fun(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strata, nativelyr, restorelyr)
 
-currtab <- curtab %>%
+currtab <- tab %>%
   flextable::width(width = 6.5 / ncol_keys(.)) %>%
   flextable::padding(padding = 0, part = 'all') %>%
   flextable::fontsize(size = 8, part = 'all')
 
 save(currtab, file = here('tabs/currtab.RData'))
+
+# target table --------------------------------------------------------------------------------
+
+fluccs <- read.csv(here('data', 'FLUCCShabsclass.csv'), stringsAsFactors = F)
+
+lulcfl <- 'lulc2017'
+subtfl <- 'sgdat2018'
+
+# from 01_inputs
+load(here('data', paste0(lulcfl, '.RData')))
+load(here('data', paste0(subtfl, '.RData')))
+lulc <- get(lulcfl)
+subt <- get(subtfl)
+data(hard)
+data(arti)
+data(tidt)
+data(livs)
+data(coastal)
+data(strata)
+data(trgsmetric)
+
+# data copied from commit e6026f831e90cffca8ea906c8fe53c85ceb5746a in hmpu-workflow
+# these are the original HMPU layers in the report
+data(restorelyr)
+
+tab <- target_fun(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strata, restorelyr, trgsmetric)
+
+targtab <- tab %>%
+  flextable::width(width = 0.45, j = 1) %>%
+  flextable::width(width = 0.66, j = 2:6) %>%
+  flextable::width(width = 2.75, j = 7) %>%
+  fontsize(size = 8, part = 'all') %>%
+  fontsize(size = 7, j = 7, part = 'body') %>%
+  flextable::padding(padding = 0, part = 'all')
+
+save(targtab, file = here('tabs/targtab.RData'))

### 2022-08-24 manu-draft.qmd
intro revision

@@
 
-The health of estuarine systems, coastal habitats, and associated fauna and flora are inextricably linked to land uses and management throughout the watershed [@Yoskowitz15] These habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18].
-The Tampa Bay Estuary Program (TBEP) is one of 28 programs administered by the US Environmental Protection Agency (USEPA) under the National Estuary Program (NEP).
-In recognition of threats to habitats from development and climate change stressors, the TBEP and partners recently created the third iteration of a plan to establish targets and goals for habitat restoration within the Tampa Bay watershed.
-As an NEP, the program has guided regional environmental restoration initiatives for the estuary since 1991.
-The methodologies used in the creation of a 2020 Habitat Master Plan Update for Tampa Bay [@Robison20] are highly transferable to both coastal and non-coastal systems.
+The health of estuarine systems and coastal habitats is inextricably linked to land uses and management of the watershed [@Yoskowitz15].
+Coastal habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18].
+Pervasive stressors can negatively impact the services provided by these habitats.
+Anthropogenic land development can remove or substantially modify existing habitats, and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14].
+Related changes in habitats in response to climate change (sea-level rise and warming) include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].
 
-## Development, climate change, and coastal squeeze
+The combined effects of land development and climate change are especially problematic for prioritizing habitat restoration or conservation activities in coastal environments.
+Landward migration of critical habitats in response to sea-level rise may not be possible due to anthropogenic barriers in the watershed.
+Sea-level rise can occur quicker than landward migration of salt marshes and the upland slope may already be lost to urban development and hardening [@Titus09].
+The coastal squeeze of development and climate change diminishes the available space for future restoration in urban estuaries, thereby impacting the variety of ecosystem services provided by these habitats and the wildlife they support [@Enwright15].
+Given projected habitat losses without management intervention and the limited resources available (i.e., time, land, funding, and labor force), appropriate and realistic sites for restoration need to be identified that account for future stressors and past trends.
 
-Development causes multiple perturbations within any watershed.
-The accumulated impacts of construction and associated infrastructure remove or substantially modify existing habitats, and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14].
-Given the degree to which the Tampa Bay watershed has been urbanized, the synergistic effects of continued coastal development and future climate change are primary concerns for maintenance of estuarine and coastal habitat health.
-Observed and potential adverse effects of climate change and sea level rise on marine and estuarine ecosystems are well-documented [@Scavia02; @Spalding07; @Titus09].
-With regard to estuarine habitats, the primary concerns are that sea level rise is now occurring at such a rapid rate that the landward migration of tidal wetlands in response cannot keep pace; or that the upland slope has already been lost to urban development and hardening, leaving no place for tidal wetland migration [@Titus09].
-Geological, physical and chemical changes could include alterations in sediment deposition and erosion patterns, micro-topography, and water quality [@Arnell15; @Whitehead09].
+Past approaches for guiding restoration planning have been successfully used in other contexts, but they do not fully balance competing needs among public and private sectors, nor do they fully account for anticipated effects of multiple stressors.
+For example, an integrated watershed approach [@EPA96] has been utilized since the early 1990s to diagnose and manage water quantity and quality problems by addressing issues within hydrologically-defined geographic areas.
+Additionally, the habitat mosaic approach [@Henningsen05] of including multiple habitat types within restoration projects has been recognized as an effective means of allowing ecosystem state changes in response to different environmental pressures [@Duarte09; @Palmer09].
+Adaptive management [@Holling78; @Gregory06] components have been increasingly used to address challenges of sea-level rise, climate change, and development stressors, including monitoring to identify critical restoration decision points and needed intervention with contingency plans.
+Elements of each of these approaches could be combined to create a more holistic approach to guide restoration and conservation activities for coastal habitats in urban settings.
 
-Related changes in habitats in response to climate change (sea level rise and warming) include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].
-Species dependent upon these habitats will be forced to change use patterns or adapt to the new conditions [@Iwamura13].
-For example, while black needle rush (*Juncus roemerianus*) can tolerate a wide salinity range [@Eleuterius79; @Stout84], the largest remaining *J.
-roemerianus* marshes in Tampa Bay are located in the lower-salinity reaches of tidal rivers and creeks.
-The greatest extents of those marshes occur in river systems where their upstream extent is constrained by impoundments for public water supplies.
-Spatial restriction in these hydrologically truncated rivers may make these marshes particularly vulnerable to "pinching out", as upstream migration in response to sea level rise will be cut off by anthropogenic barriers.
-Similarly, landward migration of salt barrens (high marsh areas in Tampa Bay) in response to sea level rise will be restricted by the filling and hardening of coastal uplands associated with existing or future urban development.
+The Tampa Bay watershed (Florida, USA) is a valuable case study for developing a habitat restoration and conservation plan that addresses pervasive coastal stressors.
+Compared to other estuaries, the ratio of watershed to estuary area is small and the area is heavily developed with 42% of land use classified as urban and suburban residential [@SWFWMD18].
+A retrospective approach to setting habitat protection and restoration targets in Tampa Bay was previously employed [@Lewis96; @Robison10; @Cicchetti11; @Russell15].
+Priority was given to restoration activities focused on habitat types that were important for a suite of estuarine faunal guilds disproportionately lost or degraded compared to a benchmark period.
+Criticisms of this approach included a lack of consideration for future sea-level rise and other climate change factors [@Yoskowitz15], use of expanded and different habitats outside the Tampa Bay watershed [@Robison10], lack of attention to upland or freshwater wetland habitats, and little recognition of land development trends or actual available space for restoration efforts.
+These challenges are shared by restoration practitioners in other coastal environments and an approach that accommodates these challenges for planning would be highly transferable.
 
-## Need for a paradigm shift
-
-The synergistic effects of development and climate change diminish the available space for future restoration in urbanizing estuaries, thereby impacting the variety of ecosystem services provided by these habitats and the wildlife they support [@Enwright15].
-Given projected habitat losses without intervention efforts and the limited resources (including time, land, funding, and labor force) available, it is important to appropriately and realistically site restoration projects to increase the likelihood of success.
-To achieve this objective, a new restoration approach on a broad watershed scale will be implemented.
-
-Previously, a "retrospective" approach to setting habitat protection and restoration targets in Tampa Bay was employed [@Lewis96; @Robison10; @Cicchetti11; @Russell15].
-Under this paradigm, priority was given to restoration activities focused on habitat types, important for a suite of ten estuarine faunal guilds, that were disproportionately lost or degraded compared to a benchmark period.
-Primary criticisms of this approach included a lack of consideration for future sea level rise and other climate change factors [@Yoskowitz15], use of expanded and different habitats outside the Tampa Bay watershed by the faunal guilds [@Robison10], lack of attention to upland or freshwater wetland habitats, and little recognition of other stressors such as land development trends or actual available space for restoration efforts.
-
-Past approaches for guiding restoration planning have been successfully used in other contexts, but they do not fully balance competing needs.
-For example, an integrated watershed approach [@EPA96] has been utilized since the early 1990s to diagnose and manage water quantity and quality problems that have contributed to seagrass restoration in the system.
-Additionally, the habitat mosaic approach [@Henningsen05] of including multiple habitat types within restoration projects is recognized as necessary in Tampa Bay [@Hughes11] and elsewhere to allow for ecosystem state changes in response to different environmental pressures [@Duarte09; @Palmer09].
-
-Adaptive management [@Holling78; @Gregory06] components have been increasingly used to address challenges of sea level rise, climate change, and development stressors, including monitoring to identify critical restoration decision points and needed intervention with contingency plans.
-Rising sea levels and temperatures and altered rainfall patterns are causing observable changes to habitats on a global scale [@Cavanaugh14; @Garner15; @Yoskowitz15], including within Tampa Bay [@Price17], and those changes are expected to become more pronounced over the next several decades [@Sheehan16; @Nerem18].
-
-## Updating the approach
-
-The new approach integrates the whole watershed, addresses historical changes, focuses on trajectories that have occurred during more contemporary time periods, and considers both current and future stressors -- particularly land development and sea level rise.
-There is relatively consistent extent and distribution data for most Tampa Bay habitats of interest (1988 to 2018), representing a time period when federal, state and local regulations were in effect and regional impacts from climate change are documented [@Raabe12; @Cavanaugh14].
-This approach establishes a broader framework that guides both watershed-level habitat master planning and site-level restoration design activities and incorporates applicable elements of the other habitat restoration paradigms discussed above [@Palmer09].
-Our broader framework for guiding restoration activities includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in acreage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories.
-
-With regard to habitat restoration projects, the design approach must envision not only what is possible today, but also what the coastal landscape will look like in 50 years and beyond.
-Design features should continue to use the historical "habitat mosaic" approach, but should also include coastal upland features that accommodate tidal inundation and the landward advance of emergent tidal wetlands [@Enwright16].
+This paper describes an approach for habitat restoration and conservation planning that addresses the above challenges by considering the whole watershed, addressing historical changes, focusing on trajectories that have occurred during contemporary time periods, and considering both current and future stressors -- particularly land development and sea-level rise.
+Relatively consistent data are available for most Tampa Bay habitats of interest (1988 to 2018), representing a time period when federal, state and local regulations were in effect and regional impacts from climate change are documented [@Raabe12; @Cavanaugh14].
+The approach establishes a framework that can guide both watershed-level habitat planning and site-level restoration activities and incorporates applicable elements of other habitat restoration paradigms discussed above [@Palmer09].
+The general approach includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in coverage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories.
+The outcomes are spatially-specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.
 
@@
 
-Addressing the full suite of habitats in a watershed is now recognized as critical for large-scale restoration planning efforts [@Palmer09; @Lamb18]**.** The major habitat types of Tampa Bay can be described and organized relative to tidal influence and location in the watershed.
+Addressing the full suite of habitats in a watershed is critical for large-scale restoration planning efforts [@Palmer09; @Lamb18]**.** The major habitat types of Tampa Bay can be described and organized relative to tidal influence and location in the watershed.
 Subtidal habitats include those that are submerged all or most of the time; emergent tidal wetlands include those that are submerged during high tides but exposed during low tides; and supratidal habitats include those that occur above the high tide line.
@@
 The first is the coastal stratum, which extends from the local Mean Lower Low Water elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation by 2070.
-The coastal stratum is the zone where emergent tidal wetland restoration would be conducted and includes low lying coastal uplands which serve as important tidal wetland buffers that will be critically important in the future as lands reserved to accommodate ideal wetland migration in response to sea level rise.
+The coastal stratum is the zone where emergent tidal wetland restoration would be conducted and includes low lying coastal uplands which serve as important tidal wetland buffers that will be critically important in the future as lands reserved to accommodate ideal wetland migration in response to sea-level rise.
 The river floodplain stratum includes all hydrologically contiguous forested and non-forested wetlands within the river and stream corridors of the Tampa Bay watershed.
@@
 
+Table 3
+
 Integration of all of the opportunity layers provides a summary of the available restoration for the TBEP habitats of interest (Table 5).
@@
 
+Table 4
+
 Recommended targets were based on habitat status and trends, habitat restoration data, identified restoration opportunities, and current and anticipated trends in development, available funding, and regulations [@Hilderbrand05; @Hobbs07].
@@
 However, it also recognizes that the identified habitat protection and restoration areas will change over time, and will be revisited on a 10-year recurring cycle.
-A 30-year planning horizon (2050) is also identified based upon sea level rise projections developed specifically for Tampa Bay [@Burke19].
-The coastal stratum (from the existing mean low water line to the approximate 5-foot contour) is projected to directly experience the effects of sea level rise by 2050, and is the primary focus area for coastal habitat protection and restoration activities.
+A 30-year planning horizon (2050) is also identified based upon sea-level rise projections developed specifically for Tampa Bay [@Burke19].
+The coastal stratum (from the existing mean low water line to the approximate 5-foot contour) is projected to directly experience the effects of sea-level rise by 2050, and is the primary focus area for coastal habitat protection and restoration activities.
 Land acquisition or protection (through conservation easements or other mechanisms) will be needed to ensure completion of targets and goals for both salt marsh and upland habitats (Fig 3).
@@
 Evolving information such as an Oyster Habitat Suitability Index [@Boswell12] and ongoing mapping exercises will be used to identify optimal locations to conduct restoration activities that help achieve targets and goals.
-Coordination with the establishment of state-mandated minimum flows [@Munson07] will be necessary for restoration and maintenance of low-salinity salt marsh habitats that will experience higher salinities and rapid transition to mangroves under existing sea level rise scenarios [@Sherwood14; @Geselbracht15].
+Coordination with the establishment of state-mandated minimum flows [@Munson07] will be necessary for restoration and maintenance of low-salinity salt marsh habitats that will experience higher salinities and rapid transition to mangroves under existing sea-level rise scenarios [@Sherwood14; @Geselbracht15].
 
@@
 
-The intertidal zone in Tampa Bay is currently experiencing dynamic change, driven by sea level rise and climate change, whereby mangrove forests are outcompeting salt marshes and salt barrens for the available niche space.
+The intertidal zone in Tampa Bay is currently experiencing dynamic change, driven by sea-level rise and climate change, whereby mangrove forests are outcompeting salt marshes and salt barrens for the available niche space.
 Without increasing the total area of the intertidal zone, restoring a greater coverage of mangroves would reduce the niche space available for salt marshes and salt barrens.
-This phenomenon has been observed throughout the Gulf of Mexico, and has been attributed to both climate change (e.g., fewer freeze events) and sea level rise [@Comeaux12].
+This phenomenon has been observed throughout the Gulf of Mexico, and has been attributed to both climate change (e.g., fewer freeze events) and sea-level rise [@Comeaux12].
 
 The observed gains in wetlands (Table 3b) are likely a reflection of: 1) the effectiveness of state and federal wetland regulatory programs; and 2) the cumulative gains resulting from, primarily, publicly-funded habitat restoration projects (Table 4).
-Minor gains in some emergent tidal wetlands (e.g., salt barrens) may also be a reflection of the landward expansion of the complex suite of these habitats associated with climate change and sea level rise.
+Minor gains in some emergent tidal wetlands (e.g., salt barrens) may also be a reflection of the landward expansion of the complex suite of these habitats associated with climate change and sea-level rise.
 Gains in vegetated non-forested freshwater wetlands are related to the clearing of forested wetlands followed by the creation of herbaceous mitigation areas and stormwater systems.
@@
 
-As discussed, land acquisition for coastal habitat restoration must prioritize adjacent low-lying coastal uplands to serve as buffers to accommodate future landward migration of tidal wetlands in response to sea level rise.
+As discussed, land acquisition for coastal habitat restoration must prioritize adjacent low-lying coastal uplands to serve as buffers to accommodate future landward migration of tidal wetlands in response to sea-level rise.
 Where public acquisition is not possible, other conservation mechanisms need to be explored.
@@
 The application of rolling easements in Tampa Bay could disincentivize more intense urban development (e.g., discourage up-zoning) of low-lying coastal uplands that may be currently in less intense agricultural or recreational (e.g., golf courses) land uses.
-Under a rolling easement, landowners would be able to maintain current economic uses, while "reserving" such lands to accommodate tidal wetland migration with advancing sea level rise.
+Under a rolling easement, landowners would be able to maintain current economic uses, while "reserving" such lands to accommodate tidal wetland migration with advancing sea-level rise.
 

### 2022-08-26 manu-draft.qmd
adding lulc map

@@
 Related changes in habitats in response to climate change (sea-level rise and warming) include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].
+Management of coastal habitats requires an understanding of how stressores impact these resources, while balancing competing societal uses.
 
@@
 
-Tampa Bay is a large open water estuary (open water area approximately 983 km^2^) on the west-central coast of Florida (Fig 1).
-Its watershed encompasses another 5,872 km^2^, for a total combined area of approximately 6,855 km^2^.
-It is subtropical, and within the current (2020) ecotone for mangrove and salt marsh habitats.
+Tampa Bay is a large open water estuary (open water area approximately 983 km^2^) on the west-central coast of Florida (@fig-lulcmap).
+The watershed includes another 5,872 km^2^, for a total combined area of approximately 6,855 km^2^.
+The climate is subtropical and within the current (2020) ecotone for mangrove and salt marsh habitats.
 The upstream watershed includes multiple habitats, including pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands.
-The watershed is heavily developed with an estimated (2019) population of 3.3 million people in the four counties that comprise most of the watershed [@Rayer20]*.* Numerous anthropogenic changes have been made to the natural systems within and surrounding Tampa Bay, including direct removal of habitat (including dredge and fill of bay bottom), alteration of hydrology, and destruction and fragmentation of habitat from development.
+The watershed is heavily developed with an estimated (2019) population of 3.3 million people in the four counties that comprise most of the watershed [@Rayer20]*.* Numerous anthropogenic changes have been made to the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom), alteration of hydrology, and destruction and fragmentation of habitat from development.
 
@@
 
+![Land use and land cover for the Tampa Bay watershed, Florida, USA.](figs/lulcmap.png){#fig-lulcmap fig-align="center"}
+
 ![Spatial analysis workflow used to identify opportunity areas and restoration potential in the Tampa Bay watershed.

### 2022-08-26 R/dat_copy.R

@@ -11,3 +11,9 @@ file.copy(fls, to = here('data/'), overwrite = T)
 #  get R functions ----------------------------------------------------------------------------
 
 file.copy(here('../hmpu-workflow/R/funcs.R'), here('R/'), overwrite = T)
+
+# fluccslkup ----------------------------------------------------------------------------------
+
+load(file = url('https://github.com/tbep-tech/reasonable-assurance-analysis/raw/main/data/fluccslkup.RData'))
+
+save(fluccslkup, file = here('data/fluccslkup.RData'))

### 2022-08-26 R/dat_proc.R

@@ -5,47 +5,6 @@ library(units)
 
 source(here('R/funcs.R'))
 
-
-# targets as metric (ha, mi) ------------------------------------------------------------------
-
-trgsmetric <- structure(
-  list(
-    Category = structure(
-      c(1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 3L),
-      .Label = c("Subtidal", "Intertidal", "Supratidal"),
-      class = "factor"
-      ),
-    HMPU_TARGETS = structure(
-      1:15,
-      .Label = c("Hard Bottom", "Artificial Reefs", "Tidal Flats", "Seagrasses", "Oyster Bars", "Living Shorelines", "Total Intertidal", "Mangrove Forests", "Salt Barrens", "Salt Marshes", "Tidal Tributaries", "Coastal Uplands", "Non-Forested Freshwater Wetlands", "Forested Freshwater Wetlands", "Native Uplands"),
-      class = "factor"
-      ),
-    Target2030 = c(171, 67, 6564, 16188, 89, 34, 8641, 6192, 221, 1945, 6, 1525, 27898, 61628, 57082),
-    Target2050 = c(171, 67, 6564, 16188, 191, 90, 9633, 6192, 322, 2208, 29, 1707, 29052, 61810, 57507),
-    rationale = c(
-      "Protect existing hard bottom; continue to identify new hard bottom area using proven mapping techniques",
-      "Protect existing artificial reefs; enhance habitat complexity where feasible; expand reef area to promote fish and wildlife benefits",
-      "Identify and protect existing persistent tidal flats; assess restoration potential of other non-vegetated subtidal areas",
-      "Protect existing seagrasses; establish new HMPU lower limit of 16,188 hectares; assess restoration potential of non-vegetated subtidal areas",
-      "2030: Protect existing oysters + restore 20 hectares; increase target by 20 hectares each out-decade; consider filtration rate to refine long-term goal; an oyster habitat suitability index (HSI) will inform opportunity space",
-      "2030: Construct 1.6 kilometers of LS each year; includes privately owned seawalls; need better definition of opportunity areas; increase target to 2.4 & 3.2 kilometers per year for out decades",
-      "2030: Protect existing intertidal mosaic + restore 405 hectares (based on hydric soils); increase target by 61 hectares each out-decade; includes the mosaic of mangrove, salt barren, and salt marsh habitats",
-      "Protect existing mangrove forests; restore opportunistically within the intertidal mosaic",
[…]

### 2022-08-26 R/figs.R

@@ -6,8 +6,142 @@ library(here)
 library(sf)
 library(ggmap)
 library(ggspatial)
+library(tibble)
+library(grid)
+library(USAboundaries)
 
 source(here('R/funcs.R'))
+
+# lulc map ------------------------------------------------------------------------------------
+
+load(file = here('data/lulc2020.RData'))
+load(file = here('data/sgdat2020.RData'))
+load(file = here('data/fluccslkup.RData'))
+
+# from T:\05_GIS\TBEP\2017_LULC_TBEP_Extent+2020_Seagrass.mxd
+fluccsconv <- list(
+    'Urban/Open Lands' = c(110, 120, 130, 140, 150, 170, 180, 182, 190, 740, 810, 820, 830),
+    'Mining' = c(160),
+    'Reclaimed Mined Lands' = c(165),
+    'Agriculture' = c(210, 214, 220, 230, 240, 250, 255, 260),
+    'Vegetated/Forested/Natural' = c(310, 320, 330, 410, 411, 412, 420, 434, 440, 710, 720),
+    'Wetlands' = c(610, 611, 612, 615, 620, 621, 630, 641 ,642, 643, 644, 652, 653, 660),
+    'Oyster' = c(654),
+    'Tidal Flat' = c(651, 721),
+    'Seagrass/SAV' = c(911, 912),
+    'Water' = c(510, 520, 530, 540, 572)
+  ) %>%
+  enframe(value = 'LEV3') %>%
+  unnest('LEV3') %>%
+  left_join(fluccslkup, by = 'LEV3') %>%
+  mutate(
+    col = case_when(
+      name == 'Urban/Open Lands' ~ '#FF0000',
+      name == 'Mining' ~ '#C500FF',
+      name == 'Reclaimed Mined Lands' ~ '#6D276B',
+      name == 'Agriculture' ~ '#FFFF00',
+      name == 'Vegetated/Forested/Natural' ~ '#00734C',
+      name == 'Wetlands' ~ '#38A800',
+      name == 'Oyster' ~ '#FFAA00',
+      name == 'Tidal Flat' ~ '#FFEABE',
+      name == 'Seagrass/SAV' ~ '#55FF00',
+      name == 'Water' ~ '#0070FF'
+    )
+  )
+
+tomap1 <- lulc2020 %>%
+  left_join(fluccsconv, by = 'FLUCCSCODE') %>%
+  filter(!is.na(name) )%>%
+  st_transform(crs = 4326)
+
+tomap2 <- sgdat2020 %>%
[…]

### 2022-08-26 manu-draft.qmd
working on methods

@@
 
-Tampa Bay is a large open water estuary (open water area approximately 983 km^2^) on the west-central coast of Florida (@fig-lulcmap).
-The watershed includes another 5,872 km^2^, for a total combined area of approximately 6,855 km^2^.
+Tampa Bay is a large open water estuary (surface area approximately 983 km^2^) on the west-central coast of Florida (@fig-lulcmap).
+The watershed covers approximately 5,872 km^2^, for a total combined area of approximately 6,855 km^2^.
 The climate is subtropical and within the current (2020) ecotone for mangrove and salt marsh habitats.
 The upstream watershed includes multiple habitats, including pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands.
-The watershed is heavily developed with an estimated (2019) population of 3.3 million people in the four counties that comprise most of the watershed [@Rayer20]*.* Numerous anthropogenic changes have been made to the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom), alteration of hydrology, and destruction and fragmentation of habitat from development.
+The watershed is heavily developed with an estimated population of 3.3 million people in the four major counties of the watershed [@Rayer20]*.* Numerous anthropogenic changes have altered the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom, mining activities), alteration of hydrology, and destruction and fragmentation of habitat from development.
 
@@
 
-Addressing the full suite of habitats in a watershed is critical for large-scale restoration planning efforts [@Palmer09; @Lamb18]**.** The major habitat types of Tampa Bay can be described and organized relative to tidal influence and location in the watershed.
-Subtidal habitats include those that are submerged all or most of the time; emergent tidal wetlands include those that are submerged during high tides but exposed during low tides; and supratidal habitats include those that occur above the high tide line.
+Addressing the full suite of habitats in a watershed is critical for large-scale restoration planning efforts [@Palmer09; @Lamb18]**.** The major habitat types of Tampa Bay were stratified relative to tidal influence and location in the watershed.
+Subtidal habitats included those that are submerged all or most of the time; intertidal habitats included emergent tidal wetlands that are submerged during high tides but exposed during low tides; and supratidal habitats included those that occur above the high tide line (i.e., the reaminder of the watershed).
 
-Habitats generally described as 'subtidal' include hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13].
-Mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15] are classified as emergent tidal wetlands.
-For the purposes of this planning effort, supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* As discussed below, uplands are sub-divided into coastal and non-coastal uplands, based on location relative to the 5-foot contour.
+Subtidal included hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13].
+Mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15] were classified as emergent tidal wetlands.
+Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.*  Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below).
 
@@
 
-Recommended targets were created stepwise with geospatial analyses: 1) quantifying habitat status and historical trends; 2) synthesizing historic habitat restoration efforts; and 3) identifying and defining remaining 'opportunities' for restoration that integrated results from the first two analyses.
-Final development of restoration targets and goals integrated strata and opportunity area information with an analysis of past restoration implementation to use previous completed projects as a guide to a feasible amount of restoration that could be conducted by partners.
-This approach will ensure that the estuarine-dependent species and faunal guilds throughout the watershed, as defined in the original approach, will continue to be supported.
+Opportunity areas for restoration and recommended coverage targets for habitat types were identifed by integrating multiple datasets available for the region.
+Habitat status and historical trends were first quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time.
+Historic habitat restoration efforts conducted in the Tampa Bay watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners.
+Finally, remaining opportunities for restoration that integrated results from the first two analyses were identified based on current status of habitat coverage in the watershed.
+This included combining current coverages with existing protected areas and considering their location within the three strata (sub-, inter-, supratidal) and areas anticipated to be impacted by sea-level rise and development near coastal areas.
+The final results included spatially-explicit maps that identified opportunity areas and restoration potential for remaining native habitats in the watershed.
+Short-term (2030) goals and long-term (2050) targets for coverage (hectares) of native habitats were informed by current status, past trends, and historical restoration efforts in the region.
 
@@
 
-For the majority of subtidal, intertidal and supratidal habitats, primary data derived from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD) were utilized.
-However, to address data gaps for some habitats, results from special studies were integrated with the primary data sources.
-These included hard bottom [@Kaufman17; @CSA19], dredged holes [@Griffen05; @Raulerson19] and oyster habitat [@OKeife06]*.*
+For the majority of subtidal, intertidal and supratidal habitats, coverages were quantified from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD).
+For subtidal habitats, 2018 source data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19].
+Historical datasets for subtidal habitats begin in 1988 with updates occurring on an approximate bi-annual basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.
 
-The source data used to estimate the most current coverage of seagrasses, tidal flats, and oysters was the *Seagrass in 2018* geospatial database [@SWFWMD19].
-The bi-annual seagrass monitoring program was initiated in 1988 under SWFWMD's Surface Water Improvement and Management program [@Sherwood17; @Tomasko20]*.* SWFWMD has estimated oyster bed coverage as part of this program since 2014.
+Current intertidal and supratidal habitat coverages were estimated using the 2017 SWFWMD Land Use Land Cover (LULC) geospatial database [@SWFWMD18].
+Land use and cover types (natural and developed) are classified pursuant to the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually.
+While the photointerpretation of specific freshwater wetland types is often difficult, forested wetlands from non-forested wetlands can be distinguished with these data.
+Therefore, all applicable FLUCCS codes representing the suite of natural freshwater wetlands were combined within those two classifications.
+Native upland habitats were also combined in one classification.
+Historical estimates for these habitat coverages were also quantified starting with the earliest database in 1990 and occurring every two to three years until the current estimate in 2017.
 
-The source data used to estimate and map trends in development, emergent tidal wetlands, freshwater wetlands, and native upland habitats was the SWFWMD Land Use Land Cover (LULC) series geospatial database [@SWFWMD18].
-This comprehensive database classifies the land use and cover types (natural and developed) pursuant to the Florida Land Use Cover and Forms Classification System [@FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually.
-While the photointerpretation of specific freshwater wetland types is often very difficult, it is possible to accurately distinguish forested wetlands from non-forested wetlands.
-Therefore, for this analysis, all applicable FLUCCS codes representing the suite of natural freshwater wetlands were combined within those two classifications.
-Similarly, within the target- and goal-setting exercise, uplands were combined in one classification.
-These classifications were reported for each mapping exercise conducted every 2-3 years from the start of the program 1990 through 2017, the most recent year with available data.
+To address data gaps for habitats not included in the routine SWFWMD datasets, results from special studies were compiled to obtain current estimates.
+These included hard bottom subtidal habitats, artificial reefs, tidal creeks, and living shorelines.
 
-To quantify the extent of tidal creeks, GIS data [@JEI16; @JEI20] was clipped to the Florida Department of Environmental Protection stream segments that were classified as estuarine.
+Current and historical coverage estimates were obtained for all habitat types, including estimates of developed lands, for each year of available data spanning approximately three decades at two-year time steps.
+In addition, a habitat coverage change analysis between the terminal years of data (1988 to 2018 for subtidal, 1990 to 2017 for intertidal and supratidal) was conducted to understand how habitats were changing between types.
+This required an intersection of the data layers to quantify if habitat types were unchanged or changed for any given location and identifying the type of change (e.g., seagrass to tidal flats).
+The results were summarized as Alluvial diagrams showing relative proportions of habitat change by type and between years.
 
@@
 
-We quantified past restoration efforts in each of the major habitat types to guide decisions on future targets and goals.
+Past restoration projects were quantified for each of the major habitat types to inform expectations for setting short-term goals and long-term targets.
 Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years were compiled, reviewed, and consolidated into a single, consistent geospatial database.
-Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act (GPRA) reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the TBEP.
-Primary data collected included project name, year, description, lead partner, size (area or length), and latitude and longitude.
-Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and documenting the knowledge of local professionals.
+Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act (GPRA) reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the Tampa Bay Estuary Program.
+The collected data included project name, year, description, lead partner, size (area or length), and latitude and longitude.
+Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and expert knowledge form local professionals.
 Living shoreline projects, including seawall enhancements and oyster reef modules, were inventoried separately.
@@
 
-@fig-gisworkflow
+Spatially-explicit estimates of the remaining opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow.
+The two main processes included binning existing datasets into relevant categories and overlaying multiple datasets to identify restoration opportunities.
+Opportunity areas were defined as locations where habitat protection and restoration activities are possible, and where they should best be focused to attain defined targets.
+The definition and mapping of opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible under current and future projected conditions, related both to land use change and sea-level rise.
+
+The Land Use Land Cover 2017 dataset from SWFWMD was used as the baseline for binning existing coverages into the relevant habitat types in the intertidal and subtidal strata.
+
+ All FLUCCS classification codes were placed into one of three categories.
+First, native habitats cover the full range of natural plant communities and other habitats that are endemic to the Tampa Bay watershed, and were further grouped into three major habitat types (tidal wetlands, freshwater wetlands, and uplands).
+Second, restorable habitats include existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through the restoration of more natural hydrology, soils strata, and/or topography.
+Third, existing development includes developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration activities.
 
@@
 These habitats provide important aquifer recharge and wildlife habitat functions.
-
-Opportunity areas, defined here as locations where habitat protection and restoration activities are possible, and where they should best be focused to attain defined targets, were also analyzed.
-The definition and mapping of opportunity areas is necessary to quantify the "restoration potential" for a particular habitat type, which is a measure of what is actually possible under current and future projected conditions.
-The most appropriate opportunity areas are generally not developed and located on existing public lands or areas identified for acquisition.
-
-The Land Use Land Cover 2017 geospatial database (Table 1) was used as the baseline for cataloguing existing and opportunity areas for intertidal and supratidal habitats in the Tampa Bay watershed.
-All FLUCCS classification codes were placed into one of three categories.
-First, native habitats cover the full range of natural plant communities and other habitats that are endemic to the Tampa Bay watershed, and were further grouped into three major habitat types (tidal wetlands, freshwater wetlands, and uplands).
-Second, restorable habitats include existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through the restoration of more natural hydrology, soils strata, and/or topography.
-Third, existing development includes developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration activities.
 

### 2022-08-29 manu-draft.qmd
methods revision done

@@
 
-Addressing the full suite of habitats in a watershed is critical for large-scale restoration planning efforts [@Palmer09; @Lamb18]**.** The major habitat types of Tampa Bay were stratified relative to tidal influence and location in the watershed.
-Subtidal habitats included those that are submerged all or most of the time; intertidal habitats included emergent tidal wetlands that are submerged during high tides but exposed during low tides; and supratidal habitats included those that occur above the high tide line (i.e., the reaminder of the watershed).
+The major habitat types of Tampa Bay were stratified relative to tidal influence and location in the watershed.
+Subtidal habitats included those that are submerged all or most of the time; intertidal habitats included emergent tidal wetlands that are submerged during high tides but exposed during low tides; and supratidal habitats included those that occur above the high tide line (i.e., the remainder of the watershed).
 
-Subtidal included hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13].
-Mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15] were classified as emergent tidal wetlands.
+Subtidal habitats included hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13].
+Intertidal habiats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15].
 Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below).
+This 5-foot contour area was considered the most likely to be impacted by sea-level rise based on current estimates.
+Since 1946, the St.
+Petersburg tidal gauge (NOAA gauge 872650) has documented a nearly 20 cm increase in mean tidal height to present day.
+Projections to the year 2100 suggest sea levels can increase from 58 to 259 cm in the region [@tbep0519].
 
@@
 
-Opportunity areas for restoration and recommended coverage targets for habitat types were identifed by integrating multiple datasets available for the region.
-Habitat status and historical trends were first quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time.
-Historic habitat restoration efforts conducted in the Tampa Bay watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners.
-Finally, remaining opportunities for restoration that integrated results from the first two analyses were identified based on current status of habitat coverage in the watershed.
-This included combining current coverages with existing protected areas and considering their location within the three strata (sub-, inter-, supratidal) and areas anticipated to be impacted by sea-level rise and development near coastal areas.
-The final results included spatially-explicit maps that identified opportunity areas and restoration potential for remaining native habitats in the watershed.
-Short-term (2030) goals and long-term (2050) targets for coverage (hectares) of native habitats were informed by current status, past trends, and historical restoration efforts in the region.
+Opportunity areas for restoration and recommended coverage targets for habitat types were identified by integrating multiple datasets available for the region.
+First, habitat status and historical trends were quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time.
+Second, historic habitat restoration efforts conducted in the Tampa Bay watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners.
+The first two steps were used to identify short-term (2030) goals and long-term (2050) targets for coverage (hectares) of native habitats.
+Finally, remaining opportunities for restoration were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be impacted by sea-level rise.
+As such, the approach identifies goals and targets and provides spatially explicit information that identifies where restoration practitioners should pursue projects to meet these goals and targets.
 
@@
 For subtidal habitats, 2018 source data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19].
-Historical datasets for subtidal habitats begin in 1988 with updates occurring on an approximate bi-annual basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.
+Historical datasets for subtidal habitats began in 1988 with updates occurring on an approximate bi-annual basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.
 
@@
 Land use and cover types (natural and developed) are classified pursuant to the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually.
-While the photointerpretation of specific freshwater wetland types is often difficult, forested wetlands from non-forested wetlands can be distinguished with these data.
+While the photointerpretation of specific freshwater wetland types is often difficult, forested wetlands and non-forested wetlands can be distinguished with these data.
 Therefore, all applicable FLUCCS codes representing the suite of natural freshwater wetlands were combined within those two classifications.
@@
 To address data gaps for habitats not included in the routine SWFWMD datasets, results from special studies were compiled to obtain current estimates.
-These included hard bottom subtidal habitats, artificial reefs, tidal creeks, and living shorelines.
+These included hard bottom subtidal habitats, artificial reefs, tidal creeks, and living shorelines, only available as current estimates with no information on historical trends.
 
@@
 This required an intersection of the data layers to quantify if habitat types were unchanged or changed for any given location and identifying the type of change (e.g., seagrass to tidal flats).
-The results were summarized as Alluvial diagrams showing relative proportions of habitat change by type and between years.
+The results were summarized as Alluvial diagrams showing relative proportions of habitat change by type and between years [@Allaire17].
 
@@
 
-Past restoration projects were quantified for each of the major habitat types to inform expectations for setting short-term goals and long-term targets.
-Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years were compiled, reviewed, and consolidated into a single, consistent geospatial database.
-Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act (GPRA) reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the Tampa Bay Estuary Program.
-The collected data included project name, year, description, lead partner, size (area or length), and latitude and longitude.
+Past restoration projects were quantified for each of the major habitat types, when available, to inform expectations for setting short-term goals and long-term targets.
+Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years were compiled, reviewed, and consolidated into a single database.
+Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the Tampa Bay Estuary Program.
+The collected data included project name, year, description, size (area or length), and location (latitude and longitude).
 Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and expert knowledge form local professionals.
-Living shoreline projects, including seawall enhancements and oyster reef modules, were inventoried separately.
+Living shoreline projects, including seawall enhancements and oyster reef installations, were inventoried separately.
 
@@
 
-Spatially-explicit estimates of the remaining opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow.
-The two main processes included binning existing datasets into relevant categories and overlaying multiple datasets to identify restoration opportunities.
-Opportunity areas were defined as locations where habitat protection and restoration activities are possible, and where they should best be focused to attain defined targets.
-The definition and mapping of opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible under current and future projected conditions, related both to land use change and sea-level rise.
+Spatially-explicit estimates of the opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow.
+The two main processes included 1) binning existing datasets into relevant categories and 2) overlaying multiple datasets to identify restoration opportunities.
+Opportunity areas were defined as locations where habitat protection and restoration activities are possible and where they could occur to attain defined goals and targets.
+The definition and mapping of opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible under current and future projected conditions, related to anticipated land use change and sea-level rise.
 
-The Land Use Land Cover 2017 dataset from SWFWMD was used as the baseline for binning existing coverages into the relevant habitat types in the intertidal and subtidal strata.
+The Land Use Land Cover 2017 dataset from SWFWMD was used for binning existing coverages into the relevant habitat types in the intertidal and subtidal strata.
+All FLUCCS classification codes were placed into one of three categories.
+First, native habitats were those that included the full range of natural plant communities and other habitats that are endemic to the watershed, and were further grouped into three major habitat types (tidal wetlands, freshwater wetlands, and uplands).
+Second, restorable habitats included existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through restoration.
+Third, existing development included developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration.
 
- All FLUCCS classification codes were placed into one of three categories.
-First, native habitats cover the full range of natural plant communities and other habitats that are endemic to the Tampa Bay watershed, and were further grouped into three major habitat types (tidal wetlands, freshwater wetlands, and uplands).
-Second, restorable habitats include existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through the restoration of more natural hydrology, soils strata, and/or topography.
-Third, existing development includes developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration activities.
+After binning, the native and restorable lands were overlaid with additional layers to identify 1) reservation native and reservation restorable areas, and 2) existing and proposed native and restorable areas.
+Collectively, each of these unique products are considered the opportunity areas in the Tampa Bay watershed.
+The reservation native and reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum.
+This area extends from the local Mean Lower Low Water elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation [@tbep0519].
+Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands could be restored.
 
-Three strata, distinct geographic breakpoints in the watershed where different habitat management and restoration activities could take place, were created to establish the process for target and goal-setting within this framework.
-The first is the coastal stratum, which extends from the local Mean Lower Low Water elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation by 2070.
-The coastal stratum is the zone where emergent tidal wetland restoration would be conducted and includes low lying coastal uplands which serve as important tidal wetland buffers that will be critically important in the future as lands reserved to accommodate ideal wetland migration in response to sea-level rise.
-The river floodplain stratum includes all hydrologically contiguous forested and non-forested wetlands within the river and stream corridors of the Tampa Bay watershed.
-Floodplain corridors provide vital watershed functions including fish and wildlife habitat and migratory pathways, floodwater attenuation and storage, erosion control, and delivery of complex organic matter to the estuarine food web.
-Finally, the upland stratum encompasses those areas outside of the coastal and river floodplain strata, including native upland habitats as well as hydrologically isolated wetlands.
-These habitats provide important aquifer recharge and wildlife habitat functions.
+The existing and proposed native and restorable areas were those that occured in public lands and parcels currently acquired or targeted for acquisition.
+To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory, consulting staff from various federal, state and local entities, and inventorying conservation and drainage easements data.
+This workflow created the existing conservation and proposed (i.e., planned to be acquired) conservation layers in @fig-gisworkflow .
+Intersecting the native and restorable lands in these areas produced four unique opportunity areas: existing conservation native, proposed conservation native, existing conservation restorable, and proposed conservation native.
 
-Layers for existing public lands and parcels targeted for acquisition were compiled by combining data from the Florida Natural Areas Inventory (Table 1), consulting staff from various federal, state and local entities, and inventorying conservation and drainage easements data.
+All opportunity areas identified as restorable included reservation restorable (in the coastal stratum), existing conservation restorable, and proposed conservation restorable.
+These layers were further grouped into their restoration potential as discrete habitat types that could be the goal of future restoration projects.
+The restoration potential was defined using the coastal stratum, soil types, and salinities of adjacent waters.
 
 Compared to vegetation communities, soil characteristics typically change slowly (e.g., decades to centuries) in response to hydrologic impacts, unless physically disturbed [@Osland12; @Stockmann14].
-Therefore, soils distributions can be used to generally represent historical habitat distribution, and can be used to provide generalized restoration guidelines (e.g., tidal wetlands, freshwater wetlands, and native uplands).
-@Ries14 created a soils suitability analysis for wetland mitigation and restoration sing data from the USDA Web Soil Survey (Table 1) and classified all soils in the Tampa Bay soils into one of three categories (xeric, mesic, and hydric).
+Therefore, soil distributions can be used to estimate historical habitat distribution and restoration potential.
+@Ries14 created a soils suitability analysis for wetland mitigation and restoration and classified all soils in the Tampa Bay watershed as xeric, mesic, or hydric.
 The mesic and hydric categories were combined to represent wetland restoration potential, while the xeric category was used to represent upland restoration potential.
 
-A distinction was made between tidal and freshwater wetland restoration potential by intersecting the combined mesic and hydric soils polygons with the coastal stratum.
-Mesic or hydric soils that occur below the 5-foot contour were classified as having tidal wetland restoration potential, while mesic or hydric soils occurring above the 5-foot contour were classified as having freshwater wetland restoration potential.
+A distinction was made between tidal and freshwater wetland restoration potential by intersecting the mesic and hydric soils with the coastal stratum.
+Mesic or hydric soils that occur below the 5-foot contour were assigned a restoration potential for tidal wetlands, while mesic or hydric soils occurring above the 5-foot contour were assigned a restoration potential for freshwater wetlands.
+A final restoration potential category was created for low salinity salt marshes (*Juncus* spp.) in the coastal stratum.
+These included restorable lands with mesic or hydric soils and also lands adjacent to waters with average salinity values less than 18 psu [@Eleuterius79].
+The salinity values were identified using salinity isohalines created from a long-term water quality data set provided by the Environmental Protection Commission of Hillsborough County.
 
-Within the Tampa Bay region, lands adjacent to waters with average salinity values greater than 18 psu are considered most appropriate for higher salinity mangrove/salt barren restoration, while lands adjacent to waters with average salinity values less than 18 psu are considered most appropriate for lower salinity salt marsh (*Juncus* spp.) restoration [@Eleuterius79].
-To estimate the relative restoration potential of mangrove/salt barrens and salt marshes, a regional long-term water quality data set was used to create salinity isohalines, which was then binned into two salinity categories: greater and less than an annual mean of 18 psu (Table 1).
-
-All acquired and created GIS layers (Table 1) were converted to 10m x 10m raster data sets and delimited using a previously established watershed boundary.
-The reservation, restorable, conservation, acquisition, soils, and salinity isohaline layers were then merged to create a comprehensive set of maps and matrices of all possible combinations of opportunity areas.
-Within an ArcGIS Pro 2.x GIS environment, the SWFWMD 2017 FLUCCS and 2018 submerged data was classified and extracted to the study area into described categories.
-The dataset was then spatially assigned to existing and proposed restoration areas, reservation areas, soil type, and salinity level.
+Two distinct mapping products were created from the above analysis.
+The first was an opportunities map that showed areas in the watershed identified as existing conservation native, existing conservation restorable, proposed conservation native, proposed conservation restorable, reservation native, and reservation restorable.
+The second was a map that identified the restorable lands (either proposed, existing, or reservation) based on their restoration potential as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands (which included salt marshes).
+All analyses were conducted using the R statistical programming language [@RCT22], specifically leveraging tools from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [@Pebesma18].
+The workflows and data are provided in an open-access repository available on GitHub (<https://github.com/tbep-tech/hmpu-workflow>) for use by others [@Beck22].
 

### 2022-08-30 manu-draft.qmd
habitat status and trend results text revised

@@
 
-#### Subtidal habitats
+Current estimates for subtidal habitats without historical trend information included 171 ha for natural hard bottom and 67 hectares for artificial reefs.
+The former estimate was obtained from previous mapping work to locate and quantify these habitats using sonar mapping [@CSA19].
+The latter estimate was obtained from county governments that manage artificial reefs in the bay.
 
-@fig-subtchgdatalluv
+Current estimates and trend information were available for seagrasses, tidal flats, and oyster bars (Table 1).
+Oyster bars were estimated at 67 ha in 2018 (Table 1), showing a 29% increase since mapping began in 2014.
+This increase may represent improved ground-truthing and photointerpretation of oyster bars from aerial photography.
+Tidal flats have generally increased in area from 1988 to the mid-2000s, followed by a decrease to 2018.
+The current estimate for tidal flats is 6,569 hectares, showing a 24% decline compared to the 1988 estimate of 8,700 hectares.
+Seagrasses have increased by 75% since 1988 to a current estimate of 6,986 ha.
+The change analysis comparing 1988 to 2018 for subtidal habitats (@fig-subtchgdatalluv) confirmed trends in Table 1 and showed that the seagrass increases were primarily associated with the colonization of non-vegetated areas of tidal flats, as well as unclassified areas of open water.
 
-While no trend information is available, the best cumulative estimate of natural hard bottom extent in Tampa Bay is 171 ha (Table 2).
-Oyster bars covered 69 ha (Table 2) in 2018, and a 30% (6 ha) increase reflected since mapping began in 2014 (Table 3a) probably represents improved ground-truthing and photointerpretation of oyster bar signatures from aerial photography.
-Twelve artificial reefs in Tampa Bay are managed by Hillsborough, Manatee, and Pinellas Counties.
-Surface area estimates were not available for the Manatee and Pinellas County reefs, but assuming an average size of 4.2 ha, based on the Hillsborough County reefs, the total coverage of artificial reefs in Tampa Bay is estimated to be approximately 67 ha (Table 2).
-
-Table 1
-
-Total seagrass coverage has increased by 7,027 ha (75%) during the 30-year period of record (Table 3a), and the most current (2018) estimate of total seagrass meadow coverage in Tampa Bay is 16,452 ha.
-The 2018 coverage of tidal flats and sand other than beaches in Tampa Bay was 6,564 ha (Table 2).
-A decrease of 4,496 ha (55%) during the 30-year period of record (Table 3a) is associated with the expansion of seagrass to previously non-vegetated bottom area.
-
-#### Intertidal habitats
-
-@fig-chgdatalluv
-
-Table 2
-
-Between 1990 and 2017, the suite of emergent tidal wetlands (mangroves, salt barrens, and salt marshes) experienced a net gain of 725 ha (10%, Table 3b).
+Total intertidal habitat (mangroves, salt barrens, and salt marshes) showed a 12% increase of 835 ha from 1990 to 2017, with a current estimate of 8,340 ha (Table 2).
+Mangroves increased by 15% to 6,276 ha, salt barrens increased by 7% to 203 ha, and salt marshes increased by 3% to 1,861 ha.
 The current estimate of the extent of mangrove forests in Tampa Bay is 6,192 ha.
@@
 Salt marshes in Tampa Bay cover 1,844 ha, and coverage increased by 30 ha (2%).
-However, from 1990 to 2017, it is estimated that a net area of 219 ha of salt marshes converted from salt marsh to mangrove habitat (<https://tbep-tech.github.io/landuse/>).
-The 2017 estimate of the extent of salt barrens in Tampa Bay is 201 ha, and coverage has increased by 14 ha (7%) during the 27-year period of record (Table 3b).
-Based on GIS data from Janicki Environmental and Mote Marine Laboratory [@JEI16; @JEI20], the extent of tidal creek habitat in the Tampa Bay watershed is approximately 623 km (Table 2).
-No trend analysis is available.
+Despite a net increase in salt marsh habitat, the change analysis showed that 153 ha were changed to mangrove habitat (@fig-chgdatalluv).
+The current extent of tidal tributary length is 622 km (no trend information is available).
 
-#### Supratidal habitats
-
-@fig-chgdatalluv
-
-Table 2
-
-As determined in the 2017 land use/land cover update, the most current estimate of the extent of non-coastal native upland habitats in the Tampa Bay watershed is 56,899 ha, and the extent of "coastal uplands" (defined as below the 5-foot contour) in the Tampa Bay watershed is 1,465 ha (Table 2).
-Over the 27-year period of record, the suite of native upland habitats has experienced a net loss of 37,051 ha (39%, Table 3b).
-
-The 2017 extent of freshwater wetlands in the Tampa Bay watershed was 88,917 ha (Table 2).
-Of this total, forested freshwater wetlands comprised 61,565 ha (69%), while non-forested freshwater wetlands comprised 27,351 ha (31%).
-From 1990-2017, the suite of freshwater wetlands experienced a net gain of 2,444 ha (3%, Table 3b).
-There has been a 5,335 ha (24%) increase in vegetated non-forested freshwater wetlands since 1990, while forested freshwater wetlands have decreased by 2,891 ha (4%).
+Trend assessments for supratidal habitats showed the effects of increasing land development in the Tampa Bay watershed (Table 2).
+Developed lands showed a 44% increase to 217,047 ha from 1990 to 2017, with a notable 30% decrease to 1,446 ha for coastal uplands, a 38% decrease to 57,836 ha for native uplands, and an 18% decrease to 189,512 ha for restorable lands.
+Non-forested freshwater wetlands showed a 24% increase to 27,358 ha, while forested freshwater wetlands decrased by 5% to 61,667 ha.
+The change analysis (@fig-chgdatalluv) showed that a majority of land conversion to developed came from restorable lands (21,292 ha) and native uplands (7,184 ha), with smaller proportions converted from forested freshwater wetlands (1,407 ha) and coastal uplands (193 ha).
+Supratidal habitats converted to restorable primarily included native uplands (8,304 ha), forested freshwater wetlands (1,700 ha), and developed (2,794 ha).
+The increase in non-forested freshwater wetlands was primarily from restorable lands (2,759 ha).
 
@@
 
-![Land use and land cover for the Tampa Bay watershed, Florida, USA.](figs/lulcmap.png){#fig-lulcmap fig-align="center"}
+![Land use and land cover for the Tampa Bay watershed, Florida, USA.
+The watershed includes the natural hydrologic boundary with minor modifications to include partners working with the Tampa Bay Estuary Program.](figs/lulcmap.png){#fig-lulcmap fig-align="center"}
 
@@
 Green indicates existing conservation, blue indicates proposed conservation, and pink indicates reservations opportunities.
-Each category is also grouped into native and restorable habitats.](figs/oppmap.png){#fig-oppmap fig-align="center"}
+Each category is also grouped into native and restorable habitats.
+The outline is the Tampa Bay watershed.](figs/oppmap.png){#fig-oppmap fig-align="center"}
 
@@
 Areas are identified as those where habitat restoration could target the four identified categories as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands.
-Categories are based on the opportunity areas and soil types.](figs/restmap.png){#fig-restmap fig-align="center"}
+Categories are based on the opportunity areas and soil types.
+The outline is the Tampa Bay watershed.](figs/restmap.png){#fig-restmap fig-align="center"}
 

### 2022-08-31 R/dat_proc.R
saving opportunity data for map and manu

@@ -183,3 +183,19 @@ subtchgdatmanu <- bind_rows(op1, op2, op3) %>%
   data.frame(stringsAsFactors = F)
 
 save(subtchgdatmanu, file = here('data', 'subtchgdatmanu.RData'), version = 2)
+
+# opportunity data for opportunity map --------------------------------------------------------
+
+# data copied from commit e6026f831e90cffca8ea906c8fe53c85ceb5746a in hmpu-workflow
+# these are the original HMPU layers in the report
+data("nativersrv")
+data("restorersrv")
+data("nativelyr")
+data("restorelyr")
+data("coastal")
+
+oppdat <- oppdat_fun(nativersrv, restorersrv, nativelyr, restorelyr, coastal)
+
+save(oppdat, file = here('data/oppdat.RData'))
+
+

### 2022-08-31 R/figs.R

@@ -164,16 +164,8 @@ saveNetwork(a2, here("figs/subtchgdatalluv.html"))
 
 # opportunity map -----------------------------------------------------------------------------
 
-# data copied from commit e6026f831e90cffca8ea906c8fe53c85ceb5746a in hmpu-workflow
-# these are the original HMPU layers in the report
-data("nativersrv")
-data("restorersrv")
-data("nativelyr")
-data("restorelyr")
-data("coastal")
 data("tbshed")
-
-oppdat <- oppdat_fun(nativersrv, restorersrv, nativelyr, restorelyr, coastal)
+load(file = here('data/oppdat.RData'))
 
 p <- oppmap_fun(oppdat, tbshed, northloc = 'tr', scaleloc = 'tl',
                 buffdist = 0.04)

### 2022-08-31 manu-draft.qmd
working on opportunity and restoration area results text, added summary objects for areas

@@
 load(file = here('tabs/targtab.RData'))
+load(file = here('data/areas.RData'))
 ```
@@
 
-### Restoration database
+### Restoration and enhancement projects
 
 Past restoration projects were quantified for each of the major habitat types, when available, to inform expectations for setting short-term goals and long-term targets.
-Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years were compiled, reviewed, and consolidated into a single database.
+Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years was compiled from various sources.
+Restoration was defined as any activity that involved earthwork to reshape the land or the addition of structural elements (e.g., rock).
+Enhancement was defined as any activity not including earthwork that improved the environment (e.g., planting native vegetation, invasive species or debris removal, prescribed burns, etc.).
 Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the Tampa Bay Estuary Program.
@@
 
-### Creating and combining opportunity layers
+### Opportunity areas and restoration potential
 
@@
 
-Current estimates for subtidal habitats without historical trend information included 171 ha for natural hard bottom and 67 hectares for artificial reefs.
-The former estimate was obtained from previous mapping work to locate and quantify these habitats using sonar mapping [@CSA19].
-The latter estimate was obtained from county governments that manage artificial reefs in the bay.
-
 Current estimates and trend information were available for seagrasses, tidal flats, and oyster bars (Table 1).
@@
 This increase may represent improved ground-truthing and photointerpretation of oyster bars from aerial photography.
-Tidal flats have generally increased in area from 1988 to the mid-2000s, followed by a decrease to 2018.
+Tidal flats have generally increased from 1988 to the mid-2000s, followed by a decrease to present.
 The current estimate for tidal flats is 6,569 hectares, showing a 24% decline compared to the 1988 estimate of 8,700 hectares.
@@
 The change analysis comparing 1988 to 2018 for subtidal habitats (@fig-subtchgdatalluv) confirmed trends in Table 1 and showed that the seagrass increases were primarily associated with the colonization of non-vegetated areas of tidal flats, as well as unclassified areas of open water.
+Current estimates for subtidal habitats without historical trend information included 171 ha for natural hard bottom and 67 hectares for artificial reefs.
 
@@
 Mangroves increased by 15% to 6,276 ha, salt barrens increased by 7% to 203 ha, and salt marshes increased by 3% to 1,861 ha.
-The current estimate of the extent of mangrove forests in Tampa Bay is 6,192 ha.
-Mangrove forest coverage increased by 684 ha (12%).
-Salt marshes in Tampa Bay cover 1,844 ha, and coverage increased by 30 ha (2%).
-Despite a net increase in salt marsh habitat, the change analysis showed that 153 ha were changed to mangrove habitat (@fig-chgdatalluv).
+Despite a net increase in salt marsh habitat, the change analysis showed that 153 ha were replaced by mangrove habitat (@fig-chgdatalluv).
 The current extent of tidal tributary length is 622 km (no trend information is available).
 
-Trend assessments for supratidal habitats showed the effects of increasing land development in the Tampa Bay watershed (Table 2).
+Trend assessments for supratidal habitats showed the effects of increasing land development and loss of restorable habitats in the Tampa Bay watershed (Table 2).
 Developed lands showed a 44% increase to 217,047 ha from 1990 to 2017, with a notable 30% decrease to 1,446 ha for coastal uplands, a 38% decrease to 57,836 ha for native uplands, and an 18% decrease to 189,512 ha for restorable lands.
-Non-forested freshwater wetlands showed a 24% increase to 27,358 ha, while forested freshwater wetlands decrased by 5% to 61,667 ha.
+Non-forested freshwater wetlands showed a 24% increase to 27,358 ha, while forested freshwater wetlands decreased by 5% to 61,667 ha.
 The change analysis (@fig-chgdatalluv) showed that a majority of land conversion to developed came from restorable lands (21,292 ha) and native uplands (7,184 ha), with smaller proportions converted from forested freshwater wetlands (1,407 ha) and coastal uplands (193 ha).
-Supratidal habitats converted to restorable primarily included native uplands (8,304 ha), forested freshwater wetlands (1,700 ha), and developed (2,794 ha).
+Habitats converted to restorable primarily included native uplands (8,304 ha), forested freshwater wetlands (1,700 ha), and developed (2,794 ha).
 The increase in non-forested freshwater wetlands was primarily from restorable lands (2,759 ha).
 
-### Habitat restoration
+### Habitat restoration and enhancement
 
-A total of 460 projects were documented between 1971 and 2019 (Table 4), addressing the full range of habitat types, including Estuarine (n=228), Freshwater (n=53), Uplands (n=119), and Mixed (n=60).
-A total of 1,978 ha have been restored, and 12,930 ha and 42.8 km of linear projects were enhanced during the time period.
-Forty lead partners were documented as responsible for the projects, although some of these lead partners are departments within the same overall agency.
-Eighty-nine living shoreline projects, seawall enhancements, and oyster reef module installations along shorelines were inventoried, with a linear footprint of 18.2 km.
+A total of 460 restoration projects were documented in Tampa Bay and its watershed between 1971 and 2019.
+These projects were divided among habitat types that included estuarine (n = 228), freshwater (n = 53), uplands (n = 119), and a mix of all three (n = 60).
+A total of 1,978 ha have been restored, whereas 12,930 ha and 42.8 km (as shoreline or tributaries) were enhanced.
+Forty partners were responsible for these projects, although some were from departments within the same agency.
+Eighty-nine living shoreline projects, seawall enhancements, and oyster reef installations were documented, with a total linear length of 18.2 km.
 
-### Habitat restoration opportunities
+### Summary of opportunity areas and restoration potential
 
-@fig-oppmap
+The geospatial analysis that produced a map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap).
+Native lands categorized as existing conservation native, proposed conservation native, or reservation native totaled.
+Similarly, restorable lands categorized as existing conservation restorable, proposed conservation restorable, or reservation restorable totalled.
+Understandably, most of the native and restorable lands occurred in undeveloped areas of the watershed, particularly in the north and southwest.
+Existing conservation lands (existing conservations native, existing conservation restorable) totalled and proposed conservation lands (proposed conservation native, proposed conservation restorable).
+Areas in the coastal stratum (the MLLW to 5-foot contour) included of reservation native habitats and ha of reservation restorable habitats.
 
+Combining the restorable lands in @fig-oppmap with the coastal stratum, soils, and salinity data provided a spatial summary of the restoration potential grouped by habitat type (@fig-restmap).
+The restortation potential
 In 2017, 1,555 km^2^ of land (26.5%) in the Tampa Bay watershed above the MLLW line was classified as natural and 2,144 km^2^ (37%) was considered restorable.
@@
 Reservation lands, from the MLLW line landward to elevation 5-feet (NAVD 88), currently (2017) include 4,622 ha proposed for conservation in the Tampa Bay watershed, with 3,303 ha of native habitats and 1,319 ha of restorable habitats (Fig 1).
-
-Xeric soils were roughly aggregated in an east-west band through the middle of the watershed (Fig 4). Approximately 2,107 km2 was classified as xeric, while 3,760 km2 was classified as mesic/hydric.
 

### 2022-08-31 R/dat_proc.R

@@ -198,4 +198,118 @@ oppdat <- oppdat_fun(nativersrv, restorersrv, nativelyr, restorelyr, coastal)
 
 save(oppdat, file = here('data/oppdat.RData'))
 
+# restoration data for restoration potential map ----------------------------------------------
+
+# data copied from commit e6026f831e90cffca8ea906c8fe53c85ceb5746a in hmpu-workflow
+# these are the original HMPU layers in the report
+data("restorelyr")
+
+restdat <- restdat_fun(restorelyr)
+
+save(restdat, file = here('data/restdat.RData'))
+
+
+# summary of areas in opportunity and restoration potential -----------------------------------
+
+load(file = here('data/oppdat.RData'))
+load(file = here('data/restdat.RData'))
+
+opparea <- oppdat %>%
+  mutate(
+    areaha = st_area(.),
+    areaha = set_units(areaha, 'hectare')
+  ) %>%
+  st_set_geometry(NULL) %>%
+  group_by(cat) %>%
+  summarise(
+    areaha = sum(areaha),
+    areaper = 100 * (areaha / 587200)
+  ) %>%
+  pivot_longer(cols = c('areaha', 'areaper'), names_to = 'var', values_to = 'val')
+
+native <- opparea %>%
+  filter(grepl('Native', cat)) %>%
+  group_by(var) %>%
+  summarise(
+    val = sum(val),
+    .groups = 'drop'
+  ) %>%
+  mutate(
+    val = case_when(
+      var == 'areaha' ~ paste(formatC(round(val, 0), format = 'd', big.mark = ','), 'ha'),
+      var == 'areaper' ~ paste0(round(val, 1), '%')
+    )
+  ) %>%
+  deframe %>%
+  as.list
+restorable <- opparea %>%
+  filter(grepl('Restorable', cat)) %>%
+  group_by(var) %>%
+  summarise(
+    val = sum(val),
+    .groups = 'drop'
+  ) %>%
+  mutate(
+    val = case_when(
+      var == 'areaha' ~ paste(formatC(round(val, 0), format = 'd', big.mark = ','), 'ha'),
+      var == 'areaper' ~ paste0(round(val, 1), '%')
+    )
+  ) %>%
+  deframe %>%
+  as.list
+reservation <- opparea %>%
+  filter(grepl('Reservation', cat)) %>%
+  group_by(var) %>%
+  summarise(
+    val = sum(val),
+    .groups = 'drop'
[…]

### 2022-08-31 R/figs.R

@@ -176,9 +176,7 @@ dev.off()
 
 # restoration potential map -------------------------------------------------------------------
 
-data("restorelyr")
-
-restdat <- restdat_fun(restorelyr)
+load(file = here('data/restdat.RData'))
 
 p <- restmap_fun(restdat, tbshed, northloc = 'tr', scaleloc = 'tl',
                  buffdist = 0.04)

### 2022-08-31 R/dat_proc.R
update opp area summaries to include proposed and existing conservation

@@ -208,11 +208,33 @@ restdat <- restdat_fun(restorelyr)
 
 save(restdat, file = here('data/restdat.RData'))
 
-
 # summary of areas in opportunity and restoration potential -----------------------------------
 
 load(file = here('data/oppdat.RData'))
 load(file = here('data/restdat.RData'))
+
+# summarizes area and formats for manu by type in oppareas
+areasumfun <- function(x, flt){
+
+  out <- x %>%
+    filter(grepl(flt, cat)) %>%
+    group_by(var) %>%
+    summarise(
+      val = sum(val),
+      .groups = 'drop'
+    ) %>%
+    mutate(
+      val = case_when(
+        var == 'areaha' ~ paste(formatC(round(val, 0), format = 'd', big.mark = ','), 'ha'),
+        var == 'areaper' ~ paste0(round(val, 1), '%')
+      )
+    ) %>%
+    deframe %>%
+    as.list
+
+  return(out)
+
+}
 
 opparea <- oppdat %>%
   mutate(
@@ -227,51 +249,11 @@ opparea <- oppdat %>%
   ) %>%
   pivot_longer(cols = c('areaha', 'areaper'), names_to = 'var', values_to = 'val')
 
-native <- opparea %>%
-  filter(grepl('Native', cat)) %>%
-  group_by(var) %>%
-  summarise(
-    val = sum(val),
-    .groups = 'drop'
-  ) %>%
-  mutate(
-    val = case_when(
-      var == 'areaha' ~ paste(formatC(round(val, 0), format = 'd', big.mark = ','), 'ha'),
-      var == 'areaper' ~ paste0(round(val, 1), '%')
-    )
-  ) %>%
-  deframe %>%
-  as.list
-restorable <- opparea %>%
-  filter(grepl('Restorable', cat)) %>%
-  group_by(var) %>%
-  summarise(
-    val = sum(val),
-    .groups = 'drop'
-  ) %>%
-  mutate(
-    val = case_when(
-      var == 'areaha' ~ paste(formatC(round(val, 0), format = 'd', big.mark = ','), 'ha'),
-      var == 'areaper' ~ paste0(round(val, 1), '%')
-    )
-  ) %>%
-  deframe %>%
-  as.list
-reservation <- opparea %>%
-  filter(grepl('Reservation', cat)) %>%
-  group_by(var) %>%
-  summarise(
-    val = sum(val),
-    .groups = 'drop'
-  ) %>%
-  mutate(
-    val = case_when(
[…]

### 2022-08-31 manu-draft.qmd
adding area estimates to text for opp area summary

@@
 The reservation native and reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum.
-This area extends from the local Mean Lower Low Water elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation [@tbep0519].
+This area extends from the local Mean Lower Low Water (MLLW) elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation [@tbep0519].
 Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands could be restored.
@@
 
-The geospatial analysis that produced a map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap).
-Native lands categorized as existing conservation native, proposed conservation native, or reservation native totaled.
-Similarly, restorable lands categorized as existing conservation restorable, proposed conservation restorable, or reservation restorable totalled.
+The map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap).
+Native lands (existing conservation native, proposed conservation native, reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW).
+Similarly, restorable lands (existing conservation restorable, proposed conservation restorable, reservation restorable) totalled `r areas$restorable$areaha` (`r areas$restorable$areaper` of watershed).
 Understandably, most of the native and restorable lands occurred in undeveloped areas of the watershed, particularly in the north and southwest.
-Existing conservation lands (existing conservations native, existing conservation restorable) totalled and proposed conservation lands (proposed conservation native, proposed conservation restorable).
-Areas in the coastal stratum (the MLLW to 5-foot contour) included of reservation native habitats and ha of reservation restorable habitats.
+Existing conservation lands (existing conservation native, existing conservation restorable) totalled `r areas$existing$areaha` (`r areas$existing$areaper` of the watershed) and proposed conservation lands (proposed conservation native, proposed conservation restorable) totalled `r areas$proposed$areaha` (`r areas$proposed$areaper` of the watershed).
+Reservation areas in the coastal stratum (reservation native, reservation restorable) totalled `r areas$reservation$areaha` (`r areas$reservation$areaper` of the watershed).
 
 Combining the restorable lands in @fig-oppmap with the coastal stratum, soils, and salinity data provided a spatial summary of the restoration potential grouped by habitat type (@fig-restmap).
-The restortation potential
-In 2017, 1,555 km^2^ of land (26.5%) in the Tampa Bay watershed above the MLLW line was classified as natural and 2,144 km^2^ (37%) was considered restorable.
-Developed areas in the Tampa Bay watershed encompassed 2,172 km^2^ (36.5%) of the 5,872 km^2^ watershed area.
-Between 2014 and 2017, the developed footprint increased by seven percent (7%, Fig 2).
-
-The Tampa Bay watershed included a total of 1,260 km^2^ of existing conservation lands, either publicly owned or in conservation easements.
-However, excluding subtidal areas owned by the State of Florida results in a total of 816 km^2^ of conservation lands, about 13.9% of the watershed area occurring above the MLLW line.
-(Fig 3).
-Proposed conservation lands in the Tampa Bay watershed total 1,254 km^2^.
-The mapped proposed conservation lands generally link eastward and provide wildlife connectivity to the larger-scale Florida Wildlife Corridor (Fig 3).
-
-Reservation lands, from the MLLW line landward to elevation 5-feet (NAVD 88), currently (2017) include 4,622 ha proposed for conservation in the Tampa Bay watershed, with 3,303 ha of native habitats and 1,319 ha of restorable habitats (Fig 1).
-
-@fig-restmap
 

### 2022-09-01 manu-draft.qmd
results revision done, simplified current extent and target tables

@@
 Petersburg tidal gauge (NOAA gauge 872650) has documented a nearly 20 cm increase in mean tidal height to present day.
-Projections to the year 2100 suggest sea levels can increase from 58 to 259 cm in the region [@tbep0519].
+Projections to the year 2100 suggest sea levels can increase from 58 to 259 cm in the region [@Burke19].
 
@@
 Second, historic habitat restoration efforts conducted in the Tampa Bay watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners.
-The first two steps were used to identify short-term (2030) goals and long-term (2050) targets for coverage (hectares) of native habitats.
+The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for coverage (hectares) of native habitats.
 Finally, remaining opportunities for restoration were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be impacted by sea-level rise.
@@
 All opportunity areas identified as restorable included reservation restorable (in the coastal stratum), existing conservation restorable, and proposed conservation restorable.
-These layers were further grouped into their restoration potential as discrete habitat types that could be the goal of future restoration projects.
-The restoration potential was defined using the coastal stratum, soil types, and salinities of adjacent waters.
+To identify discrete habitat types that could be the goal of future restoration projects, restorable lands in the coastal stratum and on existing conservation areas (reservation restorable and existing conservation restorable) were further grouped into their restoration potential.
+Proposed conservation areas were excluded from the analysis to provide a more confident assessment of restoration potential in areas that have already been acquired (i.e., existing conservation) or are immediately threatened by sea-level rise and/or coastal development (i.e., the coastal stratum).
+The restoration potential was defined using soil types, described as follows.
 
@@
 The mesic and hydric categories were combined to represent wetland restoration potential, while the xeric category was used to represent upland restoration potential.
-
 A distinction was made between tidal and freshwater wetland restoration potential by intersecting the mesic and hydric soils with the coastal stratum.
 Mesic or hydric soils that occur below the 5-foot contour were assigned a restoration potential for tidal wetlands, while mesic or hydric soils occurring above the 5-foot contour were assigned a restoration potential for freshwater wetlands.
-A final restoration potential category was created for low salinity salt marshes (*Juncus* spp.) in the coastal stratum.
-These included restorable lands with mesic or hydric soils and also lands adjacent to waters with average salinity values less than 18 psu [@Eleuterius79].
-The salinity values were identified using salinity isohalines created from a long-term water quality data set provided by the Environmental Protection Commission of Hillsborough County.
 
@@
 The second was a map that identified the restorable lands (either proposed, existing, or reservation) based on their restoration potential as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands (which included salt marshes).
-All analyses were conducted using the R statistical programming language [@RCT22], specifically leveraging tools from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [@Pebesma18].
+All analyses were conducted using the R statistical programming language [@RCT22], specifically leveraging functions from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [@Pebesma18].
 The workflows and data are provided in an open-access repository available on GitHub (<https://github.com/tbep-tech/hmpu-workflow>) for use by others [@Beck22].
@@
 
+Summaries of the current extent of each habitat type are shown in Table 3 as representations of the opportunity areas and restoration potential.
+The extent of each habitat in existing conservation lands and proposed conservation lands is shown.
+Summaries of the restoration potential under existing and proposed conservation lands is also shown.
+The majority of restoration opportunities on existing conservation lands are for native uplands and freshwater wetlands.
+Less opportunities exist for intertidal wetlands (mangrove forests, salt barrens, and salt marshes).
+These summaries are also shown spatially in @fig-oppmap and @fig-restmap.
+
 The map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap).
 Native lands (existing conservation native, proposed conservation native, reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW).
-Similarly, restorable lands (existing conservation restorable, proposed conservation restorable, reservation restorable) totalled `r areas$restorable$areaha` (`r areas$restorable$areaper` of watershed).
-Understandably, most of the native and restorable lands occurred in undeveloped areas of the watershed, particularly in the north and southwest.
+Similarly, restorable lands (existing conservation restorable, proposed conservation restorable, reservation restorable) totalled `r areas$restorable$areaha` (`r areas$restorable$areaper` of the watershed).
+Understandably, most of the native and restorable lands occurred in undeveloped areas in the north and southwest.
 Existing conservation lands (existing conservation native, existing conservation restorable) totalled `r areas$existing$areaha` (`r areas$existing$areaper` of the watershed) and proposed conservation lands (proposed conservation native, proposed conservation restorable) totalled `r areas$proposed$areaha` (`r areas$proposed$areaper` of the watershed).
@@
 
-Combining the restorable lands in @fig-oppmap with the coastal stratum, soils, and salinity data provided a spatial summary of the restoration potential grouped by habitat type (@fig-restmap).
+Combining the restorable lands on existing conservation areas and in the coastal stratum with soils data provided a spatial summary of the restoration potential grouped by habitat type (@fig-restmap).
+A total of `r areas$potentialall$areaha` (`r areas$potentialall$areaper` of the watershed) were identified as potentially restorable as coastal uplands (`r areas$potential[['Coastal Uplands']]$areaha`, `r areas$potential[['Coastal Uplands']]$areaper` of the watershed), freshwater wetlands (`r areas$potential[['Freshwater Wetlands']]$areaha`, `r areas$potential[['Freshwater Wetlands']]$areaper` of the watershed), native uplands (`r areas$potential[['Native Uplands']]$areaha`, `r areas$potential[['Native Uplands']]$areaper` of the watershed), or tidal wetlands (`r areas$potential[['Tidal Wetlands']]$areaha`, `r areas$potential[['Tidal Wetlands']]$areaper` of the watershed).
 
-Table 3
+### Establishment of targets and goals
 
-Integration of all of the opportunity layers provides a summary of the available restoration for the TBEP habitats of interest (Table 5).
-The "Native Habitats" columns show the total current extent as well as the portion of the current extent occurring on existing conservation lands and proposed conservation lands, respectively.
-The "Restorable Habitats" columns show the total restoration opportunity as well as the portion of the total restoration opportunity on existing and proposed conservation lands.
-The majority of restoration opportunities on existing conservation lands are for native uplands and freshwater wetlands (Table 5).
-However, there are approximately 627 ha of emergent tidal wetland restoration opportunities on existing conservation lands, including 530 ha applicable to higher salinity mangrove forests and salt barrens, and 16 ha applicable to lower salinity salt marsh (e.g., *Juncus roemerianus*) restoration and creation.
-
-The best estimates of total restoration opportunities for urban shorelines and tidal tributaries are provided by the Tampa Bay Living Shoreline Suitability Model (LSSM) prepared by the Florida Fish and Wildlife Conservation Commission [@Boland18].
-There is 2,566 km of shoreline in Tampa Bay, and approximately 33% is recommended for living shoreline enhancement.
-
-### Establishment of goals and targets
-
-Table 4
-
-Recommended targets were based on habitat status and trends, habitat restoration data, identified restoration opportunities, and current and anticipated trends in development, available funding, and regulations [@Hilderbrand05; @Hobbs07].
-
-The targets and goals (Table 6) identify where the 2017 CCMP Bay Habitat goals and strategies and the 2021-2025 Strategic Plan [@Burke20] thriving habitats and abundant wildlife programmatic priorities can be implemented.
-However, it also recognizes that the identified habitat protection and restoration areas will change over time, and will be revisited on a 10-year recurring cycle.
-A 30-year planning horizon (2050) is also identified based upon sea-level rise projections developed specifically for Tampa Bay [@Burke19].
-The coastal stratum (from the existing mean low water line to the approximate 5-foot contour) is projected to directly experience the effects of sea-level rise by 2050, and is the primary focus area for coastal habitat protection and restoration activities.
-Land acquisition or protection (through conservation easements or other mechanisms) will be needed to ensure completion of targets and goals for both salt marsh and upland habitats (Fig 3).
-
-Targets that maintain current coverage ("hold-the line strategies") were identified for habitats that appear to currently be sustained at acceptable levels.
-Evolving information such as an Oyster Habitat Suitability Index [@Boswell12] and ongoing mapping exercises will be used to identify optimal locations to conduct restoration activities that help achieve targets and goals.
-Coordination with the establishment of state-mandated minimum flows [@Munson07] will be necessary for restoration and maintenance of low-salinity salt marsh habitats that will experience higher salinities and rapid transition to mangroves under existing sea-level rise scenarios [@Sherwood14; @Geselbracht15].
+Identifying short-term (2030) targets and long-term (2050) goals for the extent of native habitats in Tampa Bay was informed by the assessment of current extents, past trends, and relative effort for past restoration and enhancement projects.
+Table 4 shows the targets and goals identified through this analysis and the associated rationale.
+For example, the targets and goals are established based on the current extent and the restoration potential.
+If restoration potential exists and coverage restored from past projects suggests a reasonable level of effort, the targets and goals reflect the current extent plus the restoration opportunity.
+Conversely, other habitats with no identified restoration opportunity, or current extents are sufficient (e.g., mangrove forests), were assigned targets and goals similar to the current extent, i.e., these habitats should be protected and restoration is not recommended.
+Implicit in the targets and the goals is recurring re-assessment over time to evaluate progress and adjust expectations as appropriate.
 

### 2022-09-01 R/dat_proc.R

@@ -278,7 +278,7 @@ potential <- restdat %>%
             var == 'areaha' ~ paste(formatC(round(val, 0), format = 'd', big.mark = ','), 'ha'),
             var == 'areaper' ~ paste0(round(val, 1), '%')
           ),
-          val = ifelse(val == '0%', '< 1%', val)
+          val = ifelse(val == '0%', '< 0.1%', val)
         ) %>%
         deframe %>%
         as.list
@@ -286,13 +286,35 @@ potential <- restdat %>%
   ) %>%
   deframe()
 
+potentialall <-  restdat %>%
+  mutate(
+    areaha = st_area(.),
+    areaha = set_units(areaha, 'hectare')
+  ) %>%
+  st_set_geometry(NULL) %>%
+  summarise(
+    areaha = sum(areaha),
+    areaper = 100 * (areaha / 587200)
+  ) %>%
+  pivot_longer(cols = c('areaha', 'areaper'), names_to = 'var', values_to = 'val') %>%
+  mutate(
+    val = case_when(
+      var == 'areaha' ~ paste(formatC(round(val, 0), format = 'd', big.mark = ','), 'ha'),
+      var == 'areaper' ~ paste0(round(val, 1), '%')
+    ),
+    val = ifelse(val == '0%', '< 0.1%', val)
+  ) %>%
+  deframe %>%
+  as.list
+
 areas <- list(
   native = native,
   restorable = restorable,
   existing = existing,
   proposed = proposed,
   reservation = reservation,
-  potential = potential
+  potential = potential,
+  potentialall = potentialall
 )
 
 save(areas, file = here('data/areas.RData'))

### 2022-09-01 R/funcs.R

@@ -399,11 +399,6 @@ curexcmp_fun <- function(cursum, nativesum, restoresum){
         !is.na(val) ~ paste(prettyNum(round(val, 0), big.mark = ','), unis),
         T ~ 'N/A'
       ),
-      val = case_when(
-        (HMPU_TARGETS %in% 'Salt Marshes') & (var %in% c('total restorable', 'restorable Existing',
-                                                         'restorable Proposed')) ~ paste(val, '(JU)'),
-        T ~ val
-      ),
       Category = factor(Category, levels = c('Subtidal', 'Intertidal', 'Supratidal')),
       HMPU_TARGETS = factor(HMPU_TARGETS, levels = levels(strata$HMPU_TARGETS))
     ) %>%
@@ -412,13 +407,11 @@ curexcmp_fun <- function(cursum, nativesum, restoresum){
     mutate(
       `native Existing` = case_when(
         Category == 'Subtidal' ~ `Current Extent`,
-        HMPU_TARGETS == c('Living Shorelines') ~ 'LSSM',
         T ~ `native Existing`
       ),
       `total restorable` = case_when(
         HMPU_TARGETS == 'Seagrasses' ~ '5,719 ha',
         HMPU_TARGETS %in% c('Tidal Flats', 'Oyster Bars') ~ 'I/D',
-        HMPU_TARGETS %in% c('Living Shorelines', 'Tidal Tributaries') ~ 'LSSM',
         T ~ `total restorable`
       ),
       `restorable Existing` = case_when(
@@ -462,9 +455,7 @@ curexcmp_fun <- function(cursum, nativesum, restoresum){
     merge_at(i = 15:16, j = 8, part = 'body') %>%
     add_header_row(colwidths = c(2, 3, 3), values = c('', 'Native Habitats', 'Restorable Habitats')) %>%
     add_footer_lines(values = "") %>%
-    footnote(i = 1, j = 1, sep = "", value = as_paragraph("N/A - Not Applicable; I/D - Insufficient Data; LSSM - Living Shoreline Suitability Model; JU - Potential"), part = 'body', inline = T, ref_symbols = "") %>%
-    footnote(i = 1, j = 2, sep = " ", value = as_paragraph(as_i("Juncus")), part = "body", inline = T, ref_symbols = "") %>%
-    footnote(i = 1, j = 3, sep = " ", value = as_paragraph("Marsh Opportunity"), inline = T, ref_symbols = "") %>%
[…]

### 2022-09-02 manu-draft.qmd
starting discussion revision and additions

@@
 Forty partners were responsible for these projects, although some were from departments within the same agency.
-Eighty-nine living shoreline projects, seawall enhancements, and oyster reef installations were documented, with a total linear length of 18.2 km.
+Eighty-nine living shoreline projects, seawall enhancements, and oyster reef installations were documented, with a total length of 18.2 km.
+Although projects were documented for the whole period of record, few projects were completed prior to 1990.
+From 1990 to 2010 and from 2010 to 2019, an annual mean of 68 ha/yr and 81 ha/yr of habitat was restored, respectively.
+These means were used to define appropriate expectations for future restoration, described below.
 
@@
 The map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap).
-Native lands (existing conservation native, proposed conservation native, reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW).
-Similarly, restorable lands (existing conservation restorable, proposed conservation restorable, reservation restorable) totalled `r areas$restorable$areaha` (`r areas$restorable$areaper` of the watershed).
+Native lands currently protected, proposed for protection, or in the coastal stratum (existing conservation native, proposed conservation native, reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW).
+Similarly, restorable lands currently protected, proposed for protection, or in the coastal stratum (existing conservation restorable, proposed conservation restorable, reservation restorable) totalled `r areas$restorable$areaha` (`r areas$restorable$areaper` of the watershed).
 Understandably, most of the native and restorable lands occurred in undeveloped areas in the north and southwest.
@@
 
-Identifying short-term (2030) targets and long-term (2050) goals for the extent of native habitats in Tampa Bay was informed by the assessment of current extents, past trends, and relative effort for past restoration and enhancement projects.
+Identifying short-term (2030) targets and long-term (2050) goals for the restoration extent of native habitats in Tampa Bay was informed by the assessment of current extents, past trends, and relative effort for past restoration and enhancement projects.
 Table 4 shows the targets and goals identified through this analysis and the associated rationale.
@@
 
-Because of multiple stressors such as encroaching development and climate change, habitat protection and restoration priorities should be tempered and "reality tested" by what is actually possible today, and what is possible in the future.
-Many native and potentially restorable habitats are limited, and there will always be restrictions on the financial resources that can be dedicated to public conservation land acquisition and habitat restoration activities.
-
-This replicable method for setting restoration targets and goals provides a systematic attempt to identify habitat protection and restoration targets that are based on what is actually achievable within those limitations.
-It focuses on existing opportunities for all habitat types, and what is realistically possible in the future, rather than attempting to mimic previous ecological conditions.
+Priorities for habitat restoration should be informed and validated by current assessments and what is possible to achieve in the future.
+These priorities are necessary to define given anticipated impacts of land development and climate change, while also considering competing societal interests for the use of the environment and the reality of limited resources for land acquisition and restoration.
+Our approach balances these tradeoffs by providing a systematic workflow to identify targets and goals that is informed by current extent, past trends, and realistic effort from past projects.
+Further, locations are identified where these targets and goals could be achieved based on existing opportunities for protected or restorable habitats, including areas anticipated be impacted by coastal stressors (i.e., sea level rise and land development).
+This approach departs from previous restoration paradigms by identifying what is possible rather than attempting to recreate historical baselines.
+Methods are also provided using open source tools [@Beck22] that 1) allow for the most current datasets to be synthesized to update progress on achieving targets and goals, and 2) can be used in other locations with similar needs for identifying restoration priorities.
 
@@
 
-When viewed as a whole, the most significant and meaningful trends in the TBEP habitats of interest over the periods of record examined include: 1) the 75 percent gain in seagrasses since 1988; 2) the slight gains in emergent tidal wetlands (10%) and freshwater wetlands (2%) since 1990; and 3) the 39% loss in native upland habitats since 1990.
+Identifying appropriate targets and goals would not have been possible without a detailed assessment of current extent and past trends of native habitats in Tampa Bay and its watershed.
+The most notable trends included 1) the increase of seagrasses by 75%, 2) the increase in emergent tidal wetlands (12%) and freshwater wetlands (24%), and 3) the loss in native uplands (38%).
 
-The intertidal zone in Tampa Bay is currently experiencing dynamic change, driven by sea-level rise and climate change, whereby mangrove forests are outcompeting salt marshes and salt barrens for the available niche space.
-Without increasing the total area of the intertidal zone, restoring a greater coverage of mangroves would reduce the niche space available for salt marshes and salt barrens.
-This phenomenon has been observed throughout the Gulf of Mexico, and has been attributed to both climate change (e.g., fewer freeze events) and sea-level rise [@Comeaux12].
+Seagrass recovery in Tampa Bay is a well-known success story that demonstrated how public-private partnerships can effectively reduce total nitrogen loads into Tampa Bay [@Greening14; @Sherwood17] .
+The nutrient reductions, primarily from point-source controls and advanced wastewater treatment, contributed to improvements in water quality and light environments that were favorable for seagrass growth.
+However, the most recent coverage estimate in 2020 showed a seagrass loss of 18% since the peak in 2016, falling below the target defined herein.
+These data were unavailable at the time the habitat restoration workflow was initially developed and recent seagrass changes have prompted bay managers to assess barriers in achieving the seagrass restoration goal.
+In particular, much of the seagrass losses have occurred in Old Tampa Bay, where recurring algal blooms of *Pyrodinium bahamense* have contributed to water quality decline [@Lopez19] .
+Ongoing research to understand mechanisms for mitigating these blooms will be critical for restoring seagrass in Tampa Bay.
 
-The observed gains in wetlands (Table 3b) are likely a reflection of: 1) the effectiveness of state and federal wetland regulatory programs; and 2) the cumulative gains resulting from, primarily, publicly-funded habitat restoration projects (Table 4).
-Minor gains in some emergent tidal wetlands (e.g., salt barrens) may also be a reflection of the landward expansion of the complex suite of these habitats associated with climate change and sea-level rise.
-Gains in vegetated non-forested freshwater wetlands are related to the clearing of forested wetlands followed by the creation of herbaceous mitigation areas and stormwater systems.
+Emergent tidal and freshwater wetlands in Tampa Bay have also experienced dramatic changes over the last three decades.
+Dual pressures from sea-level rise and climate change have impacted tidal wetlands, such that mangrove forests are outcompeting salt marshes and salt barrens for available niche space.
+Mangrove expansion as a result of climate change has been observed throughout the Gulf of Mexico [@Comeaux12].
+As such, the identified targets and goals for mangroves indicate protection of mangrove habitats, without the need for additional restoration.
+However, mangroves will continue to expand in the intertidal zone contributing to additional losses of salt marshes and salt barrens.
+The reservation areas identified in @fig-oppmap represent critical remaining areas in the intertidal zone that could be protected to prevent additional losses of tidal wetlands.
+Likewise, gains in non-forested freshwater wetlands are a reflection of 1) the effectiveness of state and federal wetland regulatory programs, and 2) the cumulative gains from publicly-funded habitat restoration projects.
+A similar approach could be used for salt marshes or salt barrens.
 
-The decrease in native uplands (Table 3b) is the result of continued development in the Tampa Bay watershed, combined with the lack of regulatory protection of native uplands.
-Attaining the target and goal will require concerted restoration of native upland habitats on existing conservation lands, as well as new conservation lands to offset the continued loss of these habitats to development, and amendments to existing planning, zoning and land development policies or regulations.
-While federal and state regulations related to listed species management impart some protection to certain rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats, such as pine flatwoods, are left largely unprotected.
-Unless local governments in the Tampa Bay watershed improve local protections for native uplands, such as strengthening language within comprehensive plans and development ordinances, this trend will likely continue.
+The decrease in native uplands is the result of continued development in the Tampa Bay watershed (@fig-chgdatalluv) and lack of regulatory protection of these habitats.
+Attaining the target and goal will require restoration of upland habitats on existing conservation lands (i.e., restoration potential in @fig-restmap) and new conservation lands to offset the continued loss of these habitats to development.
+The long-term conversion of restorable lands to developed (@fig-chgdatalluv) presents additional challenges for restoration of native uplands.
+Amendments to existing planning, zoning and land development policies or regulations will be needed to address these issues.
+While federal and state regulations related to listed species management provide some protection to rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats are largely unprotected (e.g., pine flatwoods).
+Reductions in native uplands will likely continue unless local governments improve regulatory protections, such as strengthening language within comprehensive plans and development ordinances.
 
@@
 
+There is a general consensus among restoration practitioners and natural resource managers that habitat restoration and management is most cost-effective on publicly-owned conservation lands and provides long-term benefits.
+Furthermore, given current development trends in the Tampa Bay watershed, public acquisition of remaining critical lands (e.g., coastal uplands, river floodplain wetlands) is a high priority, and some restoration targets (e.g., salt marshes) will not be feasible without additional public acquisition or public-private partnerships [@Holl01; @Benson18].
+Therefore, varied approaches to leverage resources, including traditional grants, partner funding, and use of volunteers for habitat restoration are recommended to maximize the potential for successful target and goal achievement.
 Increases observed in tidal and freshwater wetlands are primarily due to publicly-funded habitat restoration projects, state and federal wetland regulatory programs, and to a lesser extent, regulatory mitigation.
-While restoration activities date to 1971, few projects were completed prior to 1990, and from 1990-2010, an annual mean of 68 ha of habitat was restored.
-Over the past decade (2010-2019), this rate of restoration project completion has increased to over 81 ha/yr.
-Assuming that funding levels remain in the same range as the past decade, this annual mean can be used to set reasonable limits on restoration potential and targets.
+Assuming that funding levels remain in the same range as the past decade, estimates of past restoration effort can be used to set reasonable limits on restoration potential and targets.
+
+Four major types of disturbed sites around the Tampa Bay coastline have been identified as priority estuarine habitat restoration sites by stakeholders over the past two decades, including dredged holes, filled and spoil disposal areas, abandoned aquaculture ponds; and coastal borrow pits and stormwater ponds.
+Substantial opportunities also exist for upland restoration on reclaimed mined lands within the watershed.
 
@@
 
-Four major types of disturbed sites around the Tampa Bay coastline have been identified as priority estuarine habitat restoration sites by TBEP stakeholders over the past two decades, including dredged holes, filled and spoil disposal areas, abandoned aquaculture ponds; and coastal borrow pits and stormwater ponds.
-Substantial opportunities also exist for upland restoration on reclaimed mined lands within the watershed.
-
-There is a general consensus among restoration practitioners and natural resource managers that habitat restoration and management is most cost-effective on publicly-owned conservation lands and provides long-term benefits.
-Furthermore, given current development trends in the Tampa Bay watershed, public acquisition of remaining critical lands (e.g., coastal uplands, river floodplain wetlands) is a high priority, and some restoration targets (e.g., salt marshes) will not be feasible without additional public acquisition or public-private partnerships [@Holl01; @Benson18].
-Therefore, varied approaches to leverage resources, including traditional grants, partner funding, and use of volunteers for habitat restoration are recommended to maximize the potential for successful target and goal achievement.
-
-### Rolling easements, mitigation, and restoration consortium
+### Rolling easements and mitigation
 
@@
 
-TBEP will form a public-private partnership to provide the framework for the development of a coordinated approach linking regulatory (compensatory mitigation) and resource management (publicly funded habitat enhancement, restoration, and establishment) programs in the watershed.
-The creation of the Habitat Management Consortium is expected to provide benefits such as optimizing and improving the cost-effectiveness of habitat protection, restoration and mitigation activities in the watershed.
+### Limitations of the approach
+
+Identifying restoration priorities was data intensive and would not have been possible without the resources available for the region.
+The workflow for identifying priorities required detailed and spatially-explicit datasets specific to the Tampa Bay watershed.
+Long-term datasets describing land use and cover in the watershed and the extent of in-bay habitats were necessary to categorize current extent and past trends.
+Similarly, supporting datasets included those that described existing and proposed conservation areas, soils, past restoration activities, and relevant spatial boundaries (i.e., watershed and coastline).
+Many of these datasets are available outside of Tampa Bay, although temporal and spatial resolutions may limit application to other areas.
+Additionally, considerable effort was made in working with regional partners to identify and fill knowledge gaps for relevant habitat types.
+For example, inventories of hard bottom habitats, living shorelines, tidal tributaries, and artificial reefs were created through special studies or were available only as current estimates from regional entities.
+Tracking progress towards targets in goals for these habitats will require regular updates to these datasets that are not part of routine land use and cover maps.
+
+An additional assumption of the workflow, particularly for tracking progress, is that projects reported by partners will ultimately manifest in map products.
+Specifically, restoration effort by regional partners is catalogued in the available restoration database, which not only depends on voluntary reporting, but also represents a source of information on restoration extent that is separate from land cover maps.
+An expectation is that acreage restored for a given habitat as reported by a partner will ultimately be shown as a change in land use and cover on regional maps.
+The temporal lag between an actual project and how it may be reflected in a GIS product is unknown, which may create a disconnect between the updates in achieving targets and goals as new data layers are released and the effort reported by partners.
+The spatial resolution of mapping products may also be insufficient to detect habitat changes as reported in the restoration database.
+For these reasons, projects reported by partners are currently summarized separately from the assessments above that depend on GIS layers.
+Additional work is needed to reconcile these datasets for more streamline reporting.
 

### 2022-09-06 manu-draft.qmd
working on restoration text in discussion

@@
 
-Priorities for habitat restoration should be informed and validated by current assessments and what is possible to achieve in the future.
+Priorities for habitat restoration should be informed by current assessments and what is possible to achieve in the future.
 These priorities are necessary to define given anticipated impacts of land development and climate change, while also considering competing societal interests for the use of the environment and the reality of limited resources for land acquisition and restoration.
 Our approach balances these tradeoffs by providing a systematic workflow to identify targets and goals that is informed by current extent, past trends, and realistic effort from past projects.
-Further, locations are identified where these targets and goals could be achieved based on existing opportunities for protected or restorable habitats, including areas anticipated be impacted by coastal stressors (i.e., sea level rise and land development).
+Further, locations are identified where these targets and goals could be achieved based on existing opportunities for protected or restorable habitats, including areas anticipated to be impacted by coastal stressors (i.e., sea level rise and land development).
 This approach departs from previous restoration paradigms by identifying what is possible rather than attempting to recreate historical baselines.
@@
 Identifying appropriate targets and goals would not have been possible without a detailed assessment of current extent and past trends of native habitats in Tampa Bay and its watershed.
-The most notable trends included 1) the increase of seagrasses by 75%, 2) the increase in emergent tidal wetlands (12%) and freshwater wetlands (24%), and 3) the loss in native uplands (38%).
+The most notable trends included 1) the increase of seagrasses by 75%, 2) the increase of emergent tidal wetlands (12%) and freshwater wetlands (24%), and 3) the loss of native uplands (38%).
 
@@
 However, the most recent coverage estimate in 2020 showed a seagrass loss of 18% since the peak in 2016, falling below the target defined herein.
-These data were unavailable at the time the habitat restoration workflow was initially developed and recent seagrass changes have prompted bay managers to assess barriers in achieving the seagrass restoration goal.
+These data were unavailable at the time the habitat restoration workflow was initially developed and trends informed by the new restoration paradigm have prompted bay managers to assess barriers in achieving the seagrass restoration goal.
 In particular, much of the seagrass losses have occurred in Old Tampa Bay, where recurring algal blooms of *Pyrodinium bahamense* have contributed to water quality decline [@Lopez19] .
@@
 Mangrove expansion as a result of climate change has been observed throughout the Gulf of Mexico [@Comeaux12].
-As such, the identified targets and goals for mangroves indicate protection of mangrove habitats, without the need for additional restoration.
+As such, the identified targets and goals for mangroves indicate protection of these habitats, without the need for additional restoration.
 However, mangroves will continue to expand in the intertidal zone contributing to additional losses of salt marshes and salt barrens.
@@
 The long-term conversion of restorable lands to developed (@fig-chgdatalluv) presents additional challenges for restoration of native uplands.
-Amendments to existing planning, zoning and land development policies or regulations will be needed to address these issues.
+Amendments to existing planning, zoning, and land development policies will be needed to address these issues.
 While federal and state regulations related to listed species management provide some protection to rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats are largely unprotected (e.g., pine flatwoods).
@@
 
-There is a general consensus among restoration practitioners and natural resource managers that habitat restoration and management is most cost-effective on publicly-owned conservation lands and provides long-term benefits.
-Furthermore, given current development trends in the Tampa Bay watershed, public acquisition of remaining critical lands (e.g., coastal uplands, river floodplain wetlands) is a high priority, and some restoration targets (e.g., salt marshes) will not be feasible without additional public acquisition or public-private partnerships [@Holl01; @Benson18].
-Therefore, varied approaches to leverage resources, including traditional grants, partner funding, and use of volunteers for habitat restoration are recommended to maximize the potential for successful target and goal achievement.
-Increases observed in tidal and freshwater wetlands are primarily due to publicly-funded habitat restoration projects, state and federal wetland regulatory programs, and to a lesser extent, regulatory mitigation.
-Assuming that funding levels remain in the same range as the past decade, estimates of past restoration effort can be used to set reasonable limits on restoration potential and targets.
+Habitat restoration and management will have long-term benefits and will be most cost-effective on publicly-owned conservation lands.
+Moreover, public acquisition of remaining critical lands (e.g., coastal uplands) is a high priority given current development trends in the Tampa Bay watershed.
+Other restoration targets (e.g., salt marshes) will not be feasible without additional public acquisition or public-private partnerships as a mechanism for doing so [@Holl01; @Benson18].
+Therefore, varied approaches to leverage resources are recommended to achieve the defined targets and goals.
+These approaches could include pursuing traditional grants, matching funds from multiple partners, and volunteer initiatives for habitat restoration.
+Recent gains in tidal and freshwater wetlands are primarily due to publicly-funded habitat restoration projects, state and federal wetland regulatory programs, and to a lesser extent, regulatory mitigation.
+Restoration activities for habitats without similar regulatory frameworks  should pursue the options above to achieve the defined targets and goals.
 
-Four major types of disturbed sites around the Tampa Bay coastline have been identified as priority estuarine habitat restoration sites by stakeholders over the past two decades, including dredged holes, filled and spoil disposal areas, abandoned aquaculture ponds; and coastal borrow pits and stormwater ponds.
-Substantial opportunities also exist for upland restoration on reclaimed mined lands within the watershed.
-
-While existing development areas are not considered feasible for major habitat restoration activities at this time, there are many opportunities to enhance and restore habitat functions and improve coastal resilience in urbanized locations.
+Other restoration activities could be pursued for the opportunity areas identified herein.
+Substantial opportunities exist for upland restoration on reclaimed mined lands within the watershed (@fig-lulcmap, @fig-oppmap).
+For estuarine habitats, opportunity areas could include dredged holes, and filled and spoil disposal areas, either for enhancing existing subtidal habitats or as areas that could be colonized by seagrasses.
+Opportunities also exist on developed lands primarily through enhancement projects, although these have not been explicitly identified in the products created herein.
 Examples include the construction of living shorelines, placement of submerged habitat modules along developed urban shorelines and seawalls, and creation of backyard habitats.
-Tidal tributary restoration could also entail improvements including removal of salinity barriers and filling of dredged channel sections with low dissolved oxygen.
+Tidal tributary restoration could also include removal of salinity barriers and filling of dredged channel sections.
+Overall, restoration practitioners must consider several options and choose those that are most feasible given the available resources and likelihood of success.
 

### 2022-09-07 manu-draft.qmd
discussion revision done, fixes to current and target tables

@@
 author: 
+  - Marcus W Beck:
+      email: mbeck@tbep.org
+      institute: [TBEP]
+      correspondence: true
+  - Douglas E Robison:
+      email: DRobison@esassoc.com 
+      institute: [ESA]
+      correspondence: false
   - Gary E Raulerson:
@@
       correspondence: false
-  - Douglas E Robison:
-      email: DRobison@esassoc.com 
-      institute: [ESA]
-      correspondence: false
-  - Marcus W Beck:
-      email: mbeck@tbep.org
-      institute: [TBEP]
-      correspondence: true
   - Maya C Burke:
@@
 To address these pressures, restoration practitioners require robust tools that identify reasonable goals given historical land use trends, current status of the habitat mosaic, and anticipated future impacts from coastal stressors.
-A target-setting approach for native habitats was created for the Tampa Bay watershed that identifies current restoration opportunities and establishes short (2030) and long-term (2050) goals.
-The approach was informed through a three-decade habitat change analysis and over forty years of habitat restoration experience in the region.
+A target-setting approach for native habitats was created for the Tampa Bay watershed that identifies current restoration opportunities and establishes short-term (2030) targets and long-term (2050) goals.
+The approach was informed through a three-decade habitat change analysis and over forty years of habitat restoration projects in the region.
 Restoration goals were defined based on what is possible today and the projected needs for the future, rather than attempting to replicate past ecological conditions.
 The new paradigm also accounts for the expected impacts of sea-level rise, climate change, and watershed development - stressors which are pervasive in Florida.
-The resulting habitat goals are spatially explicit with maps that identify remaining restoration and conservation opportunities, while also providing an approach for the entire watershed that targets subtidal, intertidal, and coastal uplands.
+The resulting habitat goals are spatially explicit with maps that identify remaining restoration opportunities, while also providing an approach for the entire watershed that targets subtidal, intertidal, and uplands.
 This approach represents a general framework to support coastal planning decisions that need to address competing interests and could be applied in other coastal settings where sustainable urbanization practices need to co-exist with natural environments.
@@
 
-The health of estuarine systems and coastal habitats is inextricably linked to land uses and management of the watershed [@Yoskowitz15].
+The health of estuarine systems and coastal habitats is tightly linked to land use and management of the watershed [@Yoskowitz15].
 Coastal habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18].
@@
 Anthropogenic land development can remove or substantially modify existing habitats, and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14].
-Related changes in habitats in response to climate change (sea-level rise and warming) include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].
-Management of coastal habitats requires an understanding of how stressores impact these resources, while balancing competing societal uses.
+Habitat changes in response to climate change include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].
+Management of coastal habitats requires an understanding of how stressors impact these resources, while balancing competing societal uses.
 
@@
 The coastal squeeze of development and climate change diminishes the available space for future restoration in urban estuaries, thereby impacting the variety of ecosystem services provided by these habitats and the wildlife they support [@Enwright15].
-Given projected habitat losses without management intervention and the limited resources available (i.e., time, land, funding, and labor force), appropriate and realistic sites for restoration need to be identified that account for future stressors and past trends.
+Given projected habitat losses without management intervention and the limited resources available, appropriate and realistic sites for restoration need to be identified that account for future stressors and past trends.
 
@@
 The reservation native and reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum.
-This area extends from the local Mean Lower Low Water (MLLW) elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation [@tbep0519].
+This area extends from the local Mean Lower Low Water (MLLW) elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation [@Burke19].
 Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands could be restored.
@@
 Priorities for habitat restoration should be informed by current assessments and what is possible to achieve in the future.
-These priorities are necessary to define given anticipated impacts of land development and climate change, while also considering competing societal interests for the use of the environment and the reality of limited resources for land acquisition and restoration.
+These priorities are necessary given anticipated impacts of land development and climate change, while also considering competing societal interests for use of the environment and limited resources for land acquisition and restoration.
 Our approach balances these tradeoffs by providing a systematic workflow to identify targets and goals that is informed by current extent, past trends, and realistic effort from past projects.
-Further, locations are identified where these targets and goals could be achieved based on existing opportunities for protected or restorable habitats, including areas anticipated to be impacted by coastal stressors (i.e., sea level rise and land development).
+Further, locations are identified where these targets and goals could be achieved based on existing opportunities for restorable habitats, including areas anticipated to be impacted by coastal stressors (i.e., sea level rise and land development).
 This approach departs from previous restoration paradigms by identifying what is possible rather than attempting to recreate historical baselines.
@@
 
-### Habitat restoration
+### Achieving restoration targets and goals
 
-Habitat restoration and management will have long-term benefits and will be most cost-effective on publicly-owned conservation lands.
-Moreover, public acquisition of remaining critical lands (e.g., coastal uplands) is a high priority given current development trends in the Tampa Bay watershed.
+Achieving the defined targets and goals will require diverse approaches for habitat restoration and management.
+Focusing efforts on publicly-owned conservation lands is expected to have long-term benefits and will be most cost-effective given the level of restoration effort compared to habitats that have already been impacted by anthropogenic activities.
+As such, public acquisition of remaining critical lands (e.g., coastal uplands) is a high priority given current development trends in the Tampa Bay watershed.
 Other restoration targets (e.g., salt marshes) will not be feasible without additional public acquisition or public-private partnerships as a mechanism for doing so [@Holl01; @Benson18].
-Therefore, varied approaches to leverage resources are recommended to achieve the defined targets and goals.
-These approaches could include pursuing traditional grants, matching funds from multiple partners, and volunteer initiatives for habitat restoration.
+Therefore, varied approaches to leverage resources for restoration are needed and could include pursuing traditional grants, matching funds from multiple partners, and volunteer initiatives for habitat restoration.
 Recent gains in tidal and freshwater wetlands are primarily due to publicly-funded habitat restoration projects, state and federal wetland regulatory programs, and to a lesser extent, regulatory mitigation.
@@
 Substantial opportunities exist for upland restoration on reclaimed mined lands within the watershed (@fig-lulcmap, @fig-oppmap).
-For estuarine habitats, opportunity areas could include dredged holes, and filled and spoil disposal areas, either for enhancing existing subtidal habitats or as areas that could be colonized by seagrasses.
-Opportunities also exist on developed lands primarily through enhancement projects, although these have not been explicitly identified in the products created herein.
-Examples include the construction of living shorelines, placement of submerged habitat modules along developed urban shorelines and seawalls, and creation of backyard habitats.
+For estuarine habitats, opportunity areas could include dredged holes or spoil disposal areas, either for enhancing existing subtidal habitats or creating areas that could be colonized by seagrasses.
+Some opportunities also exist on developed lands primarily through enhancement projects, although these have not been explicitly identified in the products created herein.
+Examples include the construction of living shorelines in place of hardened seawalls, placement of submerged habitat modules (e.g., artificial oyster reef) along urban shorelines, and creation of backyard habitats.
 Tidal tributary restoration could also include removal of salinity barriers and filling of dredged channel sections.
@@
 
-### Rolling easements and mitigation
+Creative approaches may be required in areas impacted by sea-level rise if land acquisition is not possible.
+These approaches are necessary to accommodate future landward migration of tidal wetlands or the protection of coastal uplands, while also preventing or minimizing development activities that can inhibit landward habitat shifts.
+Coastal setbacks, buffers, or public easements are traditionally used to restrict development within a given distance from the shoreline.
+However, rolling easements may be an alternative approach whereby protected areas are allowed to "roll" landward with expected changes in sea level rise.
+Rolling easements could disincentivize more intense urban development of low-lying coastal uplands in less developed agricultural or recreational land uses.
+Landowners could maintain current economic uses with a rolling easement, while reserving such lands to accommodate landward habitat migration.
 
-As discussed, land acquisition for coastal habitat restoration must prioritize adjacent low-lying coastal uplands to serve as buffers to accommodate future landward migration of tidal wetlands in response to sea-level rise.
-Where public acquisition is not possible, other conservation mechanisms need to be explored.
-Coastal setbacks, buffers, or public easements are traditionally used to restrict development within a given distance from the shoreline.
-A rolling easement is a dynamic mechanism that "rolls" landward as sea levels rise and cause tidal encroachments onto low-lying coastal uplands [@Titus09].
-The application of rolling easements in Tampa Bay could disincentivize more intense urban development (e.g., discourage up-zoning) of low-lying coastal uplands that may be currently in less intense agricultural or recreational (e.g., golf courses) land uses.
-Under a rolling easement, landowners would be able to maintain current economic uses, while "reserving" such lands to accommodate tidal wetland migration with advancing sea-level rise.
-
-Wetland impacts and associated compensatory mitigation projects authorized under wetland regulatory programs have historically been conducted independent of watershed-level planning and monitoring processes.
-This disconnect has contributed to fragmented implementation and inconsistent compliance monitoring of mitigation projects, as well as historically poor documentation of wetland losses and gains in the Tampa Bay watershed.
-However, if properly focused and comprehensively coordinated, compensatory mitigation activities could significantly contribute to the attainment of wetland habitat restoration goals and targets for the Tampa Bay estuarine system and its contributing watershed.
+Finally, wetland impacts and associated compensatory mitigation projects authorized under wetland regulatory programs could serve as possible restoration mechanisms.
+Mitigation activities have historically been conducted independent of watershed-level planning and monitoring processes.
+This disconnect has contributed to fragmented implementation and inconsistent compliance monitoring of mitigation projects, including historically poor documentation of wetland losses and gains in the Tampa Bay watershed.
+However, if properly focused and coordinated, compensatory mitigation activities could significantly contribute to the attainment of restoration targets and goals for the region.
 
@@
 An additional assumption of the workflow, particularly for tracking progress, is that projects reported by partners will ultimately manifest in map products.
-Specifically, restoration effort by regional partners is catalogued in the available restoration database, which not only depends on voluntary reporting, but also represents a source of information on restoration extent that is separate from land cover maps.
+Specifically, restoration effort by regional partners is cataloged in the available restoration database, which not only depends on voluntary reporting, but also represents a source of information on restoration extent that is separate from land cover maps.
 An expectation is that acreage restored for a given habitat as reported by a partner will ultimately be shown as a change in land use and cover on regional maps.
@@
 
-The establishment of targets and goals considering climate change, development, land availability, and past achievements, expands the restoration palette to a more comprehensive list of habitats within the system.
-If successfully implemented, the 2050 goals would total over 4,000 ha of habitat restoration throughout the Tampa Bay watershed.
-Land acquisition will be an important component of successful completion of different targets and goals, including the salt marsh restoration target for the first ten years.
-Land acquisition will also provide new opportunities for outdoor access, given that these projects often have a public recreation component.
-
-Our new approach will continue to engage multiple partner agencies, non-governmental organizations, and private citizens in the successful implementation of the restoration plan.
-Emphases will include recognition of land types particularly vulnerable to climate change or development stressors and these needs will be communicated to restoration partners.
-Consistent education, targeted funding opportunities, and reporting will also ensure that these newly established targets and goals lead to successful restoration projects, land acquisition, and enhanced ecosystem services.
+The establishment of targets and goals that account for climate change, development trajectories, land availability, and past restoration effort expands the restoration opportunities to a more comprehensive list of habitats for the entire watershed.
+Land acquisition is critical for attaining the defined targets and goals and will also provide new opportunities for outdoor access to the broader community.
+Successful restoration is also contingent on engaging multiple partners, non-governmental organizations, and private citizens.
+The products created herein will guide these efforts for the next thirty years by providing a continuously updated assessment of where the opportunities exist and if targets and goals are expected to be met.
+The Tampa Bay area is not unique in its challenges for protecting native habitats and the approach is readily transferable to other locations where restoration priorities are needed in response to pervasive coastal stressors.
 
@@
 ```{r}
-#| tbl-cap: Summary of habitat restoration opportunities in the Tampa Bay watershed.
+#| tbl-cap: "Summary of habitat restoration opportunities in the Tampa Bay watershed.
 Summaries are based on 2017 land use data, 2018 subtidal data, best estimates for habitat types not in existing GIS layers, and current extent of existing and proposed conservation lands.
+Proposed conservation lands are those identified for acquisition.
+Total restoration opportunity does not account for lands currently existing or proposed for conservation.
+N/A: not applicable, I/D: insufficient data."
 currtab
@@
 ```{r}
-#| tbl-cap: Recommended 2030 targets and 2050 goals for habitat restoration and protection in the Tampa Bay watershed.
+#| tbl-cap: "Recommended 2030 targets and 2050 goals for habitat restoration and protection in the Tampa Bay watershed.
 Targets and goals are based on 2017 land use data, 2018 subtidal data, best estimates for habitat types not in existing GIS layers, and current extent of existing and proposed conservation lands.
+Total restoration opportunity does not account for lands currently existing or proposed for conservation.
+N/A: not applicable, I/D: insufficient data."
 targtab

### 2022-09-07 R/dat_proc.R

@@ -5,6 +5,46 @@ library(units)
 
 source(here('R/funcs.R'))
 
+# targets as metric (ha, mi) ------------------------------------------------------------------
+
+trgsmetric <- structure(
+  list(
+    Category = structure(
+      c(1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 3L),
+      .Label = c("Subtidal", "Intertidal", "Supratidal"),
+      class = "factor"
+    ),
+    HMPU_TARGETS = structure(
+      1:15,
+      .Label = c("Hard Bottom", "Artificial Reefs", "Tidal Flats", "Seagrasses", "Oyster Bars", "Living Shorelines", "Total Intertidal", "Mangrove Forests", "Salt Barrens", "Salt Marshes", "Tidal Tributaries", "Coastal Uplands", "Non-Forested Freshwater Wetlands", "Forested Freshwater Wetlands", "Native Uplands"),
+      class = "factor"
+    ),
+    Target2030 = c(171, 88, 6564, 16188, 87, 34, 8745, 6276, 223, 1962, 628, 1507, 27904, 61728, 58018),
+    Target2050 = c(171, 88, 6564, 16188, 189, 90, 9737, 6276, 324, 2225, 651, 1689, 29058, 61910, 58443),
+    rationale = c(
+      "Protect existing hard bottom; continue to identify new hard bottom area through mapping",
+      "Protect existing artificial reefs; enhance habitat complexity where feasible; expand reef area to promote fish and wildlife benefits",
+      "Identify and protect existing tidal flats; assess restoration potential of other non-vegetated subtidal areas",
+      "Protect existing seagrasses; assess restoration potential of non-vegetated subtidal areas",
+      "2030: Protect existing oysters and restore 20 hectares; increase target by 20 hectares each out-decade",
+      "2030: Construct 1.6 kilometers each year; better define opportunity areas; increase target to 2.4 and 3.2 kilometers per year for out decades",
+      "2030: Protect existing intertidal mosaic and restore 405 hectares; increase target by 61 hectares each out-decade; includes the mosaic of mangrove, salt barren, and salt marsh habitats",
+      "Protect existing mangrove forests; restore opportunistically within the intertidal mosaic",
+      "2030: Protect existing salt barrens and restore 20 hectares; increase target by 20 hectares per out decade",
+      "2030: Protect existing low salinity salt marshes and restore 101 hectares; increase target by 20 hectares each out-decade",
[…]

### 2022-09-07 R/funcs.R

@@ -439,10 +439,10 @@ curexcmp_fun <- function(cursum, nativesum, restoresum){
       Category = 'Stratum',
       HMPU_TARGETS = 'Habitat Type',
       `native Existing` = 'Existing Conservation Lands',
-      `native Proposed` = 'Proposed Conservation Lands*',
-      `total restorable` = 'Total Restoration Opportunity**',
+      `native Proposed` = 'Proposed Conservation Lands',
+      `total restorable` = 'Total Restoration Opportunity',
       `restorable Existing` = 'Existing Conservation Lands Restoration Opportunity',
-      `restorable Proposed` = 'Proposed Conservation Lands Restoration Opportunity*'
+      `restorable Proposed` = 'Proposed Conservation Lands Restoration Opportunity'
     ) %>%
     merge_at(i = 1, part = 'body') %>%
     merge_at(i = 7, part = 'body') %>%
@@ -454,11 +454,6 @@ curexcmp_fun <- function(cursum, nativesum, restoresum){
     merge_at(i = 15:16, j = 7, part = 'body') %>%
     merge_at(i = 15:16, j = 8, part = 'body') %>%
     add_header_row(colwidths = c(2, 3, 3), values = c('', 'Native Habitats', 'Restorable Habitats')) %>%
-    add_footer_lines(values = "") %>%
-    footnote(i = 1, j = 1, sep = "", value = as_paragraph("N/A - Not Applicable; I/D - Insufficient Data"), part = 'body', inline = T, ref_symbols = "") %>%
-    add_footer_lines(values = "*All lands identified for acquisition by partners, does not represent a 2030 target or 2050 goal") %>%
-    add_footer_lines(values = "**Does not account for lands neither currently protected nor currently under consideration for acquisition") %>%
-    fontsize(size = 8, part = 'footer') %>%
     align(align = "center", part = "header") %>%
     align(i = c(2:6, 8:12, 14:17), j = 3:8, align = "center", part = "body") %>%
     # bg(i = c(1, 7, 13), bg = 'chartreuse3', part = "body") %>%
@@ -979,7 +974,7 @@ targetcmp_fun <- function(cursum, restoresum, trgsmetric){
     set_header_labels(
       Category = 'Stratum',
       HMPU_TARGETS = 'Habitat Type',
-      `total restorable` = 'Total Restoration Opportunity*',
+      `total restorable` = 'Total Restoration Opportunity',
[…]

### 2022-09-08 manu-draft.qmd
proofing intro, methods, updating workflow graphic

@@
 
-Native habitats in Florida face dual pressures at the land-sea interface from urban development and sea-level rise associated with climate change.
-To address these pressures, restoration practitioners require robust tools that identify reasonable goals given historical land use trends, current status of the habitat mosaic, and anticipated future impacts from coastal stressors.
+Native habitats in Florida face dual pressures at the land-sea interface from urban development and sea-level rise.
+To address these pressures, restoration practitioners require robust tools that identify reasonable goals given historical land use trends, current status of native habitats, and anticipated future impacts from coastal stressors.
 A target-setting approach for native habitats was created for the Tampa Bay watershed that identifies current restoration opportunities and establishes short-term (2030) targets and long-term (2050) goals.
@@
 Restoration goals were defined based on what is possible today and the projected needs for the future, rather than attempting to replicate past ecological conditions.
-The new paradigm also accounts for the expected impacts of sea-level rise, climate change, and watershed development - stressors which are pervasive in Florida.
+The new approach also accounts for the expected impacts of sea-level rise, climate change, and watershed development - stressors which are pervasive in Florida.
 The resulting habitat goals are spatially explicit with maps that identify remaining restoration opportunities, while also providing an approach for the entire watershed that targets subtidal, intertidal, and uplands.
-This approach represents a general framework to support coastal planning decisions that need to address competing interests and could be applied in other coastal settings where sustainable urbanization practices need to co-exist with natural environments.
+This approach also supports coastal planning decisions that need to address competing interests and could be applied in other coastal settings where sustainable urbanization practices need to co-exist with natural environments.
 Methods for repeatable analyses are also available using an open source workflow to update progress over time and for adoption by others.
@@
 
-The combined effects of land development and climate change are especially problematic for prioritizing habitat restoration or conservation activities in coastal environments.
+The combined effects of land development and climate change are especially problematic for prioritizing habitat restoration activities in coastal environments.
 Landward migration of critical habitats in response to sea-level rise may not be possible due to anthropogenic barriers in the watershed.
 Sea-level rise can occur quicker than landward migration of salt marshes and the upland slope may already be lost to urban development and hardening [@Titus09].
-The coastal squeeze of development and climate change diminishes the available space for future restoration in urban estuaries, thereby impacting the variety of ecosystem services provided by these habitats and the wildlife they support [@Enwright15].
-Given projected habitat losses without management intervention and the limited resources available, appropriate and realistic sites for restoration need to be identified that account for future stressors and past trends.
+The coastal squeeze of development and climate change diminishes the available space for future restoration in urban estuaries, thereby impacting ecosystem services provided by these habitats and the wildlife they support [@Enwright15].
+Given projected habitat losses and the limited resources available, appropriate and realistic sites for restoration need to be identified that account for future stressors and past trends.
 
@@
 Additionally, the habitat mosaic approach [@Henningsen05] of including multiple habitat types within restoration projects has been recognized as an effective means of allowing ecosystem state changes in response to different environmental pressures [@Duarte09; @Palmer09].
-Adaptive management [@Holling78; @Gregory06] components have been increasingly used to address challenges of sea-level rise, climate change, and development stressors, including monitoring to identify critical restoration decision points and needed intervention with contingency plans.
+Adaptive management [@Holling78; @Gregory06] components have also been used to address challenges of sea-level rise, climate change, and development stressors, including monitoring to identify critical restoration decision points and needed intervention with contingency plans.
 Elements of each of these approaches could be combined to create a more holistic approach to guide restoration and conservation activities for coastal habitats in urban settings.
 
-The Tampa Bay watershed (Florida, USA) is a valuable case study for developing a habitat restoration and conservation plan that addresses pervasive coastal stressors.
+The Tampa Bay watershed (Florida, USA) is a valuable case study for developing a habitat restoration plan that addresses pervasive coastal stressors.
 Compared to other estuaries, the ratio of watershed to estuary area is small and the area is heavily developed with 42% of land use classified as urban and suburban residential [@SWFWMD18].
-A retrospective approach to setting habitat protection and restoration targets in Tampa Bay was previously employed [@Lewis96; @Robison10; @Cicchetti11; @Russell15].
+A retrospective approach to setting habitat protection and restoration targets in Tampa Bay was previously used [@Lewis96; @Robison10; @Cicchetti11; @Russell15].
 Priority was given to restoration activities focused on habitat types that were important for a suite of estuarine faunal guilds disproportionately lost or degraded compared to a benchmark period.
-Criticisms of this approach included a lack of consideration for future sea-level rise and other climate change factors [@Yoskowitz15], use of expanded and different habitats outside the Tampa Bay watershed [@Robison10], lack of attention to upland or freshwater wetland habitats, and little recognition of land development trends or actual available space for restoration efforts.
+Criticisms of this approach included lack of consideration for future sea-level rise and other climate change factors [@Yoskowitz15], use of expanded and different habitats outside the Tampa Bay watershed [@Robison10], lack of attention to upland or freshwater wetland habitats, and little recognition of land development trends or actual available space for restoration efforts.
 These challenges are shared by restoration practitioners in other coastal environments and an approach that accommodates these challenges for planning would be highly transferable.
@@
 This paper describes an approach for habitat restoration and conservation planning that addresses the above challenges by considering the whole watershed, addressing historical changes, focusing on trajectories that have occurred during contemporary time periods, and considering both current and future stressors -- particularly land development and sea-level rise.
-Relatively consistent data are available for most Tampa Bay habitats of interest (1988 to 2018), representing a time period when federal, state and local regulations were in effect and regional impacts from climate change are documented [@Raabe12; @Cavanaugh14].
+Current and historical data are available for most Tampa Bay habitats, representing a time period when federal, state and local regulations were in effect and regional impacts from climate change have been documented [@Raabe12; @Cavanaugh14].
 The approach establishes a framework that can guide both watershed-level habitat planning and site-level restoration activities and incorporates applicable elements of other habitat restoration paradigms discussed above [@Palmer09].
@@
 The watershed covers approximately 5,872 km^2^, for a total combined area of approximately 6,855 km^2^.
-The climate is subtropical and within the current (2020) ecotone for mangrove and salt marsh habitats.
-The upstream watershed includes multiple habitats, including pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands.
-The watershed is heavily developed with an estimated population of 3.3 million people in the four major counties of the watershed [@Rayer20]*.* Numerous anthropogenic changes have altered the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom, mining activities), alteration of hydrology, and destruction and fragmentation of habitat from development.
+The climate is subtropical and within the 2020 ecotone for mangrove and salt marsh habitats.
+Naive habitats in the watershed include pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands.
+The watershed is heavily developed with an estimated population of 3.3 million people in the four major counties [@Rayer20]*.* Numerous anthropogenic changes have altered the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom, mining activities), alteration of hydrology, and destruction and fragmentation of habitat from development.
 
@@
 
-The major habitat types of Tampa Bay were stratified relative to tidal influence and location in the watershed.
-Subtidal habitats included those that are submerged all or most of the time; intertidal habitats included emergent tidal wetlands that are submerged during high tides but exposed during low tides; and supratidal habitats included those that occur above the high tide line (i.e., the remainder of the watershed).
+The major habitat types of Tampa Bay were stratified by tidal influence and location in the watershed to define broad categories for restoration planning.
+Subtidal habitats included those that are submerged all or most of the time, intertidal habitats included emergent tidal wetlands that are submerged during high tides but exposed during low tides, and supratidal habitats included those that occur above the high tide line (i.e., the remainder of the watershed).
 
 Subtidal habitats included hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13].
-Intertidal habiats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15].
+Intertidal habitats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15].
 Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below).
-This 5-foot contour area was considered the most likely to be impacted by sea-level rise based on current estimates.
+This 5-foot contour is an area of intense urban development and is expected to be impacted by sea-level rise based on current estimates.
 Since 1946, the St.
 Petersburg tidal gauge (NOAA gauge 872650) has documented a nearly 20 cm increase in mean tidal height to present day.
-Projections to the year 2100 suggest sea levels can increase from 58 to 259 cm in the region [@Burke19].
+Projections to the year 2100 suggest sea levels can increase between 58 and 259 cm in the region [@Burke19].
 
@@
 
-Opportunity areas for restoration and recommended coverage targets for habitat types were identified by integrating multiple datasets available for the region.
+Coverage targets for habitat types and opportunity areas for restoration were identified by integrating multiple datasets available for the region.
 First, habitat status and historical trends were quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time.
-Second, historic habitat restoration efforts conducted in the Tampa Bay watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners.
+Second, historic habitat restoration efforts conducted in the watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners.
 The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for coverage (hectares) of native habitats.
@@
 For the majority of subtidal, intertidal and supratidal habitats, coverages were quantified from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD).
-For subtidal habitats, 2018 source data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19].
+For subtidal habitats, 2018 data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19].
 Historical datasets for subtidal habitats began in 1988 with updates occurring on an approximate bi-annual basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.
 
-Current intertidal and supratidal habitat coverages were estimated using the 2017 SWFWMD Land Use Land Cover (LULC) geospatial database [@SWFWMD18].
-Land use and cover types (natural and developed) are classified pursuant to the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually.
+Current intertidal and supratidal habitat coverages were estimated using the 2017 SWFWMD Land Use Land Cover (LULC) maps [@SWFWMD18].
+Land use and cover types (natural and developed) are classified following the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually.
 While the photointerpretation of specific freshwater wetland types is often difficult, forested wetlands and non-forested wetlands can be distinguished with these data.
-Therefore, all applicable FLUCCS codes representing the suite of natural freshwater wetlands were combined within those two classifications.
+Therefore, all applicable FLUCCS codes representing natural freshwater wetlands were combined for these classifications.
 Native upland habitats were also combined in one classification.
-Historical estimates for these habitat coverages were also quantified starting with the earliest database in 1990 and occurring every two to three years until the current estimate in 2017.
+Historical estimates for all intertidal and supratidal habitats were also quantified starting with the earliest database in 1990 and occurring every two to three years until the current estimate in 2017.
 
 To address data gaps for habitats not included in the routine SWFWMD datasets, results from special studies were compiled to obtain current estimates.
-These included hard bottom subtidal habitats, artificial reefs, tidal creeks, and living shorelines, only available as current estimates with no information on historical trends.
+These included hard bottom subtidal habitats, artificial reefs, tidal creeks, and living shorelines.
+No information on historical trends is available for these habitats.
 
-Current and historical coverage estimates were obtained for all habitat types, including estimates of developed lands, for each year of available data spanning approximately three decades at two-year time steps.
-In addition, a habitat coverage change analysis between the terminal years of data (1988 to 2018 for subtidal, 1990 to 2017 for intertidal and supratidal) was conducted to understand how habitats were changing between types.
+Finally, a habitat coverage change analysis between the terminal years of data (1988 to 2018 for subtidal, 1990 to 2017 for intertidal and supratidal) was conducted to understand how habitats were changing between types.
 This required an intersection of the data layers to quantify if habitat types were unchanged or changed for any given location and identifying the type of change (e.g., seagrass to tidal flats).
@@
 
-Past restoration projects were quantified for each of the major habitat types, when available, to inform expectations for setting short-term goals and long-term targets.
-Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years was compiled from various sources.
+Restoration projects conducted over the past 40 years were quantified for each of the major habitat types to inform expectations for setting short-term targets and long-term goals.
 Restoration was defined as any activity that involved earthwork to reshape the land or the addition of structural elements (e.g., rock).
@@
 The collected data included project name, year, description, size (area or length), and location (latitude and longitude).
-Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and expert knowledge form local professionals.
-Living shoreline projects, including seawall enhancements and oyster reef installations, were inventoried separately.
+Data gaps were supplemented by archival research, site visits, contacting entities, and expert knowledge from local professionals.
 
@@
 Spatially-explicit estimates of the opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow.
-The two main processes included 1) binning existing datasets into relevant categories and 2) overlaying multiple datasets to identify restoration opportunities.
-Opportunity areas were defined as locations where habitat protection and restoration activities are possible and where they could occur to attain defined goals and targets.
-The definition and mapping of opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible under current and future projected conditions, related to anticipated land use change and sea-level rise.
+The two main processes included 1) binning existing datasets into relevant categories and 2) overlaying multiple datasets to identify opportunities.
+Opportunity areas were defined as locations where habitat protection and restoration activities are possible and where they could occur to attain the targets and goals described above.
+Identifying opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible given expected land use change and sea-level rise.
 
@@
 All FLUCCS classification codes were placed into one of three categories.
-First, native habitats were those that included the full range of natural plant communities and other habitats that are endemic to the watershed, and were further grouped into three major habitat types (tidal wetlands, freshwater wetlands, and uplands).
-Second, restorable habitats included existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through restoration.
+First, native habitats were those that included the full range of natural plant communities and other habitats that are endemic to the watershed Second, restorable habitats included existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through restoration.
 Third, existing development included developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration.
@@
 The reservation native and reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum.
-This area extends from the local Mean Lower Low Water (MLLW) elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation [@Burke19].
+This area extends from the local Mean Lower Low Water (MLLW) elevation to an elevation 5 feet above Mean Sea Level and is highly vulnerable to sea-level rise [@Burke19].
 Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands could be restored.
 
-The existing and proposed native and restorable areas were those that occured in public lands and parcels currently acquired or targeted for acquisition.
-To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory, consulting staff from various federal, state and local entities, and inventorying conservation and drainage easements data.
-This workflow created the existing conservation and proposed (i.e., planned to be acquired) conservation layers in @fig-gisworkflow .
+The existing and proposed native and restorable areas were those that occurred in public lands that are currently acquired or proposed for acquisition.
+To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory and inventorying conservation and drainage easements data.
+This workflow created the existing conservation and proposed conservation layers in @fig-gisworkflow .
 Intersecting the native and restorable lands in these areas produced four unique opportunity areas: existing conservation native, proposed conservation native, existing conservation restorable, and proposed conservation native.
@@
 Therefore, soil distributions can be used to estimate historical habitat distribution and restoration potential.
-@Ries14 created a soils suitability analysis for wetland mitigation and restoration and classified all soils in the Tampa Bay watershed as xeric, mesic, or hydric.
-The mesic and hydric categories were combined to represent wetland restoration potential, while the xeric category was used to represent upland restoration potential.
+A soils suitability layer was used for the Tampa Bay watershed [@Ries14] that classified soils as xeric, mesic, or hydric.
+The mesic and hydric categories were combined to represent wetland restoration potential and the xeric category was used to represent upland restoration potential.
 A distinction was made between tidal and freshwater wetland restoration potential by intersecting the mesic and hydric soils with the coastal stratum.
-Mesic or hydric soils that occur below the 5-foot contour were assigned a restoration potential for tidal wetlands, while mesic or hydric soils occurring above the 5-foot contour were assigned a restoration potential for freshwater wetlands.
+Mesic or hydric soils that occur below the 5-foot contour were assigned a restoration potential for tidal wetlands, whereas mesic or hydric soils above the 5-foot contour were assigned a restoration potential for freshwater wetlands.
 
@@
 The first was an opportunities map that showed areas in the watershed identified as existing conservation native, existing conservation restorable, proposed conservation native, proposed conservation restorable, reservation native, and reservation restorable.
-The second was a map that identified the restorable lands (either proposed, existing, or reservation) based on their restoration potential as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands (which included salt marshes).
+The second was a map that identified the restorable lands (either existing or reservation) based on their restoration potential as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands.
 All analyses were conducted using the R statistical programming language [@RCT22], specifically leveraging functions from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [@Pebesma18].
-The workflows and data are provided in an open-access repository available on GitHub (<https://github.com/tbep-tech/hmpu-workflow>) for use by others [@Beck22].
+The workflows and data are provided in an open-access repository available on GitHub (<https://github.com/tbep-tech/hmpu-workflow>) [@Beck22].
 

### 2022-09-09 R/tabs.R
color removed from trend tables

@@ -45,8 +45,8 @@ thm <- function(x){
 
 # table
 subttab <- flextable::flextable(totab, col_keys = grep('^chgcol$', names(totab), value = T, invert = T)) %>%
-  flextable::bg(i = ~ `chgcol` == 0, j = '% change', bg = losscol) %>%
-  flextable::bg(i = ~ `chgcol` == 1, j = '% change', bg = gaincol) %>%
+  # flextable::bg(i = ~ `chgcol` == 0, j = '% change', bg = losscol) %>%
+  # flextable::bg(i = ~ `chgcol` == 1, j = '% change', bg = gaincol) %>%
   thm
 
 # subttab
@@ -75,6 +75,7 @@ totab <- acres %>%
     Hectares = Hectares / 2.471
   ) %>%
   tidyr::pivot_wider(names_from = yr, values_from = Hectares) %>%
+  filter(`Habitat Type` != 'Open Water') %>%
   mutate(
     `1990 to 2017` = `2017` - `1990`,
     `% change` = 100 * (`2017` - `1990`) / `1990`,
@@ -101,8 +102,8 @@ thm <- function(x){
 
 # table
 acretab <- flextable::flextable(totab, col_keys = grep('^chgcol$', names(totab), value = T, invert = T)) %>%
-  flextable::bg(i = ~ `chgcol` == 0, j = '% change', bg = losscol) %>%
-  flextable::bg(i = ~ `chgcol` == 1, j = '% change', bg = gaincol) %>%
+  # flextable::bg(i = ~ `chgcol` == 0, j = '% change', bg = losscol) %>%
+  # flextable::bg(i = ~ `chgcol` == 1, j = '% change', bg = gaincol) %>%
   thm
 
 # acretab

### 2022-09-09 manu-draft.qmd
finished proofing results and discussion, added acknowledgements and key words

@@
       email: gmangrove@gmail.com
-      institute: [TBEP]
       correspondence: false
@@
       institute: [TBEP]
-      corrspondence: false
+      correspondence: false
   - Justin Saarinen:
@@
 Methods for repeatable analyses are also available using an open source workflow to update progress over time and for adoption by others.
+
+*Key words*: Florida, land use change, restoration planning, sea-level rise, Tampa Bay, urbanization
 
@@
 
-Current estimates and trend information were available for seagrasses, tidal flats, and oyster bars (Table 1).
+Current estimates and trend information on subtidal habitats were available for seagrasses, tidal flats, and oyster bars (Table 1).
 Oyster bars were estimated at 67 ha in 2018 (Table 1), showing a 29% increase since mapping began in 2014.
-This increase may represent improved ground-truthing and photointerpretation of oyster bars from aerial photography.
+This increase may represent improved ground-truthing and photointerpretation, rather than a true increase in coverage.
 Tidal flats have generally increased from 1988 to the mid-2000s, followed by a decrease to present.
@@
 The change analysis comparing 1988 to 2018 for subtidal habitats (@fig-subtchgdatalluv) confirmed trends in Table 1 and showed that the seagrass increases were primarily associated with the colonization of non-vegetated areas of tidal flats, as well as unclassified areas of open water.
-Current estimates for subtidal habitats without historical trend information included 171 ha for natural hard bottom and 67 hectares for artificial reefs.
+Current estimates for subtidal habitats without historical trend information included 171 ha for hard bottom habitat and 67 hectares for artificial reefs.
 
-Total intertidal habitat (mangroves, salt barrens, and salt marshes) showed a 12% increase of 835 ha from 1990 to 2017, with a current estimate of 8,340 ha (Table 2).
+Total intertidal habitat (mangroves, salt barrens, and salt marshes) increased by 12% to 8,340 ha from 1990 to 2017 (Table 2).
 Mangroves increased by 15% to 6,276 ha, salt barrens increased by 7% to 203 ha, and salt marshes increased by 3% to 1,861 ha.
-Despite a net increase in salt marsh habitat, the change analysis showed that 153 ha were replaced by mangrove habitat (@fig-chgdatalluv).
+Despite a net increase in salt marsh habitat, the change analysis showed that 153 ha were replaced by mangroves (@fig-chgdatalluv).
 The current extent of tidal tributary length is 622 km (no trend information is available).
@@
 Trend assessments for supratidal habitats showed the effects of increasing land development and loss of restorable habitats in the Tampa Bay watershed (Table 2).
-Developed lands showed a 44% increase to 217,047 ha from 1990 to 2017, with a notable 30% decrease to 1,446 ha for coastal uplands, a 38% decrease to 57,836 ha for native uplands, and an 18% decrease to 189,512 ha for restorable lands.
-Non-forested freshwater wetlands showed a 24% increase to 27,358 ha, while forested freshwater wetlands decreased by 5% to 61,667 ha.
+Developed lands increased by 44% increase to 217,047 ha from 1990 to 2017.
+Coastal uplands decreased by 30% to 1,446 ha, native uplands decreased by 38% to 57,836 ha, and restorable lands decreased by 18% to 189,512 ha.
+Non-forested freshwater wetlands increased by 24% to 27,358 ha, whereas forested freshwater wetlands decreased by 5% to 61,667 ha.
 The change analysis (@fig-chgdatalluv) showed that a majority of land conversion to developed came from restorable lands (21,292 ha) and native uplands (7,184 ha), with smaller proportions converted from forested freshwater wetlands (1,407 ha) and coastal uplands (193 ha).
@@
 Forty partners were responsible for these projects, although some were from departments within the same agency.
-Eighty-nine living shoreline projects, seawall enhancements, and oyster reef installations were documented, with a total length of 18.2 km.
+Eighty-nine living shoreline projects, seawall enhancements, and oyster reef installations were documented, totaling 18.2 km.
 Although projects were documented for the whole period of record, few projects were completed prior to 1990.
@@
 
-Summaries of the current extent of each habitat type are shown in Table 3 as representations of the opportunity areas and restoration potential.
+The current extent of each habitat type is shown in Table 3 as summaries for the opportunity areas and restoration potential.
 The extent of each habitat in existing conservation lands and proposed conservation lands is shown.
@@
 
-The map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap).
+The map of the remaining opportunity areas provided a spatial summary of where practitioners could target future restoration projects (@fig-oppmap).
 Native lands currently protected, proposed for protection, or in the coastal stratum (existing conservation native, proposed conservation native, reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW).
-Similarly, restorable lands currently protected, proposed for protection, or in the coastal stratum (existing conservation restorable, proposed conservation restorable, reservation restorable) totalled `r areas$restorable$areaha` (`r areas$restorable$areaper` of the watershed).
-Understandably, most of the native and restorable lands occurred in undeveloped areas in the north and southwest.
-Existing conservation lands (existing conservation native, existing conservation restorable) totalled `r areas$existing$areaha` (`r areas$existing$areaper` of the watershed) and proposed conservation lands (proposed conservation native, proposed conservation restorable) totalled `r areas$proposed$areaha` (`r areas$proposed$areaper` of the watershed).
-Reservation areas in the coastal stratum (reservation native, reservation restorable) totalled `r areas$reservation$areaha` (`r areas$reservation$areaper` of the watershed).
+Similarly, restorable lands currently protected, proposed for protection, or in the coastal stratum (existing conservation restorable, proposed conservation restorable, reservation restorable) totaled `r areas$restorable$areaha` (`r areas$restorable$areaper` of the watershed).
+Understandably, most of the native and restorable lands occurred in undeveloped areas in northern and southeastern areas of the watershed.
+Existing conservation lands (existing conservation native, existing conservation restorable) totaled `r areas$existing$areaha` (`r areas$existing$areaper` of the watershed) and proposed conservation lands (proposed conservation native, proposed conservation restorable) totaled `r areas$proposed$areaha` (`r areas$proposed$areaper` of the watershed).
+Reservation areas in the coastal stratum (reservation native, reservation restorable) totaled `r areas$reservation$areaha` (`r areas$reservation$areaper` of the watershed).
 
@@
 These priorities are necessary given anticipated impacts of land development and climate change, while also considering competing societal interests for use of the environment and limited resources for land acquisition and restoration.
-Our approach balances these tradeoffs by providing a systematic workflow to identify targets and goals that is informed by current extent, past trends, and realistic effort from past projects.
-Further, locations are identified where these targets and goals could be achieved based on existing opportunities for restorable habitats, including areas anticipated to be impacted by coastal stressors (i.e., sea level rise and land development).
+Our approach balances these tradeoffs by identifying targets and goals that are informed by current extent, past trends, and realistic effort from past projects.
+Further, locations are identified where these targets and goals could be achieved based on existing opportunities for restorable habitats, including areas anticipated to be impacted by coastal stressors (i.e., sea-level rise and land development).
 This approach departs from previous restoration paradigms by identifying what is possible rather than attempting to recreate historical baselines.
-Methods are also provided using open source tools [@Beck22] that 1) allow for the most current datasets to be synthesized to update progress on achieving targets and goals, and 2) can be used in other locations with similar needs for identifying restoration priorities.
+Methods are also provided using open source tools [@Beck22] that 1) allow for the most current datasets to be synthesized to assess progress, and 2) can be used in other locations with similar needs for identifying restoration priorities.
 
@@
 
-Identifying appropriate targets and goals would not have been possible without a detailed assessment of current extent and past trends of native habitats in Tampa Bay and its watershed.
+Identifying appropriate targets and goals would not have been possible without a detailed assessment of current extent and past trends over thirty years of native habitats in Tampa Bay and its watershed.
 The most notable trends included 1) the increase of seagrasses by 75%, 2) the increase of emergent tidal wetlands (12%) and freshwater wetlands (24%), and 3) the loss of native uplands (38%).
 
-Seagrass recovery in Tampa Bay is a well-known success story that demonstrated how public-private partnerships can effectively reduce total nitrogen loads into Tampa Bay [@Greening14; @Sherwood17] .
+Seagrass recovery in Tampa Bay is a well-known success story that demonstrated how public-private partnerships can effectively reduce total nitrogen loads into Tampa Bay [@Greening14; @Sherwood17].
 The nutrient reductions, primarily from point-source controls and advanced wastewater treatment, contributed to improvements in water quality and light environments that were favorable for seagrass growth.
@@
 As such, the identified targets and goals for mangroves indicate protection of these habitats, without the need for additional restoration.
-However, mangroves will continue to expand in the intertidal zone contributing to additional losses of salt marshes and salt barrens.
+However, mangroves are expected to continue colonization of the intertidal zone, contributing to additional losses of salt marshes and salt barrens.
 The reservation areas identified in @fig-oppmap represent critical remaining areas in the intertidal zone that could be protected to prevent additional losses of tidal wetlands.
@@
 Amendments to existing planning, zoning, and land development policies will be needed to address these issues.
-While federal and state regulations related to listed species management provide some protection to rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats are largely unprotected (e.g., pine flatwoods).
+Although federal and state regulations for endangered species provide some protection to rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats are largely unprotected (e.g., pine flatwoods).
 Reductions in native uplands will likely continue unless local governments improve regulatory protections, such as strengthening language within comprehensive plans and development ordinances.
@@
 Focusing efforts on publicly-owned conservation lands is expected to have long-term benefits and will be most cost-effective given the level of restoration effort compared to habitats that have already been impacted by anthropogenic activities.
-As such, public acquisition of remaining critical lands (e.g., coastal uplands) is a high priority given current development trends in the Tampa Bay watershed.
-Other restoration targets (e.g., salt marshes) will not be feasible without additional public acquisition or public-private partnerships as a mechanism for doing so [@Holl01; @Benson18].
+As such, public acquisition of remaining critical lands (e.g., coastal uplands) is a high priority given current development trends in the watershed.
+Other restoration targets (e.g., salt marshes) will not be obtained without additional public acquisition or public-private partnerships as a mechanism for doing so [@Holl01; @Benson18].
 Therefore, varied approaches to leverage resources for restoration are needed and could include pursuing traditional grants, matching funds from multiple partners, and volunteer initiatives for habitat restoration.
@@
 
-Other restoration activities could be pursued for the opportunity areas identified herein.
+Other restoration activities could be pursued for the opportunity areas.
 Substantial opportunities exist for upland restoration on reclaimed mined lands within the watershed (@fig-lulcmap, @fig-oppmap).
 For estuarine habitats, opportunity areas could include dredged holes or spoil disposal areas, either for enhancing existing subtidal habitats or creating areas that could be colonized by seagrasses.
-Some opportunities also exist on developed lands primarily through enhancement projects, although these have not been explicitly identified in the products created herein.
-Examples include the construction of living shorelines in place of hardened seawalls, placement of submerged habitat modules (e.g., artificial oyster reef) along urban shorelines, and creation of backyard habitats.
+Some opportunities also exist on developed lands primarily through enhancement projects, although these have not been explicitly identified in the products herein.
+Examples include the construction of living shorelines in place of hardened seawalls, placement of submerged habitat modules along urban shorelines (e.g., artificial oyster reefs), and creation of backyard habitats.
 Tidal tributary restoration could also include removal of salinity barriers and filling of dredged channel sections.
@@
 Coastal setbacks, buffers, or public easements are traditionally used to restrict development within a given distance from the shoreline.
-However, rolling easements may be an alternative approach whereby protected areas are allowed to "roll" landward with expected changes in sea level rise.
+However, rolling easements may be an alternative approach whereby protected areas are allowed to "roll" landward with expected changes in sea-level rise.
 Rolling easements could disincentivize more intense urban development of low-lying coastal uplands in less developed agricultural or recreational land uses.
@@
 The workflow for identifying priorities required detailed and spatially-explicit datasets specific to the Tampa Bay watershed.
-Long-term datasets describing land use and cover in the watershed and the extent of in-bay habitats were necessary to categorize current extent and past trends.
+Long-term datasets describing land use and cover and the extent of subtidal habitats were necessary to categorize current extent and past trends.
 Similarly, supporting datasets included those that described existing and proposed conservation areas, soils, past restoration activities, and relevant spatial boundaries (i.e., watershed and coastline).
@@
 For example, inventories of hard bottom habitats, living shorelines, tidal tributaries, and artificial reefs were created through special studies or were available only as current estimates from regional entities.
-Tracking progress towards targets in goals for these habitats will require regular updates to these datasets that are not part of routine land use and cover maps.
+Tracking progress towards targets and goals for these habitats will require regular updates to these datasets that are not part of routine land use and cover maps.
 
@@
 Specifically, restoration effort by regional partners is cataloged in the available restoration database, which not only depends on voluntary reporting, but also represents a source of information on restoration extent that is separate from land cover maps.
-An expectation is that acreage restored for a given habitat as reported by a partner will ultimately be shown as a change in land use and cover on regional maps.
+An expectation is that the reported coverage restored by a partner will ultimately be shown as a change in land use and cover on regional maps.
 The temporal lag between an actual project and how it may be reflected in a GIS product is unknown, which may create a disconnect between the updates in achieving targets and goals as new data layers are released and the effort reported by partners.
@@
 
+## Acknowledgments
+
+The Technical Advisory Committee of the Tampa Bay Estuary Program reviewed drafts of the original reports that are the basis of this paper and their comments are immensely appreciated.
+We also recognize the extensive efforts of regional partners, of which there are too many to mention, for past restoration efforts in the Tampa Bay region that have contributed to native habitats.
+These efforts will be critical for continued success of restoration in the future.
+Substantial efforts by partners in creating and curating routine datasets for the region, especially by the Southwest Florida Water Management District and the Florida Natural Areas Inventory, are also gratefully recognized.
+
+\newpage
+
 ## Figures
 
-![Land use and land cover for the Tampa Bay watershed, Florida, USA.
+![Land use, cover, and subtidal habitats for the Tampa Bay watershed, Florida, USA.
 The watershed includes the natural hydrologic boundary with minor modifications to include partners working with the Tampa Bay Estuary Program.](figs/lulcmap.png){#fig-lulcmap fig-align="center"}
@@
 ![Spatial analysis workflow used to identify opportunity areas and restoration potential in the Tampa Bay watershed.
-Workflows are divided into binning of land use/land cover categories into relevant habitat types and spatial overlay of datasets to identify the opportunity areas.](figs/gisworkflow.png){#fig-gisworkflow fig-align="center"}
+Workflows are divided into binning of land use/land cover categories into relevant habitat types and spatial overlay of datasets to identify the opportunity areas and restoration potential.](figs/gisworkflow.png){#fig-gisworkflow fig-align="center"}
 
 ![Change analysis of habitat categories in the subtidal strata of the Tampa Bay watershed.
-The left column shows relative areas in 1988 and the right column shows relative areas in 2018 for each habitat category.
-The grey lines show the proportional change of each habitat category between the years.](figs/subtchgdatalluv.png){#fig-subtchgdatalluv fig-align="center"}
+The left column shows relative areas in 1988 and the right column shows relative areas in 2018 for each habitat category, where the bar heights are proportional to extents in each year.
+The grey lines show the proportional change in area of each habitat category between the years.](figs/subtchgdatalluv.png){#fig-subtchgdatalluv fig-align="center"}
 
 ![Change analysis of habitat categories in the intertidal and supratidal strata of the Tampa Bay watershed.
-The left column shows relative areas in 1990 and the right column shows relative areas in 2017 for each habitat category.
-The grey lines show the proportional change of each habitat category between the years.](figs/chgdatalluv.png){#fig-chgdatalluv fig-align="center"}
+The left column shows relative areas in 1990 and the right column shows relative areas in 2017 for each habitat category, where the bar heights are proportional to extents in each year.
+The grey lines show the proportional change in area of each habitat category between the years.](figs/chgdatalluv.png){#fig-chgdatalluv fig-align="center"}
 
-![Opportunity areas for habitat conservation and reservation in the Tampa Bay watershed.
+![Opportunity areas for habitat restoration in the Tampa Bay watershed.
 Green indicates existing conservation, blue indicates proposed conservation, and pink indicates reservations opportunities.
@@
 Areas are identified as those where habitat restoration could target the four identified categories as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands.
-Categories are based on the opportunity areas and soil types.
+Categories are based on the opportunity areas, soil types, and coastal boundaries.
 The outline is the Tampa Bay watershed.](figs/restmap.png){#fig-restmap fig-align="center"}

### 2022-10-02 manu-draft.qmd
adding kfw and es comments

@@
 
+load(file = here('tabs/opptab.RData'))
 load(file = here('tabs/subttab.RData'))
@@
 The new approach also accounts for the expected impacts of sea-level rise, climate change, and watershed development - stressors which are pervasive in Florida.
-The resulting habitat goals are spatially explicit with maps that identify remaining restoration opportunities, while also providing an approach for the entire watershed that targets subtidal, intertidal, and uplands.
+The resulting habitat goals are provided as maps that identify remaining restoration opportunities, while also providing an approach for the entire watershed that targets subtidal, intertidal, and uplands.
 This approach also supports coastal planning decisions that need to address competing interests and could be applied in other coastal settings where sustainable urbanization practices need to co-exist with natural environments.
@@
 Sea-level rise can occur quicker than landward migration of salt marshes and the upland slope may already be lost to urban development and hardening [@Titus09].
-The coastal squeeze of development and climate change diminishes the available space for future restoration in urban estuaries, thereby impacting ecosystem services provided by these habitats and the wildlife they support [@Enwright15].
+Coastal development and climate change diminish the available space for future restoration in urban estuaries, thereby impacting ecosystem services provided by these habitats and the wildlife they support [@Enwright15].
 Given projected habitat losses and the limited resources available, appropriate and realistic sites for restoration need to be identified that account for future stressors and past trends.
@@
 
-This paper describes an approach for habitat restoration and conservation planning that addresses the above challenges by considering the whole watershed, addressing historical changes, focusing on trajectories that have occurred during contemporary time periods, and considering both current and future stressors -- particularly land development and sea-level rise.
+In this paper, we describe an approach for habitat restoration and conservation planning that addresses the above challenges by considering the whole watershed, addressing historical changes, focusing on trajectories that have occurred during contemporary time periods, and considering both current and future stressors -- particularly land development and sea-level rise.
 Current and historical data are available for most Tampa Bay habitats, representing a time period when federal, state and local regulations were in effect and regional impacts from climate change have been documented [@Raabe12; @Cavanaugh14].
@@
 
-Subtidal habitats included hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13].
+Subtidal habitats included hard bottom [@Jaap90; @Ash05; @Kaufman17; @CSA19], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13].
 Intertidal habitats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15].
@@
 First, habitat status and historical trends were quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time.
-Second, historic habitat restoration efforts conducted in the watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners.
-The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for coverage (hectares) of native habitats.
-Finally, remaining opportunities for restoration were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be impacted by sea-level rise.
+Second, historic habitat restoration efforts conducted in the watershed were synthesized to inform on a practical and feasible level of effort that could be conducted by restoration practitioners in the future.
+The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for native habitat coverage (hectares).
+Finally, remaining restoration opportunities were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be impacted by sea-level rise.
 As such, the approach identifies goals and targets and provides spatially explicit information that identifies where restoration practitioners should pursue projects to meet these goals and targets.
@@
 For subtidal habitats, 2018 data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19].
-Historical datasets for subtidal habitats began in 1988 with updates occurring on an approximate bi-annual basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.
+Historical datasets for subtidal habitats began in 1988 with updates occurring on an approximate biennial basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.
 
@@
 To address data gaps for habitats not included in the routine SWFWMD datasets, results from special studies were compiled to obtain current estimates.
-These included hard bottom subtidal habitats, artificial reefs, tidal creeks, and living shorelines.
+These included hard bottom subtidal habitats, artificial reefs, tidal creeks, and living shorelines [@Robison20].
 No information on historical trends is available for these habitats.
@@
 Opportunity areas were defined as locations where habitat protection and restoration activities are possible and where they could occur to attain the targets and goals described above.
-Identifying opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible given expected land use change and sea-level rise.
+Identifying opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible given underlying soil conditions, expected land use change, and sea-level rise.
 
@@
 
-After binning, the native and restorable lands were overlaid with additional layers to identify 1) reservation native and reservation restorable areas, and 2) existing and proposed native and restorable areas.
-Collectively, each of these unique products are considered the opportunity areas in the Tampa Bay watershed.
-The reservation native and reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum.
+After binning, the native and restorable lands were overlaid with additional layers to identify 1) coastal reservation native and coastal reservation restorable areas, and 2) existing and proposed native and restorable areas.
+Collectively, each of these unique products are considered the opportunity areas in the Tampa Bay watershed (Table 1).
+The coastal reservation native and coastal reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum.
 This area extends from the local Mean Lower Low Water (MLLW) elevation to an elevation 5 feet above Mean Sea Level and is highly vulnerable to sea-level rise [@Burke19].
-Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands could be restored.
+Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands or coastal uplands could be restored.
 
 The existing and proposed native and restorable areas were those that occurred in public lands that are currently acquired or proposed for acquisition.
-To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory and inventorying conservation and drainage easements data.
-This workflow created the existing conservation and proposed conservation layers in @fig-gisworkflow .
-Intersecting the native and restorable lands in these areas produced four unique opportunity areas: existing conservation native, proposed conservation native, existing conservation restorable, and proposed conservation native.
+To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory and permit databases of conservation and drainage easements.
+This workflow created the existing conservation and proposed conservation layers in @fig-gisworkflow.
+Intersecting the native and restorable lands in these areas produced four unique opportunity areas: existing conservation native, proposed conservation native, existing conservation restorable, and proposed conservation restorable.
 
-All opportunity areas identified as restorable included reservation restorable (in the coastal stratum), existing conservation restorable, and proposed conservation restorable.
-To identify discrete habitat types that could be the goal of future restoration projects, restorable lands in the coastal stratum and on existing conservation areas (reservation restorable and existing conservation restorable) were further grouped into their restoration potential.
+All opportunity areas identified as restorable included coastal reservation restorable, existing conservation restorable, and proposed conservation restorable.
+To identify discrete habitat types that could be the goal of future restoration projects, restorable lands in the coastal stratum and on existing conservation areas (coastal reservation native and existing conservation restorable) were further grouped into their restoration potential by underlying soil types.
 Proposed conservation areas were excluded from the analysis to provide a more confident assessment of restoration potential in areas that have already been acquired (i.e., existing conservation) or are immediately threatened by sea-level rise and/or coastal development (i.e., the coastal stratum).
-The restoration potential was defined using soil types, described as follows.
 
@@
 Two distinct mapping products were created from the above analysis.
-The first was an opportunities map that showed areas in the watershed identified as existing conservation native, existing conservation restorable, proposed conservation native, proposed conservation restorable, reservation native, and reservation restorable.
-The second was a map that identified the restorable lands (either existing or reservation) based on their restoration potential as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands.
+The first was an opportunities map that showed areas in the watershed identified as existing conservation native, existing conservation restorable, proposed conservation native, proposed conservation restorable, coastal reservation native, and coastal reservation restorable.
+The second was a map that identified the restorable lands (either existing or coastal reservation) based on their restoration potential as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands.
 All analyses were conducted using the R statistical programming language [@RCT22], specifically leveraging functions from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [@Pebesma18].
@@
 
-Current estimates and trend information on subtidal habitats were available for seagrasses, tidal flats, and oyster bars (Table 1).
-Oyster bars were estimated at 67 ha in 2018 (Table 1), showing a 29% increase since mapping began in 2014.
+Current estimates and trend information on subtidal habitats were available for seagrasses, tidal flats, and oyster bars (Table 2).
+Oyster bars were estimated at 67 ha in 2018 (Table 2), showing a 29% increase since mapping began in 2014.
 This increase may represent improved ground-truthing and photointerpretation, rather than a true increase in coverage.
@@
 The current estimate for tidal flats is 6,569 hectares, showing a 24% decline compared to the 1988 estimate of 8,700 hectares.
-Seagrasses have increased by 75% since 1988 to a current estimate of 6,986 ha.
-The change analysis comparing 1988 to 2018 for subtidal habitats (@fig-subtchgdatalluv) confirmed trends in Table 1 and showed that the seagrass increases were primarily associated with the colonization of non-vegetated areas of tidal flats, as well as unclassified areas of open water.
+Seagrasses have increased by 75% (6,986 ha) since 1988 to a current estimate of 16,293 ha.
+The change analysis comparing 1988 to 2018 for subtidal habitats (@fig-subtchgdatalluv) confirmed trends in Table 2 and showed that the seagrass increases were primarily associated with the colonization of non-vegetated areas of tidal flats, as well as unclassified areas of open water.
 Current estimates for subtidal habitats without historical trend information included 171 ha for hard bottom habitat and 67 hectares for artificial reefs.
 
-Total intertidal habitat (mangroves, salt barrens, and salt marshes) increased by 12% to 8,340 ha from 1990 to 2017 (Table 2).
+Total intertidal habitat (mangroves, salt barrens, and salt marshes) increased by 12% to 8,340 ha from 1990 to 2017 (Table 3).
 Mangroves increased by 15% to 6,276 ha, salt barrens increased by 7% to 203 ha, and salt marshes increased by 3% to 1,861 ha.
@@
 
-Trend assessments for supratidal habitats showed the effects of increasing land development and loss of restorable habitats in the Tampa Bay watershed (Table 2).
-Developed lands increased by 44% increase to 217,047 ha from 1990 to 2017.
+Trend assessments for supratidal habitats showed the effects of increasing land development and loss of restorable habitats in the Tampa Bay watershed (Table 3).
+Developed lands increased by 44% to 217,047 ha from 1990 to 2017.
 Coastal uplands decreased by 30% to 1,446 ha, native uplands decreased by 38% to 57,836 ha, and restorable lands decreased by 18% to 189,512 ha.
 Non-forested freshwater wetlands increased by 24% to 27,358 ha, whereas forested freshwater wetlands decreased by 5% to 61,667 ha.
-The change analysis (@fig-chgdatalluv) showed that a majority of land conversion to developed came from restorable lands (21,292 ha) and native uplands (7,184 ha), with smaller proportions converted from forested freshwater wetlands (1,407 ha) and coastal uplands (193 ha).
-Habitats converted to restorable primarily included native uplands (8,304 ha), forested freshwater wetlands (1,700 ha), and developed (2,794 ha).
+The change analysis (@fig-chgdatalluv) showed that a majority of conversion to developed lands came from restorable areas (21,292 ha) and native uplands (7,184 ha), with smaller proportions converted from forested freshwater wetlands (1,407 ha) and coastal uplands (193 ha).
+Habitats converted to restorable areas primarily included native uplands (8,304 ha), forested freshwater wetlands (1,700 ha), and developed lands (2,794 ha).
 The increase in non-forested freshwater wetlands was primarily from restorable lands (2,759 ha).
@@
 
-The current extent of each habitat type is shown in Table 3 as summaries for the opportunity areas and restoration potential.
+The current extent of each habitat type is shown in Table 4 as summaries for the opportunity areas and restoration potential.
 The extent of each habitat in existing conservation lands and proposed conservation lands is shown.
 Summaries of the restoration potential under existing and proposed conservation lands is also shown.
-The majority of restoration opportunities on existing conservation lands are for native uplands and freshwater wetlands.
+Most restoration opportunities on existing conservation lands are for native uplands and freshwater wetlands.
 Less opportunities exist for intertidal wetlands (mangrove forests, salt barrens, and salt marshes).
@@
 The map of the remaining opportunity areas provided a spatial summary of where practitioners could target future restoration projects (@fig-oppmap).
-Native lands currently protected, proposed for protection, or in the coastal stratum (existing conservation native, proposed conservation native, reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW).
-Similarly, restorable lands currently protected, proposed for protection, or in the coastal stratum (existing conservation restorable, proposed conservation restorable, reservation restorable) totaled `r areas$restorable$areaha` (`r areas$restorable$areaper` of the watershed).
-Understandably, most of the native and restorable lands occurred in undeveloped areas in northern and southeastern areas of the watershed.
+Native habitats currently protected (existing conservation native), proposed for protection (proposed conservation native), or in the coastal stratum (coastal reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW).
+Similarly, restorable lands currently protected (existing conservation restorable), proposed for protection (proposed conservation restorable), or in the coastal stratum (coastal reservation native) totaled `r areas$restorable$areaha` (`r areas$restorable$areaper` of the watershed).
+Understandably, most of the native and restorable lands occurred in undeveloped areas in northern and southeastern areas of the watershed (@fig-restmap).
 Existing conservation lands (existing conservation native, existing conservation restorable) totaled `r areas$existing$areaha` (`r areas$existing$areaper` of the watershed) and proposed conservation lands (proposed conservation native, proposed conservation restorable) totaled `r areas$proposed$areaha` (`r areas$proposed$areaper` of the watershed).
-Reservation areas in the coastal stratum (reservation native, reservation restorable) totaled `r areas$reservation$areaha` (`r areas$reservation$areaper` of the watershed).
+Reservation areas in the coastal stratum (coastal reservation native, coastal reservation restorable) totaled `r areas$reservation$areaha` (`r areas$reservation$areaper` of the watershed).
 
 Combining the restorable lands on existing conservation areas and in the coastal stratum with soils data provided a spatial summary of the restoration potential grouped by habitat type (@fig-restmap).
-A total of `r areas$potentialall$areaha` (`r areas$potentialall$areaper` of the watershed) were identified as potentially restorable as coastal uplands (`r areas$potential[['Coastal Uplands']]$areaha`, `r areas$potential[['Coastal Uplands']]$areaper` of the watershed), freshwater wetlands (`r areas$potential[['Freshwater Wetlands']]$areaha`, `r areas$potential[['Freshwater Wetlands']]$areaper` of the watershed), native uplands (`r areas$potential[['Native Uplands']]$areaha`, `r areas$potential[['Native Uplands']]$areaper` of the watershed), or tidal wetlands (`r areas$potential[['Tidal Wetlands']]$areaha`, `r areas$potential[['Tidal Wetlands']]$areaper` of the watershed).
+A total of `r areas$potentialall$areaha` (`r areas$potentialall$areaper` of the watershed) of potentially restorable lands were identified, further partitioned as coastal uplands (`r areas$potential[['Coastal Uplands']]$areaha`, `r areas$potential[['Coastal Uplands']]$areaper` of the watershed), freshwater wetlands (`r areas$potential[['Freshwater Wetlands']]$areaha`, `r areas$potential[['Freshwater Wetlands']]$areaper` of the watershed), native uplands (`r areas$potential[['Native Uplands']]$areaha`, `r areas$potential[['Native Uplands']]$areaper` of the watershed), or tidal wetlands (`r areas$potential[['Tidal Wetlands']]$areaha`, `r areas$potential[['Tidal Wetlands']]$areaper` of the watershed).
 
@@
 Identifying short-term (2030) targets and long-term (2050) goals for the restoration extent of native habitats in Tampa Bay was informed by the assessment of current extents, past trends, and relative effort for past restoration and enhancement projects.
-Table 4 shows the targets and goals identified through this analysis and the associated rationale.
+Table 5 shows the targets and goals identified through this analysis and the associated rationale.
 For example, the targets and goals are established based on the current extent and the restoration potential.
 If restoration potential exists and coverage restored from past projects suggests a reasonable level of effort, the targets and goals reflect the current extent plus the restoration opportunity.
-Conversely, other habitats with no identified restoration opportunity, or current extents are sufficient (e.g., mangrove forests), were assigned targets and goals similar to the current extent, i.e., these habitats should be protected and restoration is not recommended.
+Conversely, other habitats with no identified restoration opportunity, or with sufficient current extents (e.g., mangrove forests), were assigned targets and goals similar to the current extent, i.e., these habitats should be protected and further restoration will only increase resilience.
 Implicit in the targets and the goals is recurring re-assessment over time to evaluate progress and adjust expectations as appropriate.
@@
 
-Priorities for habitat restoration should be informed by current assessments and what is possible to achieve in the future.
+Priorities for comprehensive, watershed-wide habitat restoration should be informed by current assessments and what is possible to achieve in the future.
 These priorities are necessary given anticipated impacts of land development and climate change, while also considering competing societal interests for use of the environment and limited resources for land acquisition and restoration.
 Our approach balances these tradeoffs by identifying targets and goals that are informed by current extent, past trends, and realistic effort from past projects.
-Further, locations are identified where these targets and goals could be achieved based on existing opportunities for restorable habitats, including areas anticipated to be impacted by coastal stressors (i.e., sea-level rise and land development).
-This approach departs from previous restoration paradigms by identifying what is possible rather than attempting to recreate historical baselines.
+Further, spatially-explicit locations are identified where these targets and goals could be achieved based on existing opportunities for restorable habitats, including areas anticipated to be impacted by coastal stressors (i.e., sea-level rise and land development).
+This approach departs from previous restoration paradigms by identifying what is possible rather than attempting to recreate an ideal historical baseline.
 Methods are also provided using open source tools [@Beck22] that 1) allow for the most current datasets to be synthesized to assess progress, and 2) can be used in other locations with similar needs for identifying restoration priorities.
@@
 Identifying appropriate targets and goals would not have been possible without a detailed assessment of current extent and past trends over thirty years of native habitats in Tampa Bay and its watershed.
-The most notable trends included 1) the increase of seagrasses by 75%, 2) the increase of emergent tidal wetlands (12%) and freshwater wetlands (24%), and 3) the loss of native uplands (38%).
+The most notable trends included 1) an increase of seagrasses by 75%, 2) an increase of emergent tidal wetlands (12%) and freshwater wetlands (24%), and 3) a loss of native uplands (38%).
 
@@
 The nutrient reductions, primarily from point-source controls and advanced wastewater treatment, contributed to improvements in water quality and light environments that were favorable for seagrass growth.
-However, the most recent coverage estimate in 2020 showed a seagrass loss of 18% since the peak in 2016, falling below the target defined herein.
-These data were unavailable at the time the habitat restoration workflow was initially developed and trends informed by the new restoration paradigm have prompted bay managers to assess barriers in achieving the seagrass restoration goal.
+However, the most recent (2020) coverage estimate showed a seagrass loss of 18% since peak coverages in 2016, falling below the target defined herein.
+These data were unavailable at the time this habitat restoration workflow was initially developed and trends informed by the new restoration paradigm have prompted bay managers to assess barriers in achieving the seagrass restoration goal.
 In particular, much of the seagrass losses have occurred in Old Tampa Bay, where recurring algal blooms of *Pyrodinium bahamense* have contributed to water quality decline [@Lopez19] .
@@
 Dual pressures from sea-level rise and climate change have impacted tidal wetlands, such that mangrove forests are outcompeting salt marshes and salt barrens for available niche space.
-Mangrove expansion as a result of climate change has been observed throughout the Gulf of Mexico [@Comeaux12].
+Mangrove expansion as a result of climate change has been observed throughout the Gulf of Mexico [@Comeaux12; @Osland22].
 As such, the identified targets and goals for mangroves indicate protection of these habitats, without the need for additional restoration.
@@
 Attaining the target and goal will require restoration of upland habitats on existing conservation lands (i.e., restoration potential in @fig-restmap) and new conservation lands to offset the continued loss of these habitats to development.
-The long-term conversion of restorable lands to developed (@fig-chgdatalluv) presents additional challenges for restoration of native uplands.
+The long-term conversion of restorable areas to developed lands (@fig-chgdatalluv) presents additional challenges for restoration of native uplands.
 Amendments to existing planning, zoning, and land development policies will be needed to address these issues.
 Although federal and state regulations for endangered species provide some protection to rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats are largely unprotected (e.g., pine flatwoods).
-Reductions in native uplands will likely continue unless local governments improve regulatory protections, such as strengthening language within comprehensive plans and development ordinances.
+Reductions in native uplands will likely continue unless local governments improve regulatory protections, such as strengthening language within comprehensive plans and development ordinances to maintain a defined extent of these habitats within a rapidly urbanizing coastal watershed.
 
@@
 As such, public acquisition of remaining critical lands (e.g., coastal uplands) is a high priority given current development trends in the watershed.
-Other restoration targets (e.g., salt marshes) will not be obtained without additional public acquisition or public-private partnerships as a mechanism for doing so [@Holl01; @Benson18].
-Therefore, varied approaches to leverage resources for restoration are needed and could include pursuing traditional grants, matching funds from multiple partners, and volunteer initiatives for habitat restoration.
+Other restoration targets (e.g., salt marshes) will not be obtained without additional public acquisition or initiating novel public-private partnerships as a mechanism for doing so [@Holl01; @Benson18].
+Therefore, varied approaches to leverage resources for restoration are needed and could include pursuing traditional grants, matching funds from multiple partners, or voluntary initiatives that incentivize habitat restoration [e.g., Blue Carbon investments, @Sheehan19].
 Recent gains in tidal and freshwater wetlands are primarily due to publicly-funded habitat restoration projects, state and federal wetland regulatory programs, and to a lesser extent, regulatory mitigation.
@@
 Overall, restoration practitioners must consider several options and choose those that are most feasible given the available resources and likelihood of success.
+Further, finer-scale land cover classification datasets are currently being investigated to refine identification of opportunity areas within the urbanized, developed landscape of the watershed.
 
@@
 
-Finally, wetland impacts and associated compensatory mitigation projects authorized under wetland regulatory programs could serve as possible restoration mechanisms.
+Finally, wetland impacts and associated compensatory mitigation projects authorized under wetland regulatory programs could serve as more focused and directed restoration mechanisms to help achieve watershed-wide goals.
 Mitigation activities have historically been conducted independent of watershed-level planning and monitoring processes.
-This disconnect has contributed to fragmented implementation and inconsistent compliance monitoring of mitigation projects, including historically poor documentation of wetland losses and gains in the Tampa Bay watershed.
+This disconnect has contributed to fragmented implementation and habitat function and inconsistent compliance monitoring of mitigation projects, including historically poor documentation of wetland losses and gains in the Tampa Bay watershed.
 However, if properly focused and coordinated, compensatory mitigation activities could significantly contribute to the attainment of restoration targets and goals for the region.
@@
 Many of these datasets are available outside of Tampa Bay, although temporal and spatial resolutions may limit application to other areas.
+Despite the region’s data richness, limitations still exist in classifying restoration opportunities based on the spatial-scale of the LULC datasets.
+Additional refinements within the classified developed lands and coastal reservation space are currently being explored with 1-m scale national land cover datasets.
 Additionally, considerable effort was made in working with regional partners to identify and fill knowledge gaps for relevant habitat types.
 For example, inventories of hard bottom habitats, living shorelines, tidal tributaries, and artificial reefs were created through special studies or were available only as current estimates from regional entities.
-Tracking progress towards targets and goals for these habitats will require regular updates to these datasets that are not part of routine land use and cover maps.
+Tracking progress towards these habitat targets and goals is heavily reliant on regular updates to these datasets, as well as routine land use and cover map updates.
 
-An additional assumption of the workflow, particularly for tracking progress, is that projects reported by partners will ultimately manifest in map products.
+An additional assumption of the workflow, particularly for tracking progress, is that implemented restoration projects reported by partners will ultimately manifest into a classification within the map products.
 Specifically, restoration effort by regional partners is cataloged in the available restoration database, which not only depends on voluntary reporting, but also represents a source of information on restoration extent that is separate from land cover maps.
 An expectation is that the reported coverage restored by a partner will ultimately be shown as a change in land use and cover on regional maps.
-The temporal lag between an actual project and how it may be reflected in a GIS product is unknown, which may create a disconnect between the updates in achieving targets and goals as new data layers are released and the effort reported by partners.
+The temporal lag between an actual project and how it may be reflected in a GIS product is unknown, which may create a disconnect between the updates in achieving targets and goals as new data layers are released and the effort reported by partners becomes represented within the data layers.
 The spatial resolution of mapping products may also be insufficient to detect habitat changes as reported in the restoration database.
 For these reasons, projects reported by partners are currently summarized separately from the assessments above that depend on GIS layers.
-Additional work is needed to reconcile these datasets for more streamline reporting.
+Additional work is needed to reconcile these datasets for more streamlined reporting.
 
@@
 The products created herein will guide these efforts for the next thirty years by providing a continuously updated assessment of where the opportunities exist and if targets and goals are expected to be met.
-The Tampa Bay area is not unique in its challenges for protecting native habitats and the approach is readily transferable to other locations where restoration priorities are needed in response to pervasive coastal stressors.
+The Tampa Bay region is not unique in the challenges resource managers face to protect and restore native habitats, and the approach described herein is readily transferable to other locations where restoration priorities are needed in response to pervasive coastal stressors.
 
@@
 ![Opportunity areas for habitat restoration in the Tampa Bay watershed.
-Green indicates existing conservation, blue indicates proposed conservation, and pink indicates reservations opportunities.
+Green indicates existing conservation, blue indicates proposed conservation, and pink indicates reservation opportunities.
 Each category is also grouped into native and restorable habitats.
@@
 ```{r}
-#| tbl-cap: Change over time in hectares for subtidal habitats in Tampa Bay.
+#| tbl-cap: "Table 1: Description of the opportunity areas in the Tampa Bay watershed identified through spatial analysis.
+Native habitats include those in the watershed not considered developed or restorable (e.g., freshwater wetlands, forested uplands, etc.).
+Restorable areas are altered but non-hardened and pervious lands that could potentially support native habitats through restoration.
+Existing and proposed conservation areas are those that are publicly owned or on conservation easements that currently exist or are proposed for acquisition, respectively, as identified primarily in the Florida Natural Areas Inventory.
+Coastal reservation areas occur within the coastal stratum identified from the bay shoreline to the 5-foot elevation contour."
+opptab
+```
+
+\newpage
+
+```{r}
+#| tbl-cap: "Table 2: Change over time in hectares for subtidal habitats in Tampa Bay.
 Columns show years with available data and the final two columns show the change and percent change from 1988 to 2018.
-Oyster bars were not meaningfully quantified prior to 2014.
+Oyster bars were not meaningfully quantified prior to 2014."
 subttab
@@
 ```{r}
-#| tbl-cap: Change over time in hectares for intertidal and supratidal habitats in Tampa Bay.
-Columns show years with available data and the final two columns show the change and percent change from 1990 to 2017.
+#| tbl-cap: "Table 3: Change over time in hectares for intertidal and supratidal habitats in Tampa Bay.
+Columns show years with available data and the final two columns show the change and percent change from 1990 to 2017."
 acretab
@@
 ```{r}
-#| tbl-cap: "Summary of habitat restoration opportunities in the Tampa Bay watershed.
+#| tbl-cap: "Table 4: Summary of habitat restoration opportunities in the Tampa Bay watershed.
 Summaries are based on 2017 land use data, 2018 subtidal data, best estimates for habitat types not in existing GIS layers, and current extent of existing and proposed conservation lands.
@@
 ```{r}
-#| tbl-cap: "Recommended 2030 targets and 2050 goals for habitat restoration and protection in the Tampa Bay watershed.
+#| tbl-cap: "Table 5: Recommended 2030 targets and 2050 goals for habitat restoration and protection in the Tampa Bay watershed.
 Targets and goals are based on 2017 land use data, 2018 subtidal data, best estimates for habitat types not in existing GIS layers, and current extent of existing and proposed conservation lands.

### 2022-10-02 R/dat_proc.R

@@ -250,24 +250,57 @@ save(restdat, file = here('data/restdat.RData'))
 
 # summary of areas in opportunity and restoration potential -----------------------------------
 
+fluccs <- read.csv(here('data', 'FLUCCShabsclass.csv'), stringsAsFactors = F)
+
+lulcfl <- 'lulc2017'
+subtfl <- 'sgdat2018'
+
+# from 01_inputs
+load(here('data', paste0(lulcfl, '.RData')))
+load(here('data', paste0(subtfl, '.RData')))
+lulc <- get(lulcfl)
+subt <- get(subtfl)
+load(file = here('data/hard.RData'))
+load(file = here('data/arti.RData'))
+load(file = here('data/tidt.RData'))
+load(file = here('data/livs.RData'))
+load(file = here('data/coastal.RData'))
+load(file = here('data/strata.RData'))
 load(file = here('data/oppdat.RData'))
 load(file = here('data/restdat.RData'))
 
-# summarizes area and formats for manu by type in oppareas
-areasumfun <- function(x, flt){
+# data copied from commit e6026f831e90cffca8ea906c8fe53c85ceb5746a in hmpu-workflow
+# these are the original HMPU layers in the report
+load(file = here('data/restorelyr.RData'))
+load(file = here('data/nativelyr.RData'))
 
-  out <- x %>%
-    filter(grepl(flt, cat)) %>%
-    group_by(var) %>%
-    summarise(
-      val = sum(val),
-      .groups = 'drop'
+# get info from table 3, then tabulate for native, restorable, existing, and proposed
+notab <- curex_fun(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strata, nativelyr, restorelyr, notab = T)
+
+notabfrm <- notab %>%
+  mutate_if(
+    is.character, function(x) gsub('\\D', '', x) %>%  as.numeric
+  ) %>%
+  filter(Category != 'Subtidal')
+
+# filter by type, calc total, and format
+sumfun <- function(notabfrm, flt){
+
+  out <- notabfrm %>%
+    select(matches(flt)) %>%
+    apply(2, function(x) ifelse(duplicated(x), NA, x)) %>%
+    sum(na.rm = TRUE) %>%
+    tibble(areaha = .) %>%
+    mutate(
+      areaper = 100 * (areaha / 587200)
     ) %>%
[…]

### 2022-10-02 R/figs.R

@@ -167,6 +167,15 @@ saveNetwork(a2, here("figs/subtchgdatalluv.html"))
 data("tbshed")
 load(file = here('data/oppdat.RData'))
 
+oppdat <- oppdat %>%
+  mutate(
+    cat = case_when(
+      cat == 'Reservation Native' ~ 'Coastal Reservation Native',
+      cat == 'Reservation Restorable' ~ 'Coastal Reservation Restorable',
+      T ~ cat
+    )
+  )
+
 p <- oppmap_fun(oppdat, tbshed, northloc = 'tr', scaleloc = 'tl',
                 buffdist = 0.04)
 

### 2022-10-02 R/funcs.R

@@ -172,7 +172,8 @@ fixgeo <- function(dat){
 # nativelyr is current existing/proposed native sf object
 # restorelyr is current existing/proposed restoration layer
 # crplyr is optional cropping layer
-curex_fun <- function(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strata, nativelyr, restorelyr, crplyr = NULL){
+# notab will return the summarized data not as a flextable object
+curex_fun <- function(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strata, nativelyr, restorelyr, crplyr = NULL, notab = F){
 
   # crop all sf objects by optional crop layer
   if(!is.null(crplyr)){
@@ -334,7 +335,7 @@ curex_fun <- function(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strat
 
   # final table
 
-  tab <- curexcmp_fun(cursum, nativesum, restoresum)
+  tab <- curexcmp_fun(cursum, nativesum, restoresum, notab = notab)
 
   return(tab)
 
@@ -385,7 +386,7 @@ curexleg_fun <- function(){
 }
 
 # final table compilation function for curex_fun, curexleg_fun
-curexcmp_fun <- function(cursum, nativesum, restoresum){
+curexcmp_fun <- function(cursum, nativesum, restoresum, notab = F){
 
   # combine all for table
 
@@ -430,6 +431,9 @@ curexcmp_fun <- function(cursum, nativesum, restoresum){
       `restorable Existing`,
       `restorable Proposed`
     )
+
+  if(notab)
+    return(allsum)
 
   # make table
 
@@ -565,8 +569,8 @@ oppmap_fun <- function(oppmap, bndry, northloc = 'tr', scaleloc = 'tl', buffdist
     `Existing Conservation Restorable` = 'green4',
     `Proposed Conservation Native` = 'dodgerblue1',
     `Proposed Conservation Restorable` = 'dodgerblue4',
-    `Reservation Native` = 'violetred1',
-    `Reservation Restorable` = 'violetred3'
+    `Coastal Reservation Native` = 'violetred1',
+    `Coastal Reservation Restorable` = 'violetred3'
     ) %>%
     unlist
 

### 2022-10-02 R/tabs.R

@@ -6,6 +6,32 @@ library(units)
 
 source(here('R/funcs.R'))
 
+# opportunity area description ----------------------------------------------------------------
+
+totab <- tibble(
+  `Opportunity area` = c('Existing Conservation Native', 'Existing Conservation Restorable', 'Proposed Conservation Native', 'Proposed Conservation Restorable', 'Coastal Reservation Native', 'Coastal Reservation Restorable'),
+  `Description`= c(
+    'Native habitats currently within existing conservation lands',
+    'Restorable areas currently within existing conservation lanbs',
+    'Native habits within proposed conservation lands',
+    'Restorable areas within proposed conservations lands',
+    'Native habitats within the coastal stratum',
+    'Restorable areas within the coastal stratum')
+)
+
+thm <- function(x){
+  flextable::width(x, width = 1.75, j = 1) %>%
+    flextable::width(width = 4.75, j = 2) %>%
+    # flextable::width(width = 2.75, j = 7) %>%
+    fontsize(size = 8, part = 'all') %>%
+    flextable::padding(padding = 0, part = 'all')
+}
+
+opptab <- flextable(totab) %>%
+  thm
+
+save(opptab, file = here('tabs/opptab.RData'))
+
 # subtidal habitat change ---------------------------------------------------------------------
 
 data(subtacres)

### 2022-10-11 manu-draft.qmd
added mcb comments

@@
 Habitat changes in response to climate change include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].
-Management of coastal habitats requires an understanding of how stressors impact these resources, while balancing competing societal uses.
+Management of coastal habitats requires an understanding of how stressors influence these resources, while balancing competing societal uses.
 
@@
 The climate is subtropical and within the 2020 ecotone for mangrove and salt marsh habitats.
-Naive habitats in the watershed include pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands.
+Native habitats in the watershed include pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands.
 The watershed is heavily developed with an estimated population of 3.3 million people in the four major counties [@Rayer20]*.* Numerous anthropogenic changes have altered the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom, mining activities), alteration of hydrology, and destruction and fragmentation of habitat from development.
@@
 Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below).
-This 5-foot contour is an area of intense urban development and is expected to be impacted by sea-level rise based on current estimates.
+This 5-foot contour is an area of intense urban development and is expected to be affected by sea-level rise based on current estimates.
 Since 1946, the St.
@@
 The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for native habitat coverage (hectares).
-Finally, remaining restoration opportunities were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be impacted by sea-level rise.
-As such, the approach identifies goals and targets and provides spatially explicit information that identifies where restoration practitioners should pursue projects to meet these goals and targets.
+Finally, remaining restoration opportunities were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be affected by sea-level rise.
+As such, the approach identifies reasonable goals and targets based on past trends and provides spatially explicit information that identifies where restoration practitioners could prioritize projects based on opportunities within their respective jurisdictions.
 
@@
 Emergent tidal and freshwater wetlands in Tampa Bay have also experienced dramatic changes over the last three decades.
-Dual pressures from sea-level rise and climate change have impacted tidal wetlands, such that mangrove forests are outcompeting salt marshes and salt barrens for available niche space.
+Dual pressures from sea-level rise and changes in the length of freeze-free season have affected tidal wetlands, such that mangrove forests are outcompeting salt marshes and salt barrens for available niche space.
 Mangrove expansion as a result of climate change has been observed throughout the Gulf of Mexico [@Comeaux12; @Osland22].
+Anthropogenic water withdrawals have also reduced freshwater flows reaching tidal marshes, contributing to reductions in coverage of key species (e.g., *Juncus roemerianus*) in place of mangroves [@Raabe12].
 As such, the identified targets and goals for mangroves indicate protection of these habitats, without the need for additional restoration.
@@
 The reservation areas identified in @fig-oppmap represent critical remaining areas in the intertidal zone that could be protected to prevent additional losses of tidal wetlands.
-Likewise, gains in non-forested freshwater wetlands are a reflection of 1) the effectiveness of state and federal wetland regulatory programs, and 2) the cumulative gains from publicly-funded habitat restoration projects.
-A similar approach could be used for salt marshes or salt barrens.
+Likewise, gains in non-forested freshwater wetlands are a reflection of 1) constructed stormwater ponds required by state and federal regulatory programs, and 2) the cumulative gains from publicly-funded habitat restoration projects.
+Creative restoration approaches (e.g., habitat acquisition and optimal management of freshwater flows) that address the likely expansion of mangroves at the expense of salt marshes and salt barrens will be required to meet the targets and goals for these habitats.
 
@@
 The long-term conversion of restorable areas to developed lands (@fig-chgdatalluv) presents additional challenges for restoration of native uplands.
-Amendments to existing planning, zoning, and land development policies will be needed to address these issues.
+Additional education about best practices in land development, market-based incentives, and amendments to existing planning, zoning, and land development regulations will be needed to address these issues.
 Although federal and state regulations for endangered species provide some protection to rare habitats, such as scrub jay (*Aphelocoma coerulescens*) habitat, common and historically abundant native habitats are largely unprotected (e.g., pine flatwoods).
-Reductions in native uplands will likely continue unless local governments improve regulatory protections, such as strengthening language within comprehensive plans and development ordinances to maintain a defined extent of these habitats within a rapidly urbanizing coastal watershed.
+Voluntary approaches to low impact urban or suburban development may also gain traction among developers as more viable methods for land conversion that minimize impacts to natural resources while increasing quality of life [@Jones09].
+Education and outreach activities that target land developers to raise awareness of the benefits of alternative practices are critical in this effort.
+Market-based approaches to mitigating urban sprawl may also be practical [e.g., conservation subdvisions, @Mohamed06] given the estimated economic gains relative to conventional approaches.
+Regardless, reductions in native uplands will likely continue in the short-term unless local governments improve regulatory protections, such as strengthening language within comprehensive plans and development regulations to maintain a defined extent of these habitats within a rapidly urbanizing coastal watershed.
 
@@
 
-Creative approaches may be required in areas impacted by sea-level rise if land acquisition is not possible.
-These approaches are necessary to accommodate future landward migration of tidal wetlands or the protection of coastal uplands, while also preventing or minimizing development activities that can inhibit landward habitat shifts.
+Creative approaches may be required in areas affected by sea-level rise if land acquisition is not possible.
+These approaches are necessary to accommodate future landward migration of tidal wetlands or the protection of coastal uplands, while also reducing risks to built infrastructure that, when inappropriately sited, can inhibit landward habitat shifts.
 Coastal setbacks, buffers, or public easements are traditionally used to restrict development within a given distance from the shoreline.
@@
 Rolling easements could disincentivize more intense urban development of low-lying coastal uplands in less developed agricultural or recreational land uses.
-Landowners could maintain current economic uses with a rolling easement, while reserving such lands to accommodate landward habitat migration.
+Landowners could maintain current economic uses with a rolling easement, while reserving such lands to accommodate future landward habitat migration.
+These approaches also offer risk-reduction to built infrastructure by offering increased protection from potential affects of sea-level rise and other coastal stressors (e.g., storm surge).
 
-Finally, wetland impacts and associated compensatory mitigation projects authorized under wetland regulatory programs could serve as more focused and directed restoration mechanisms to help achieve watershed-wide goals.
+Finally, wetland impacts and associated compensatory mitigation projects authorized under wetland regulatory programs could serve as more directed restoration mechanisms to help achieve watershed-wide goals.
 Mitigation activities have historically been conducted independent of watershed-level planning and monitoring processes.
-This disconnect has contributed to fragmented implementation and habitat function and inconsistent compliance monitoring of mitigation projects, including historically poor documentation of wetland losses and gains in the Tampa Bay watershed.
+This disconnect has contributed to fragmented implementation, marginal habitat function, and inconsistent compliance monitoring of mitigation projects, including historically poor documentation of wetland losses and gains in the Tampa Bay watershed.
 However, if properly focused and coordinated, compensatory mitigation activities could significantly contribute to the attainment of restoration targets and goals for the region.
@@
 Despite the region’s data richness, limitations still exist in classifying restoration opportunities based on the spatial-scale of the LULC datasets.
-Additional refinements within the classified developed lands and coastal reservation space are currently being explored with 1-m scale national land cover datasets.
+Additional refinements within the classified developed lands and coastal reservation space are currently being explored with 1-m scale national land cover datasets, which could expose additional opportunities in urban or suburban areas (e.g., improving stormwater infrastructure).
 Additionally, considerable effort was made in working with regional partners to identify and fill knowledge gaps for relevant habitat types.
@@
 
-![Spatial analysis workflow used to identify opportunity areas and restoration potential in the Tampa Bay watershed.
-Workflows are divided into binning of land use/land cover categories into relevant habitat types and spatial overlay of datasets to identify the opportunity areas and restoration potential.](figs/gisworkflow.png){#fig-gisworkflow fig-align="center"}
+![Spatial analysis workflow used to identify opportunity areas (existing conservation native, proposed conservation native, existing conservation restorable, proposed conservation restorable, reservation native, reservation restorable) and restoration potential (coastal uplands, uplands, freshwater wetlands, tidal wetlands) in the Tampa Bay watershed.
+Workflows are divided into binning of land use/land cover categories into relevant habitat types and spatial overlay of datasets to identify the opportunity areas and restoration potential.
+The approach was applied to both the intertidal and supratidal strata of the watershed.](figs/gisworkflow.png){#fig-gisworkflow fig-align="center"}
 

### 2022-10-11 manu-draft.qmd
added DT comments

@@
 Petersburg tidal gauge (NOAA gauge 872650) has documented a nearly 20 cm increase in mean tidal height to present day.
-Projections to the year 2100 suggest sea levels can increase between 58 and 259 cm in the region [@Burke19].
+Projections from the year 2000 to 2100 suggest sea levels can increase between 58 and 259 cm in the region [@Burke19].
 
@@
 The nutrient reductions, primarily from point-source controls and advanced wastewater treatment, contributed to improvements in water quality and light environments that were favorable for seagrass growth.
-However, the most recent (2020) coverage estimate showed a seagrass loss of 18% since peak coverages in 2016, falling below the target defined herein.
+Reducing nitrogen inputs into Tampa Bay remains the primary strategy for maintaining water quality conditions.
+However, the most recent (2020) coverage estimate showed a seagrass loss of 18% baywide since peak coverages in 2016, falling below the target defined herein.
 These data were unavailable at the time this habitat restoration workflow was initially developed and trends informed by the new restoration paradigm have prompted bay managers to assess barriers in achieving the seagrass restoration goal.
-In particular, much of the seagrass losses have occurred in Old Tampa Bay, where recurring algal blooms of *Pyrodinium bahamense* have contributed to water quality decline [@Lopez19] .
-Ongoing research to understand mechanisms for mitigating these blooms will be critical for restoring seagrass in Tampa Bay.
+In particular, much of the seagrass losses have occurred in Old Tampa Bay (northwest segment of Tampa Bay), where recurring algal blooms of *Pyrodinium bahamense* have contributed to water quality decline [@Lopez19].
+The greatest percent loss of seagrass in 2020 was observed in Hillsborough Bay (northeast segment of Tampa Bay), which does not experience *P.
+bahamense* blooms.
+Ongoing research to understand mechanisms for mitigating blooms that negatively affect water quality, in addition to identifying potential regional stressors, will be critical for restoring seagrass in Tampa Bay.
 
 Emergent tidal and freshwater wetlands in Tampa Bay have also experienced dramatic changes over the last three decades.
-Dual pressures from sea-level rise and changes in the length of freeze-free season have affected tidal wetlands, such that mangrove forests are outcompeting salt marshes and salt barrens for available niche space.
+Dual pressures from sea-level rise and changes in the length of the freeze-free season have affected tidal wetlands, such that mangrove forests are outcompeting salt marshes and salt barrens for available niche space.
 Mangrove expansion as a result of climate change has been observed throughout the Gulf of Mexico [@Comeaux12; @Osland22].
@@
 Despite the region’s data richness, limitations still exist in classifying restoration opportunities based on the spatial-scale of the LULC datasets.
-Additional refinements within the classified developed lands and coastal reservation space are currently being explored with 1-m scale national land cover datasets, which could expose additional opportunities in urban or suburban areas (e.g., improving stormwater infrastructure).
+Additional refinements within the classified developed lands and coastal reservation space are currently being explored with 1-m scale national land cover datasets, which could expose additional opportunities in urban or suburban areas (e.g., improving stormwater infrastructure) or differential changes within native habitat classes (e.g., interspecific differences in mangrove colonization).
 Additionally, considerable effort was made in working with regional partners to identify and fill knowledge gaps for relevant habitat types.

### 2022-10-14 manu-draft.qmd
FEE formatting and submission

@@
     reference-doc: my_styles.docx
+    number-sections: true
 title: "Addressing climate change and development pressures in an urban estuary through habitat restoration planning"
@@
 
-## Abstract
+## Abstract {.unnumbered}
 
@@
 Pervasive stressors can negatively impact the services provided by these habitats.
-Anthropogenic land development can remove or substantially modify existing habitats, and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14].
+Anthropogenic land development can remove or substantially modify existing habitats and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14].
 Habitat changes in response to climate change include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].
@@
 The general approach includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in coverage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories.
-The outcomes are spatially-specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.
+The outcomes are spatially specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.
 
@@
 
-Spatially-explicit estimates of the opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow.
+Spatially explicit estimates of the opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow.
 The two main processes included 1) binning existing datasets into relevant categories and 2) overlaying multiple datasets to identify opportunities.
@@
 Our approach balances these tradeoffs by identifying targets and goals that are informed by current extent, past trends, and realistic effort from past projects.
-Further, spatially-explicit locations are identified where these targets and goals could be achieved based on existing opportunities for restorable habitats, including areas anticipated to be impacted by coastal stressors (i.e., sea-level rise and land development).
+Further, spatially explicit locations are identified where these targets and goals could be achieved based on existing opportunities for restorable habitats, including areas anticipated to be impacted by coastal stressors (i.e., sea-level rise and land development).
 This approach departs from previous restoration paradigms by identifying what is possible rather than attempting to recreate an ideal historical baseline.
@@
 Identifying restoration priorities was data intensive and would not have been possible without the resources available for the region.
-The workflow for identifying priorities required detailed and spatially-explicit datasets specific to the Tampa Bay watershed.
+The workflow for identifying priorities required detailed and spatially explicit datasets specific to the Tampa Bay watershed.
 Long-term datasets describing land use and cover and the extent of subtidal habitats were necessary to categorize current extent and past trends.
@@
 
-## Acknowledgments
+## Acknowledgments {.unnumbered}
 
@@
 
-## Figures
+## Figures {.unnumbered}
 
 ![Land use, cover, and subtidal habitats for the Tampa Bay watershed, Florida, USA.
-The watershed includes the natural hydrologic boundary with minor modifications to include partners working with the Tampa Bay Estuary Program.](figs/lulcmap.png){#fig-lulcmap fig-align="center"}
+The watershed includes the natural hydrologic boundary with minor modifications to include partners working with the Tampa Bay Estuary Program.](figs/lulcmap.jpg){#fig-lulcmap fig-align="center"}
 
@@
 Workflows are divided into binning of land use/land cover categories into relevant habitat types and spatial overlay of datasets to identify the opportunity areas and restoration potential.
-The approach was applied to both the intertidal and supratidal strata of the watershed.](figs/gisworkflow.png){#fig-gisworkflow fig-align="center"}
+The approach was applied to both the intertidal and supratidal strata of the watershed.](figs/gisworkflow.jpg){#fig-gisworkflow fig-align="center"}
 
@@
 The left column shows relative areas in 1988 and the right column shows relative areas in 2018 for each habitat category, where the bar heights are proportional to extents in each year.
-The grey lines show the proportional change in area of each habitat category between the years.](figs/subtchgdatalluv.png){#fig-subtchgdatalluv fig-align="center"}
+The grey lines show the proportional change in area of each habitat category between the years.](figs/subtchgdatalluv.jpg){#fig-subtchgdatalluv fig-align="center"}
 
@@
 The left column shows relative areas in 1990 and the right column shows relative areas in 2017 for each habitat category, where the bar heights are proportional to extents in each year.
-The grey lines show the proportional change in area of each habitat category between the years.](figs/chgdatalluv.png){#fig-chgdatalluv fig-align="center"}
+The grey lines show the proportional change in area of each habitat category between the years.](figs/chgdatalluv.jpg){#fig-chgdatalluv fig-align="center"}
 
@@
 Each category is also grouped into native and restorable habitats.
-The outline is the Tampa Bay watershed.](figs/oppmap.png){#fig-oppmap fig-align="center"}
+The outline is the Tampa Bay watershed.](figs/oppmap.jpg){#fig-oppmap fig-align="center"}
 
@@
 Categories are based on the opportunity areas, soil types, and coastal boundaries.
-The outline is the Tampa Bay watershed.](figs/restmap.png){#fig-restmap fig-align="center"}
+The outline is the Tampa Bay watershed.](figs/restmap.jpg){#fig-restmap fig-align="center"}
 
@@
 
-## Tables
+## Tables {.unnumbered}
 
@@
 
-## References
+## References {.unnumbered}

### 2022-10-14 R/figs.R

@@ -139,7 +139,7 @@ plulc <- ggmap(bsmap1_transparent) +
 p <- plulc +
   inset(ggplotGrob(pinset), xmin = -81.85, xmax = -81.35, ymin = 27.4, ymax = 27.7)
 
-png(here('figs/lulcmap.png'), height = 5, width = 7, res = 300, units = 'in')
+jpeg(here('figs/lulcmap.jpg'), height = 5, width = 7, res = 300, units = 'in')
 p
 dev.off()
 
@@ -160,6 +160,8 @@ saveNetwork(a1, here("figs/chgdatalluv.html"))
 saveNetwork(a2, here("figs/subtchgdatalluv.html"))
 
 # go to chrome and use svgcrowbar to svg, then inkscape svg to png
+# can export svg to jpg in inkscape, but must set alpha of background to 100
+# this is under Document properties -> Background -> Background color
 # webshot screws up the labels
 
 # opportunity map -----------------------------------------------------------------------------
@@ -179,7 +181,7 @@ oppdat <- oppdat %>%
 p <- oppmap_fun(oppdat, tbshed, northloc = 'tr', scaleloc = 'tl',
                 buffdist = 0.04)
 
-png(here('figs/oppmap.png'), height = 5, width = 8, res = 300, units = 'in')
+jpeg(here('figs/oppmap.jpg'), height = 5, width = 8, res = 300, units = 'in')
 p
 dev.off()
 
@@ -190,6 +192,6 @@ load(file = here('data/restdat.RData'))
 p <- restmap_fun(restdat, tbshed, northloc = 'tr', scaleloc = 'tl',
                  buffdist = 0.04)
 
-png(here('figs/restmap.png'), height = 5, width = 8, res = 300, units = 'in')
+jpeg(here('figs/restmap.jpg'), height = 5, width = 8, res = 300, units = 'in')
 p
 dev.off()

### 2022-12-29 revision/review_response.qmd
starting FEE revision

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: false
bibliography: ../refs.bib
csl: ../frontiers-in-ecology-and-evolution.csl
---

We sincerely thank the reviewers for providing comments on our manuscript.  We provide a point-by-point response to these comments below.

All page and line numbers in the reviewer comments refer to the original submission.  Page numbers noted in our response refer to the revised submission.

## Reviewer 2

The authors suggest an interesting planning tool for coastal restoration. However, in the case given by the author, the basis for judging whether wetlands can be restored is very subjective, and there is no actual data to support whether such judgment is real and feasible, which is the biggest problem of the paper.

* __Response__:

Line 76. How to determine the benchmark period, or what is the benchmark period in the study area, decades ago?

Line 127-128. 5-foot contour is the absolute elevation or elevation relative to sea level?

Line 140. What are the specific contents of short-term and long-term goals? The reader may want a comparison table of the two goals

Line 206-215. The actual available space for restoration efforts depends on the relative relation between wetland elevation and tide height. Whether the slope or elevation of coastal wetlands is used to estimate the potential restoration area under the impact of sea level rise. The author should give specific elevation data and tidal data.

Line 216-220 and Figue2. In addition to the land use/land cover data, I think it is important to consider the digital elevation data as well as the relative sea level rise data.

Line 236. What is the classification standard of xeric, mesic, or hydric soil? Due to sea level rise, wetland soil salinity changes, resulting in the transformation of the spatial pattern of freshwater wetland to saltwater wetland, how to consider this.

Line 248-249. The author needs to specify which data analysis methods were used?

Line 304-316. The determination of potential repair area lacks quantitative basis. The priority of potential restoration areas at different locations depends on substrate conditions, ecological importance, engineering implementation, and restoration costs. For the whole region, the restoration potential of various ecosystems should be ranked, and only when they exceed a certain threshold can they be judged as having the operability and real potential of restoration under realistic conditions. Current opportunity maps simply overlap land use type data and soil data. I think such judgment is too subjective and arbitrary, lack of quantitative basis. Since the author mentions that there are data on restoration projects, why not use the projects already carried out to determine whether these potential restoration areas can actually be repaired in the future?

Figure 1. The classification criteria for land use types are confused. I suggest using the Ramsar Convention classification system.

## Reviewer 3

This paper presents a detailed method for spatial restoration planning using available landcover datasets for the Tampa Bay watershed. The authors provide a concise overview of the approach with the intention that the method can be applied in other regions. The research is important and relevant given the challenges related to both increasing development while considering conservation targets under potential climate change are widespread.

Overall, the paper is well-written and the authors provide a clear justification of the method, including the strengths (open source workflow), and limitations (detailed inputs, potential lag between reported restoration actions and observed landcover change in GIS products). Detailed comments on each section are provided below. Line numbers refer to those on the PDF version available for review.

* __Response__:

Abstract:

While the paper is focused on the method and spatial classification to identify areas for potential restoration, this is not represented in the Abstract. Instead, the Abstract indicates a focus on the target-setting approach and emphasizes how sea-level rise, climate change, and watershed development are incorporated into this approach. The Abstract should rather focus on the main results of the paper – which was the identification of the different areas and changes over time as measured by applying the method. That is not to say that the entire Abstract is unsuitable, rather that it needs to be amended to better reflect the contents of the paper, rather than a summary of the approach.

Introduction:

The Introduction would benefit from a stronger rationale / context in the opening paragraph. This first paragraph intends to highlight the ecological importance of estuarine / coastal ecosystems by listing ecosystem services and then mentioning impacts from anthropogenic activities and climate change. The rationale is focused on the need for management to incorporate stressors using a socio-ecological systems approach (L41-43). While this rationale is valid, the sentences leading up to it are somewhat disjointed. The readability could be improved by considering the information in the first and second paragraph together and then restructuring to provide better context. For example, L39-43 “Habitat changes in response to climate change include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration” is a very limited description, with no prior reference to the process of landward migration. A more descriptive narrative to the opening paragraphs of the Introduction will set the tone for the paper and further support the relevance and importance of the work.

Methods:

L107: Should this read as “altered the natural habitats of”

L117-126: If possible, it would help to have this information on the habitat layers and respective source material presented in a table – even as supplementary information.

L127: At first mention of the 5-foot contour please provide an estimate in metres for comparison. This will also make for a more straightforward understanding of the SLR projection, which is provided in cm (L132).

L130-132: The SLR trend and projections to 2100 are stated here, but it is not clear whether these have been incorporated into the 2030 / 2050 targets / goals for restoration. In the Introduction (L83-86), it is stated that the approach will consider future stressors, including SLR. A spatial representation of how habitat / land cover may change by 2030 / 2050 under SLR is however not part of the Results. Were the SLR projections incorporated spatially? And did this impact the potential restoration areas? For example, are transitions from mangrove to tidal flat / seagrass included in the targets / goals when considering SLR? Please provide some additional information on whether this was incorporated and how it was done. This is needed to justify the statement in L83-86.

L154-165: Please indicate the type and scale of the spatial datasets for the habitat. Are these raster datasets? What is the grid resolution? Are they obtained from Landsat, or other remotely-sensed products? Providing detail on the input datasets will greatly assist any effort to replicate this method in another region.

L170-175: Please provide more detail on how the habitat coverage change analysis was performed. Was this a simple calculation of total area change per habitat type derived from pixel count / polygon area? Or, were the maps for different years overlayed and a spatial algorithm applied to calculate differences? Please specify how the change from one habitat to another was estimated in order to generate the alluvial diagrams.

L177-188: While I understand the justification to differentiate between “restoration” and “enhancement” activities, I would suggest the authors re-consider the terminology and definitions they have provided. The term “restoration” is well-known, with the definition provided by the Society of Ecological Restoration becoming mainstreamed, particularly under the “Decade for Ecosystem Restoration”. The SER definition for restoration from the international principles and standards (https://www.ser.org/page/SERStandards): “the process of assisting the recovery of an ecosystem that has been degraded, damaged or destroyed. (Ecosystem restoration is sometimes used interchangeably with ecological restoration, but ecological restoration always addresses biodiversity conservation and ecological integrity, whereas some approaches to ecosystem restoration may focus solely on the delivery of ecosystem services.)” Therefore, it is advised that the authors do not provide their own definition for restoration to only refer to activities where earthwork is involved to reshape the land. Rather, it would be better to state what type of restoration activity has been conducted, instead of using the holistic / encompassing term “restoration” to only refer to a specific subset of activities. Given the definition provided by the authors, it needs to be clarified if all areas indicated as “potential restorable’ would therefore require earthwork in order to revert to natural habitat? Or in this case does “restorable” encompass the greater suite of possible activities?

L206-207 and Table1: Is there a difference in using the term “reservation” when referring to the coastal areas, vs. the term “conservation” being used for the upland areas? Looking at Table 4, there are coastal habitats in conservation areas. Is this to just designate specifically areas that are within the 5ft contour, regardless of protection status – this was my understanding from Table 1, but it would be beneficial to have it stated more clearly in the text.

L217-219: Again, please provide some detail on the spatial resolution of these datasets. Were there any issues / challenges with alignment with spatial datasets from other sources? How were these overcome?

Results:

L263-266: Could changes in seagrass also be attributed to improved ground-truthing and photointerpretation? Perhaps some additional information on the input datasets, such as how the areas were estimated, could provide some insight (as suggested to present in a table above in Methods section).

L285-295: This is valuable information, and impressive that the authors were able to collate and document the restoration projects that have been carried out across different agencies and partnerships. This information will be very important for monitoring and reporting on progress for targets going forward, so it is a critical step to have synthesized the work that has already been carried out. An additional useful output would be a spatial representation of these projects so that the map could be updated as more projects are completed in the future (but this would be beyond the scope of this paper).

L309-310: Should this read “(coastal reservation restorable)”?

L324-335: In this section describing the targets and goals, I anticipated more detail on the potential climate change effects on the habitat map. As mentioned above in a previous comment, if this has been included then please provide a clearer description of the approach. But for example, the distribution of the intertidal habitats could be different by 2050 under SLR, and therefore adjacent upland area may be lost / transitioned into intertidal habitat. The authors have mentioned landward migration and the possibility of coastal squeeze due to infrastructure in the Discussion, but I expected it to be more clearly incorporated into the Results, given the framing of the paper in the Abstract and Introduction. If this is not the case then those previous sections need to be revisited.

Discussion:

The authors have provided a suitable Discussion to contextualise the results of the paper.

Tables:

Table3: Suggest that the heading in the second column be changed to “Landcover Category” or something similar, instead of “Habitat Type” as this column includes “Developed” and “Restorable”, which would not typically be considered as habitats.

Table4: There are some discrepancies in the table, or perhaps a sentence is needed in the title to explain why the current extent is not equal to existing conservation + proposed conservation lands. For example, Mangroves current extent = 6276, existing conservation + proposed conservation = (4516+1604) = 6120. Does this indicate an additional 156 ha of mangrove that is not within a conservation area? Also, check the rounding up of the area estimates as there is a discrepancy in values presented for total restoration opportunity = existing conservation lands restoration opportunity + proposed conservation lands restoration opportunity.

Table5: This table is arguably the most important result from the paper. The only query I have is whether the authors have considered how the map would look if these targets / goals were to be achieved. For example, the target for total intertidal to increase by 61 hectares in a mosaic – how does this factor in SLR?

It is also not immediately clear from the table how the values for the 2030 targets and 2050 goals are calculated. For example, for coastal uplands if 513 ha is the total restoration opportunity, then how was 1689 ha estimated for the 2050 target (if the current extent is 1446 ha)?

If the values for the envisioned targets and goals for 2030 and 2050 are derived from the total restoration opportunity (which was estimated from the spatial data), then would it be possible to generate maps for where these locations are? Similar to Figure 5 / Figure 6?

### 2022-12-29 revision/review_response.qmd
revision to reviewer 2 done, need to proof

@@
 
-* __Response__:
+* __Response__: Thank you for your comments on the draft manuscript.
+We have made every effort to provide responses to your concerns and have revised the text as needed, particularly for use of our approach to identify actual restoration potential (see response to your comments on lines 304-316).
+We are confident these changes have addressed your concerns and strengthened the content therein.
 
@@
 
+* __Response__: Apologies for the omission of that critical detail.
+The benchmark period is defined as circa 1950 when the earliest historical aerial photographs of Tampa Bay and its watershed were available.
+This time period is defined as pre-development in @Robison10 and the citations that precede the sentence.
+The text was amended as follows: "Priority was given to restoration activities focused on habitat types that were important for a suite of estuarine faunal guilds disproportionately lost or degraded compared to a circa 1950 benchmark period considered as pre-development."
+
 Line 127-128. 5-foot contour is the absolute elevation or elevation relative to sea level?
 
+* __Response__: The explicit definition of the 5-foot contour was previously provided on lines 209-213 as the area of land from the local Mean Lower Low Water elevation extending landward to the elevation 5 feet above Mean Sea Level.
+The definition was moved to this section for clarity.
+
 Line 140. What are the specific contents of short-term and long-term goals? The reader may want a comparison table of the two goals
+
+* __Response__: The identified short-term targets and long-term goals are provided in Table 5 and they are not presented at this point in the text because the methods that follow describe the process by which they were identified.
+However, we have added some text to provide some clarify on on these terms: "The short-term targets provided an interim set of native habitat coverages to attain within a reasonable planning horizon, at which point progress towards attaining the long-term goals will be re-assessed."
 
@@
 
+* __Response__: We assume that this comment relates to your general concern that our approach does not provide an explicit quantification of the restoration potential of wetlands, either tidal or uplands.
+We certainly agree that identification of wetland restoration areas requires a more detailed assessment of land characteristics, including but not limited to elevation and slope data.
+The approach described herein is primarily a screening and prioritization tool that requires additional work for on-site restoration planning.
+We hope that the addition to the text for your comments on lines 304-316 below addresses this concern.
+
 Line 216-220 and Figue2. In addition to the land use/land cover data, I think it is important to consider the digital elevation data as well as the relative sea level rise data.
+
+* __Response__: Please see the response to the previous comment and below regarding comments to lines 304-316.
 
@@
 
+* __Response__: As noted in the manuscript, these classifications are defined in @Ries14, but historically are defined in @Baldwin38.
+The latter citation was added.
+Additionally, we certainly agree that soil types may change over time as affected by sea level rise.
+However, our 5-foot contour considers the likely rate of sea level rise [sensu @Burke19] within the 2030 and 2050 time period used to establish the targets and goals.
+That is, the coastal stratum provides a distinction for the restoration potential of wetlands (mesic/hydric soils) as inter-tidal within the five foot contour (e.g., mangroves, salt barrens, salt marshes) and supra-tidal above the five foot contour (forested and non-forested wetlands) within the expected time period of sea level rise.
+We have added some text to make this clear: "This distinction explicitly accounts for potential salinity changes to soil properties as a function of sea-level rise based on regional projections in the time period for establishing the targets and goals."
+
 Line 248-249. The author needs to specify which data analysis methods were used?
+
+* __Response__: The analysis methods are the set of spatial operations described in the preceding text in the same section and as outlined in Figure 2.
+The sentence was revised to make this more clear.
+We have also noted two geospatial functions that were used, as examples, in the text that followed.
+Readers can view the linked repository for specific details as needed.
 
@@
 
+* __Response__: We certainly appreciate the concerns that our approach does not provide a quantitative means of identifying actual restoration potential.
+Additional information, as noted by the reviewer, is certainly needed for understanding the true restoration potential of any location withing the watershed.
+However, we have noted in several locations in the manuscript that the approach described herein is a prioritization tool that can be used as a first step for restoration planning (lines 83, 90, 113).
+Prior to this assessment, comparable tools at the scale of the watershed were not available and we believe that our approach is valuable in identifying potential restoration areas that regional managers could leverage for planning purposes.
+In fact, our use of the term restoration "potential" was meant to implicitly account for this tool as a preliminary means to guide follow-up work once potential sites are identified.
+
+     That being said, it is worth further clarifying the intent of our approach as a prioritization tool that warrants additional work should restoration projects be pursued at the locations identified by our spatial analyses.
+We have added text to two key locations to make this clear.
+
+     First, the following was added to the introduction on line 96: "These opportunity areas provide a first assessment of where restoration could occur, where on the ground assessments could be pursued to further quantify restoration potential."
+
+     Second, the following was added to the methods on line 197: "The identification of these areas on a broad spatial scale serves as a planning tool for restoration practitioners, where follow-up assessments are expected to more fully quantify restoration potential at selected sites."
+"
 Figure 1. The classification criteria for land use types are confused. I suggest using the Ramsar Convention classification system.
+
+* __Response__: The classification criteria used in the figure was based on the FLUCCS approach described in the text (lines 157), with some categories combined for clarity in the figure.
+For example, "urban/open lands" includes 13 FLUCCS categories considered relevant for combining and displaying together in the map as a single category to show the extent of development in the Tampa Bay watershed.
+The figure caption was revised to describe this approach.
+An additional citation to @Kawula18 was added in the text and figure caption.
+This classification was chosen as this is the standard approach used for the entire state of Florida.
 
@@
 Similar to Figure 5 / Figure 6?
+
+## References

### 2022-12-29 manu-draft.qmd
revision to reviewer 2 done, need to proof

@@
 A retrospective approach to setting habitat protection and restoration targets in Tampa Bay was previously used [@Lewis96; @Robison10; @Cicchetti11; @Russell15].
-Priority was given to restoration activities focused on habitat types that were important for a suite of estuarine faunal guilds disproportionately lost or degraded compared to a benchmark period.
+Priority was given to restoration activities focused on habitat types that were important for a suite of estuarine faunal guilds disproportionately lost or degraded compared to a circa 1950 benchmark period considered as pre-development.
 Criticisms of this approach included lack of consideration for future sea-level rise and other climate change factors [@Yoskowitz15], use of expanded and different habitats outside the Tampa Bay watershed [@Robison10], lack of attention to upland or freshwater wetland habitats, and little recognition of land development trends or actual available space for restoration efforts.
@@
 The general approach includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in coverage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories.
-The outcomes are spatially specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.
+These opportunity areas provide a first assessment of where restoration could occur, where on the ground assessments could be pursued to further quantify restoration potential.
+The outcomes of the approach are spatially specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.
 
@@
 Intertidal habitats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15].
-Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below).
-This 5-foot contour is an area of intense urban development and is expected to be affected by sea-level rise based on current estimates.
+Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that covers an area of land from the local Mean Lower Low Water (MLLW) elevation landward to an elevation 5 feet above Mean Sea Level.
+This 5-foot contour or coastal stratum is an area of intense urban development and is expected to be affected by sea-level rise based on current estimates [@Burke19].
 Since 1946, the St.
@@
 The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for native habitat coverage (hectares).
+The short-term targets provided an interim set of native habitat coverages to attain within a reasonable planning horizon, at which point progress towards attaining the long-term goals will be re-assessed.
 Finally, remaining restoration opportunities were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be affected by sea-level rise.
@@
 Current intertidal and supratidal habitat coverages were estimated using the 2017 SWFWMD Land Use Land Cover (LULC) maps [@SWFWMD18].
-Land use and cover types (natural and developed) are classified following the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually.
+Land use and cover types (natural and developed) are classified following the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14; @Kawula18]*.* Mangroves, salt barrens, and salt marshes were reported individually.
 While the photointerpretation of specific freshwater wetland types is often difficult, forested wetlands and non-forested wetlands can be distinguished with these data.
@@
 Identifying opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible given underlying soil conditions, expected land use change, and sea-level rise.
+The identification of these areas on a broad spatial scale serves as a planning tool for restoration practitioners, where follow-up assessments are expected to more fully quantify restoration potential at selected sites.
 
@@
 The coastal reservation native and coastal reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum.
-This area extends from the local Mean Lower Low Water (MLLW) elevation to an elevation 5 feet above Mean Sea Level and is highly vulnerable to sea-level rise [@Burke19].
 Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands or coastal uplands could be restored.
@@
 Mesic or hydric soils that occur below the 5-foot contour were assigned a restoration potential for tidal wetlands, whereas mesic or hydric soils above the 5-foot contour were assigned a restoration potential for freshwater wetlands.
+This distinction explicitly accounts for potential salinity changes to soil properties as a function of sea-level rise based on regional projections in the time period for establishing the targets and goals.
 
@@
 The second was a map that identified the restorable lands (either existing or coastal reservation) based on their restoration potential as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands.
-All analyses were conducted using the R statistical programming language [@RCT22], specifically leveraging functions from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [@Pebesma18].
+All spatial analyses described above and as outlined in @fig-gisworkflow were conducted using the R statistical programming language [@RCT22], specifically leveraging functions from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [e.g., the *st_intersection* and *st_union* functions for intersect and union operations, @Pebesma18].
 The workflows and data are provided in an open-access repository available on GitHub (<https://github.com/tbep-tech/hmpu-workflow>) [@Beck22].
@@
 ![Land use, cover, and subtidal habitats for the Tampa Bay watershed, Florida, USA.
-The watershed includes the natural hydrologic boundary with minor modifications to include partners working with the Tampa Bay Estuary Program.](figs/lulcmap.jpg){#fig-lulcmap fig-align="center"}
+The watershed includes the natural hydrologic boundary with minor modifications to include partners working with the Tampa Bay Estuary Program.
+Categories are based on the Florida Land Use Cover and Forms Classification System [@FDOT99; @SWFWMD14; @Kawula18] with relevant codes combined for presentation.](figs/lulcmap.jpg){#fig-lulcmap fig-align="center"}
 

### 2022-12-30 revision/review_response.qmd
proof revisions to reviewer 2

@@
 This time period is defined as pre-development in @Robison10 and the citations that precede the sentence.
+
     The text was amended as follows: "Priority was given to restoration activities focused on habitat types that were important for a suite of estuarine faunal guilds disproportionately lost or degraded compared to a circa 1950 benchmark period considered as pre-development."
@@
 
-* __Response__: The explicit definition of the 5-foot contour was previously provided on lines 209-213 as the area of land from the local Mean Lower Low Water elevation extending landward to the elevation 5 feet above Mean Sea Level.
+* __Response__: The explicit definition of the 5-foot contour was previously provided on lines 211-213 as the area of land from the local Mean Lower Low Water elevation extending landward to the elevation 5 feet above Mean Sea Level.
 The definition was moved to this section for clarity.
@@
 * __Response__: The identified short-term targets and long-term goals are provided in Table 5 and they are not presented at this point in the text because the methods that follow describe the process by which they were identified.
-However, we have added some text to provide some clarify on on these terms: "The short-term targets provided an interim set of native habitat coverages to attain within a reasonable planning horizon, at which point progress towards attaining the long-term goals will be re-assessed."
+
+     However, we have added some text to provide some clarity on on these terms: "The short-term targets provided an interim set of native habitat coverages to attain within a reasonable planning horizon, after which progress in attaining the long-term goals will be re-assessed."
 
@@
 
-* __Response__: We assume that this comment relates to your general concern that our approach does not provide an explicit quantification of the restoration potential of wetlands, either tidal or uplands.
+* __Response__: We assume that this comment relates to your general concern that our approach does not provide an explicit quantification of the actual restoration potential of wetlands, either tidal or freshwater.
 We certainly agree that identification of wetland restoration areas requires a more detailed assessment of land characteristics, including but not limited to elevation and slope data.
 The approach described herein is primarily a screening and prioritization tool that requires additional work for on-site restoration planning.
-We hope that the addition to the text for your comments on lines 304-316 below addresses this concern.
+We hope that our response to your comments on lines 304-316 below addresses this concern.
 
@@
 Additionally, we certainly agree that soil types may change over time as affected by sea level rise.
-However, our 5-foot contour considers the likely rate of sea level rise [sensu @Burke19] within the 2030 and 2050 time period used to establish the targets and goals.
-That is, the coastal stratum provides a distinction for the restoration potential of wetlands (mesic/hydric soils) as inter-tidal within the five foot contour (e.g., mangroves, salt barrens, salt marshes) and supra-tidal above the five foot contour (forested and non-forested wetlands) within the expected time period of sea level rise.
-We have added some text to make this clear: "This distinction explicitly accounts for potential salinity changes to soil properties as a function of sea-level rise based on regional projections in the time period for establishing the targets and goals."
+However, the coastal stratum included in the 5-foot contour considers the likely rate of sea level rise [sensu @Burke19] within the 2030 and 2050 time period used to establish the targets and goals.
+That is, the coastal stratum provides a distinction for the restoration potential of wetlands (mesic/hydric soils) as tidal within the five foot contour (mangroves, salt barrens, salt marshes) and freshwater above the five foot contour (forested and non-forested wetlands) within the expected time period of sea level rise.
+
+     We have added some text at the end of the paragraph to make this clear: "This distinction explicitly accounts for potential salinity changes to soil properties as a function of sea-level rise based on regional projections in the time period for establishing the targets and goals."
 
@@
 The sentence was revised to make this more clear.
-We have also noted two geospatial functions that were used, as examples, in the text that followed.
+We have also noted two core geospatial functions that were used, as examples, in the text that followed.
 Readers can view the linked repository for specific details as needed.
@@
 
-* __Response__: We certainly appreciate the concerns that our approach does not provide a quantitative means of identifying actual restoration potential.
-Additional information, as noted by the reviewer, is certainly needed for understanding the true restoration potential of any location withing the watershed.
-However, we have noted in several locations in the manuscript that the approach described herein is a prioritization tool that can be used as a first step for restoration planning (lines 83, 90, 113).
-Prior to this assessment, comparable tools at the scale of the watershed were not available and we believe that our approach is valuable in identifying potential restoration areas that regional managers could leverage for planning purposes.
-In fact, our use of the term restoration "potential" was meant to implicitly account for this tool as a preliminary means to guide follow-up work once potential sites are identified.
+* __Response__: We certainly appreciate the concerns that our approach does not provide a more comprehensive quantitative means of identifying actual restoration potential.
+Additional information, as noted by the reviewer, is certainly needed for understanding the true restoration potential of any location within the watershed.
+However, we have noted in several locations in the manuscript that the approach described herein is a prioritization tool that can be used as a first step for restoration planning (e.g., lines 83, 90, 113).
+Prior to this assessment, comparable tools at the watershed scale were not available and we believe that our approach is valuable in identifying potential restoration areas that regional managers could leverage for planning purposes.
+In fact, our use of the term restoration "potential" was meant to describe this tool as a preliminary means to guide follow-up work once sites are identified from the provided opportunity areas.
 
@@
 The figure caption was revised to describe this approach.
-An additional citation to @Kawula18 was added in the text and figure caption.
+An additional citation to @Kawula18 was also added in the text and figure caption.
 This classification was chosen as this is the standard approach used for the entire state of Florida.

### 2022-12-30 manu-draft.qmd
proof revisions to reviewer 2

@@
 These opportunity areas provide a first assessment of where restoration could occur, where on the ground assessments could be pursued to further quantify restoration potential.
-The outcomes of the approach are spatially specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.
+The outcomes of the approach are also spatially specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.
 
@@
 The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for native habitat coverage (hectares).
-The short-term targets provided an interim set of native habitat coverages to attain within a reasonable planning horizon, at which point progress towards attaining the long-term goals will be re-assessed.
+The short-term targets provided an interim set of native habitat coverages to attain within a reasonable planning horizon, after which progress in attaining the long-term goals will be re-assessed.
 Finally, remaining restoration opportunities were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be affected by sea-level rise.
@@
 The watershed includes the natural hydrologic boundary with minor modifications to include partners working with the Tampa Bay Estuary Program.
-Categories are based on the Florida Land Use Cover and Forms Classification System [@FDOT99; @SWFWMD14; @Kawula18] with relevant codes combined for presentation.](figs/lulcmap.jpg){#fig-lulcmap fig-align="center"}
+Categories are based on the Florida Land Use Cover and Forms Classification System [@FDOT99; @SWFWMD14; @Kawula18] with relevant codes combined for presentation in the figure.](figs/lulcmap.jpg){#fig-lulcmap fig-align="center"}
 

### 2022-12-30 revision/review_response.qmd
started revisions for reviewer 3

@@
 
-* __Response__:
+* __Response__: Thank you for your detailed comments on our manuscript.  We have made every effort to address your concerns noted below and are confident this has improved the draft.
 
@@
 That is not to say that the entire Abstract is unsuitable, rather that it needs to be amended to better reflect the contents of the paper, rather than a summary of the approach.
+
+* __Response__:
 
@@
 
+* __Response__:
+
 Methods:
@@
 
+* __Response__: Corrected.
+
 L117-126: If possible, it would help to have this information on the habitat layers and respective source material presented in a table – even as supplementary information.
+
+* __Response__: Text was added to refer the reader to a new supplement Table S1.
 
@@
 This will also make for a more straightforward understanding of the SLR projection, which is provided in cm (L132).
+
+* __Response__: This sentence was revised as follows "...location relative to the 5-foot contour (~1.5m elevation) that covers an area of land...".
 
@@
 
+* __Response__:
+
 L154-165: Please indicate the type and scale of the spatial datasets for the habitat.
@@
 
+* __Response__: Both the subtidal and inter/supratidal data layers are vector products produced by photo-interpretation.  Details were added as follows.
+
+     For the subtidal layers, details were added starting on line 150: "These data include vector polygons coverages of the major subtidal habitat types in Tampa Bay, as interpreted from 1:24,000 scale natural color aerial photographs flown in winter 2018 under cloud free conditions.
+Accuracy assessments of the photo-interpreted map product included field verification by random sample points, with a requirement of 90% accuracy for the seagrass categories.
+The minimum mapping unit for seagrass polygons is reported as 0.25 acres."
+
+     For the inter/supratidal layers, details were added on line 158: "Similar methods as the subtidal habitats described above are used for the intertidal and supratidal coverage maps, although at a slightly higher spatial resolution (1:12,000)."
+
 L170-175: Please provide more detail on how the habitat coverage change analysis was performed.
@@
 Please specify how the change from one habitat to another was estimated in order to generate the alluvial diagrams.
+
+* __Response__: Additional detail was provided as follows: "Specifically, the spatial datasets were unioned and the total areal change of the polygons for each habitat type was quantified by taking the difference between the two terminal years.
+For example, the area that remained as native uplands between the 1990 and 2017 intertidal and suptratidal layers was quantified, whereas the area that changed from native uplands to another habitat category was also quantified.
+This process was repeated for all native habitats, including developed and restorable lands (described below)."
 
@@
 
+* __Response__:
+
 L206-207 and Table1: Is there a difference in using the term “reservation” when referring to the coastal areas, vs.
@@
 
+* __Response__:
+
 L217-219: Again, please provide some detail on the spatial resolution of these datasets.
@@
 How were these overcome?
+
+* __Response__: Details on these data were added as follows: "The FNAI data are Florida Managed Areas as vector-based polygons of public and some private lands identified as having natural resource value and that are being managed at least partially for conservation.
+The source data for this layer are provided to FNAI directly from the managing agency in digital format or as paper source maps that are digitized using appropriate topographic quadrangles, ortho-imagery, and property appraiser parcel data at a minimum spatial resolution of 1:5,000."
+
+     A citation to @FNAI20 was also added.
+
+     The common projection used for all analyses was also added to line 251: "All spatial data were transformed to the NAD83(2011) / Florida West (ftUS) projection prior to analysis."
 
@@
 
+* __Response__: The change in seagrasses in Tampa Bay over the last three decades is well-documented [e.g., @Greening14; @Sherwood17].
+The documentation of this recovery has been made possible primarily from efforts of the Southwest Florida Water Management District in providing accurate and comparable coverage maps on an approximate biennial basis since the late 1980s (as described in the data sources).
+Although some minor changes in the FLUCCS codes have occurred during this period, the methods for documenting seagrasses have been consistent.
+We are confident that these changes are not a product of changes in methodologies.
+We hope that our addition to the text describing more detail on the data sources and methods has provided some clarity.
+
 L285-295: This is valuable information, and impressive that the authors were able to collate and document the restoration projects that have been carried out across different agencies and partnerships.
@@
 
+* __Response__:
+
 L309-310: Should this read “(coastal reservation restorable)”?
+
+* __Response__: Corrected.
 
@@
 If this is not the case then those previous sections need to be revisited.
+
+* __Response__:
 
@@
 
+* __Response__: The column header was changed to "Landcover Category".  The table caption was also changed.
+
 Table4: There are some discrepancies in the table, or perhaps a sentence is needed in the title to explain why the current extent is not equal to existing conservation + proposed conservation lands.
@@
 
+* __Response__:
+
 Table5: This table is arguably the most important result from the paper.
@@
 
+* __Response__:
+
 It is also not immediately clear from the table how the values for the 2030 targets and 2050 goals are calculated.
 For example, for coastal uplands if 513 ha is the total restoration opportunity, then how was 1689 ha estimated for the 2050 target (if the current extent is 1446 ha)?
+
+* __Response__:
 
@@
 
+* __Response__:
+
 ## References

### 2022-12-30 manu-draft.qmd
started revisions for reviewer 3

@@
 Native habitats in the watershed include pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands.
-The watershed is heavily developed with an estimated population of 3.3 million people in the four major counties [@Rayer20]*.* Numerous anthropogenic changes have altered the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom, mining activities), alteration of hydrology, and destruction and fragmentation of habitat from development.
+The watershed is heavily developed with an estimated population of 3.3 million people in the four major counties [@Rayer20]*.* Numerous anthropogenic changes have altered the natural habitats of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom, mining activities), alteration of hydrology, and destruction and fragmentation of habitat from development.
 
@@
 Intertidal habitats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15].
-Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that covers an area of land from the local Mean Lower Low Water (MLLW) elevation landward to an elevation 5 feet above Mean Sea Level.
+Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Data sources are described below and in Table S1.
+Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour (~1.5m elevation) that covers an area of land from the local Mean Lower Low Water (MLLW) elevation landward to an elevation 5 feet above Mean Sea Level.
 This 5-foot contour or coastal stratum is an area of intense urban development and is expected to be affected by sea-level rise based on current estimates [@Burke19].
@@
 For subtidal habitats, 2018 data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19].
-Historical datasets for subtidal habitats began in 1988 with updates occurring on an approximate biennial basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.
+These data include vector polygons coverages of the major subtidal habitat types in Tampa Bay, as interpreted from 1:24,000 scale natural color aerial photographs flown in winter 2018 under cloud free conditions.
+Accuracy assessments of the photo-interpreted map product included field verification by random sample points, with a requirement of 90% accuracy for the seagrass categories.
+The minimum mapping unit for seagrass polygons is reported as 0.25 acres.
+Historical datasets for subtidal habitats using identical methods began in 1988 with updates occurring on an approximate biennial basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.
 
 Current intertidal and supratidal habitat coverages were estimated using the 2017 SWFWMD Land Use Land Cover (LULC) maps [@SWFWMD18].
-Land use and cover types (natural and developed) are classified following the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14; @Kawula18]*.* Mangroves, salt barrens, and salt marshes were reported individually.
+Land use and cover types (natural and developed) are classified following the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14; @Kawula18]*.* Similar methods as the subtidal habitats described above are used for the intertidal and supratidal coverage maps, although at a slightly higher spatial resolution (1:12,000).
+Mangroves, salt barrens, and salt marshes were reported individually.
 While the photointerpretation of specific freshwater wetland types is often difficult, forested wetlands and non-forested wetlands can be distinguished with these data.
@@
 This required an intersection of the data layers to quantify if habitat types were unchanged or changed for any given location and identifying the type of change (e.g., seagrass to tidal flats).
+Specifically, the spatial datasets were unioned and the total areal change of the polygons for each habitat type was quantified by taking the difference between the two terminal years.
+For example, the area that remained as native uplands between the 1990 and 2017 intertidal and suptratidal layers was quantified, whereas the area that changed from native uplands to another habitat category was also quantified.
+This process was repeated for all native habitats, including developed and restorable lands (described below).
 The results were summarized as Alluvial diagrams showing relative proportions of habitat change by type and between years [@Allaire17].
@@
 The existing and proposed native and restorable areas were those that occurred in public lands that are currently acquired or proposed for acquisition.
-To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory and permit databases of conservation and drainage easements.
+To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory (FNAI) and permit databases of conservation and drainage easements [@FNAI20].
+The FNAI data are Florida Managed Areas as vector-based polygons of public and some private lands identified as having natural resource value and that are being managed at least partially for conservation.
+The source data for this layer are provided to FNAI directly from the managing agency in digital format or as paper source maps that are digitized using appropriate topographic quadrangles, ortho-imagery, and property appraiser parcel data at a minimum spatial resolution of 1:5,000.
+Intersecting the native and restorable lands in these areas produced four unique opportunity areas: existing conservation native, proposed conservation native, existing conservation restorable, and proposed conservation restorable.
 This workflow created the existing conservation and proposed conservation layers in @fig-gisworkflow.
-Intersecting the native and restorable lands in these areas produced four unique opportunity areas: existing conservation native, proposed conservation native, existing conservation restorable, and proposed conservation restorable.
 
@@
 All spatial analyses described above and as outlined in @fig-gisworkflow were conducted using the R statistical programming language [@RCT22], specifically leveraging functions from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [e.g., the *st_intersection* and *st_union* functions for intersect and union operations, @Pebesma18].
+All spatial data were transformed to the NAD83(2011) / Florida West (ftUS) projection prior to analysis.
 The workflows and data are provided in an open-access repository available on GitHub (<https://github.com/tbep-tech/hmpu-workflow>) [@Beck22].
@@
 Native habitats currently protected (existing conservation native), proposed for protection (proposed conservation native), or in the coastal stratum (coastal reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW).
-Similarly, restorable lands currently protected (existing conservation restorable), proposed for protection (proposed conservation restorable), or in the coastal stratum (coastal reservation native) totaled `r areas$restorable$areaha` (`r areas$restorable$areaper` of the watershed).
+Similarly, restorable lands currently protected (existing conservation restorable), proposed for protection (proposed conservation restorable), or in the coastal stratum (coastal reservation restorable) totaled `r areas$restorable$areaha` (`r areas$restorable$areaper` of the watershed).
 Understandably, most of the native and restorable lands occurred in undeveloped areas in northern and southeastern areas of the watershed (@fig-restmap).
@@
 ```{r}
-#| tbl-cap: "Table 3: Change over time in hectares for intertidal and supratidal habitats in Tampa Bay.
+#| tbl-cap: "Table 3: Change over time in hectares for intertidal and supratidal landcover in Tampa Bay.
 Columns show years with available data and the final two columns show the change and percent change from 1990 to 2017."

### 2022-12-30 supplement.qmd
started revisions for reviewer 3

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "Supplement: Addressing climate change and development pressures in an urban estuary through habitat restoration planning"
shorttitle: "Tampa Bay watershed habitat restoration planning"
bibliography: refs.bib
csl: frontiers-in-ecology-and-evolution.csl
author: 
  - Marcus W Beck:
      email: mbeck@tbep.org
      institute: [TBEP]
      correspondence: true
  - Douglas E Robison:
      email: DRobison@esassoc.com 
      institute: [ESA]
      correspondence: false
  - Gary E Raulerson:
      email: gmangrove@gmail.com
      correspondence: false
  - Maya C Burke:
      email: mburke@tbep.org
      institute: [TBEP]
      correspondence: false
  - Justin Saarinen:
      email: justinsaarinen@gmail.com 
      institute: [ESA]
      correspondence: false
  - Christine Sciarrino:
      email: CSciarrino@esassoc.com 
      institute: [ESA]
      correspondence: false
  - Edward T Sherwood:
      email: esherwood@tbep.org 
      institute: [TBEP]
      correspondence: false
  - David A Tomasko:
      email: dave@sarasotabay.org
      institute: [SBEP]
      correspondence: false
institute:
  - TBEP: Tampa Bay Estuary Program, St. Petersburg, Florida 33701 USA
  - ESA: Environmental Science Associates, Tampa, Florida 33609 USA
  - SBEP: Sarasota Bay Estuary Program, Sarasota, Florida 34236 USA

filters: 
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua

execute: 
  echo: false
  warning: false
  ft.keepnext: false
  ft.align: left
---

```{r, results = 'hide'}
library(knitr)
library(here)
library(flextable)
library(ftExtra)
library(dplyr)
```

```{r}
#| tbl-cap: "Table S1: Data sources for habitat types grouped by subtidal, intertidal, and supratidal strata."

data(strata)

tab <- strata %>%
  mutate(
    Source = case_when(
      HMPU_TARGETS == 'Hard Bottom' ~ '@SWFWMD18;@CSA19',
      HMPU_TARGETS %in% c('Artificial Reefs', 'Living Shorelines') ~ '@Robison20',
      HMPU_TARGETS %in% c('Seagrasses', 'Oyster Bars', 'Tidal Flats') ~ '@SWFWMD19',
      HMPU_TARGETS %in% c('Mangrove Forests', 'Salt Barrens', 'Salt Marshes') ~ '@SWFWMD18', 
      HMPU_TARGETS == 'Tidal Tributaries' ~ '@Wessel22',
      HMPU_TARGETS %in% c('Coastal Uplands', 'Non-Forested Freshwater Wetlands', 
                          'Forested Freshwater Wetlands', 'Native Uplands') ~ '@SWFWMD18'
    ), 
    HMPU_TARGETS = as.character(HMPU_TARGETS)
  ) %>% 
  na.omit() %>% 
  arrange(Category, HMPU_TARGETS) %>% 
  select(
    Stratum = Category, 
    `Habitat Type` = HMPU_TARGETS, 
    Source
  )

thm <- function(x){
  flextable::align(x, align = 'center', part = 'header') %>%
    flextable::align(align = 'left', j = 1, part = 'all') %>%
    flextable::align(align = 'left', j = 2, part = 'body') %>%
    flextable::border_inner() %>%
    flextable::width(width = 0.7, j = 1) %>%
    flextable::width(width = 5.8 / (ncol_keys(.) - 1), j = 2:ncol_keys(.)) %>%
    flextable::padding(padding = 0, part = 'all') %>%
    flextable::fontsize(size = 8, part = 'all')
}

tab %>% 
  as_grouped_data('Stratum') %>% 
  flextable() %>%
  colformat_md() %>%
  thm
```

## References {.unnumbered}

### 2022-12-30 R/tabs.R

@@ -3,6 +3,7 @@ library(tidyverse)
 library(here)
 library(sf)
 library(units)
+library(ftExtra)
 
 source(here('R/funcs.R'))
 
@@ -88,31 +89,31 @@ gaincol <- 'green'
 # format for the table
 totab <- acres %>%
   rename(
-    `Habitat Type` = HMPU_TARGETS,
+    `Landcover Category` = HMPU_TARGETS,
     yr = name,
     Hectares = Acres
   ) %>%
   filter(yr <= 2017) %>%
   mutate(
     Stratum = case_when(
-      `Habitat Type` %in% c('Mangrove Forests', 'Salt Barrens', 'Salt Marshes') ~ 'Intertidal',
+      `Landcover Category` %in% c('Mangrove Forests', 'Salt Barrens', 'Salt Marshes') ~ 'Intertidal',
       T ~ 'Supratidal'
     ),
     Hectares = Hectares / 2.471
   ) %>%
   tidyr::pivot_wider(names_from = yr, values_from = Hectares) %>%
-  filter(`Habitat Type` != 'Open Water') %>%
+  filter(`Landcover Category` != 'Open Water') %>%
   mutate(
     `1990 to 2017` = `2017` - `1990`,
     `% change` = 100 * (`2017` - `1990`) / `1990`,
     chgcol = case_when(
-      `% change` < 0 & `Habitat Type` != 'Developed' ~ 0,
-      `% change` >= 0 & `Habitat Type` != 'Developed' ~ 1,
-      `% change` < 0 & `Habitat Type` == 'Developed' ~ 1,
-      `% change` >= 0 & `Habitat Type` == 'Developed' ~ 0,
+      `% change` < 0 & `Landcover Category` != 'Developed' ~ 0,
+      `% change` >= 0 & `Landcover Category` != 'Developed' ~ 1,
+      `% change` < 0 & `Landcover Category` == 'Developed' ~ 1,
+      `% change` >= 0 & `Landcover Category` == 'Developed' ~ 0,
     )
   ) %>%
-  arrange(Stratum, `Habitat Type`) %>%
+  arrange(Stratum, `Landcover Category`) %>%
   as_grouped_data(groups = 'Stratum')
 
 thm <- function(x){

### 2022-12-30 revision/review_response.qmd
more revisions for reviewer 3

@@
 
-* __Response__:
+* __Response__: Thanks for this suggestion.  We have revised the abstracts to include more relevant content from the paper. The entire abstract from line 17 down was edited:
+
+     "Although significant gains in subtidal habitats have been observed, expansion of mangroves into salt marshes and loss of native upland habitats to development highlights the need to target these locations for restoration.
+The long-term loss of potentially restorable lands to both coastal and upland development further underscores the diminishing restoration opportunities in the watershed.
+The established targets and goals identified habitats to maintain at their present level (e.g., mangroves) and those that require additional progress (e.g., oyster bars) based on past trends and an expected level of effort given the restoration history of the region.
+The new approach also accounts for the future effects of sea-level rise, climate change, and watershed development by prioritizing native coastal habitats relative to subtidal or upland areas.
+The restoration potential for coastal habitats was 624 ha (0.1% of the watershed) for tidal wetlands and 128 ha (< 0.1% of the watershed) for coastal uplands, whereas the total restoration opportunity in the watershed was 17,205 ha (2.9% of the watershed).
+Maps were created to identify the restoration opportunities where practitioners could focus efforts to achieve the defined targets and goals.
+Methods for repeatable analyses are also available using an open source workflow."
 
@@
 
-* __Response__:
+* __Response__: The first two paragraphs were combined and revised to make a more cohesive introduction to the paper:
+
+     "The health of estuarine systems and coastal habitats is tightly linked to land use and management of the watershed [@Yoskowitz15].
+Coastal habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18].
+Anthropogenic stressors can negatively impact the services provided by coastal habitats and restoration practitioners must consider the anticipated effects of these stressors during planning [@Elliott07;@White17].
+The combined effects of land development and climate change are especially problematic for prioritizing habitat restoration activities in coastal environments.
+Habitat changes in response to climate change include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].
+Landward migration of critical habitats in response to sea-level rise may not be possible due to anthropogenic barriers in the watershed.
+Sea-level rise can occur quicker than landward migration of salt marshes and the upland slope may already be lost to urban development and hardening [@Titus09].
+Given projected habitat losses and the limited resources available, appropriate and realistic sites for restoration need to be identified that account for future stressors and past trends."
 
@@
 
-* __Response__: Text was added to refer the reader to a new supplement Table S1.
+* __Response__: Text was added to refer the reader to a new Table S1 in supplement.
 
@@
 
-* __Response__:
+* __Response__: The expected effects of sea level rise are implicitly accounted for in the 5-foot contour or coastal stratum that was used to differentiate the opportunity areas in the coastal margin from those in the upland.
+By placing emphasis on these areas, it is anticipated that restoration practitioners can target these locations differently based on the diminishing opportunities in these areas.
+For example, figures 5 and 6 identify these areas differently from the other coastal strata.
+The following text was added to line 132 to make this clear:
+
+     "The coastal stratum within the 5-foot contour is used to better identify and prioritize coastal habitats at risk of landward migration and coastal development given that it includes the area of land within the sea level rise projections."
 

### 2022-12-30 manu-draft.qmd
more revisions for reviewer 3

@@
 To address these pressures, restoration practitioners require robust tools that identify reasonable goals given historical land use trends, current status of native habitats, and anticipated future impacts from coastal stressors.
-A target-setting approach for native habitats was created for the Tampa Bay watershed that identifies current restoration opportunities and establishes short-term (2030) targets and long-term (2050) goals.
+A restoration framework for native habitats was created for the Tampa Bay watershed that identifies current opportunities and establishes short-term (2030) targets and long-term (2050) goals.
 The approach was informed through a three-decade habitat change analysis and over forty years of habitat restoration projects in the region.
-Restoration goals were defined based on what is possible today and the projected needs for the future, rather than attempting to replicate past ecological conditions.
-The new approach also accounts for the expected impacts of sea-level rise, climate change, and watershed development - stressors which are pervasive in Florida.
-The resulting habitat goals are provided as maps that identify remaining restoration opportunities, while also providing an approach for the entire watershed that targets subtidal, intertidal, and uplands.
-This approach also supports coastal planning decisions that need to address competing interests and could be applied in other coastal settings where sustainable urbanization practices need to co-exist with natural environments.
-Methods for repeatable analyses are also available using an open source workflow to update progress over time and for adoption by others.
+Although significant gains in subtidal habitats have been observed, expansion of mangroves into salt marshes and loss of native upland habitats to development highlights the need to target these locations for restoration.
+The long-term loss of potentially restorable lands to both coastal and upland development further underscores the diminishing restoration opportunities in the watershed.
+The established targets and goals identified habitats to maintain at their present level (e.g., mangroves) and those that require additional progress (e.g., oyster bars) based on past trends and an expected level of effort given the restoration history of the region.
+The new approach also accounts for the future effects of sea-level rise, climate change, and watershed development by prioritizing native coastal habitats relative to subtidal or upland areas.
+The restoration potential for coastal habitats was `r areas$potential[['Tidal Wetlands']]$areaha` (`r areas$potential[['Tidal Wetlands']]$areaper` of the watershed) for tidal wetlands and `r areas$potential[['Coastal Uplands']]$areaha` (`r areas$potential[['Coastal Uplands']]$areaper` of the watershed) for coastal uplands, whereas the total restoration opportunity in the watershed was `r areas$potentialall$areaha` (`r areas$potentialall$areaper` of the watershed).
+Maps were created to identify the restoration opportunities where practitioners could focus efforts to achieve the defined targets and goals.
+Methods for repeatable analyses are also available using an open source workflow.
 
@@
 Coastal habitats provide multiple ecosystem services, including wildlife shelter and migratory corridors [@Yoskowitz15], fisheries production [@Houde93], water quality improvement [@Kushlan90; @Sprandel00; @AvilaGarcia20], erosion and flood attenuation [@Calil15; @Menendez18], carbon sequestration [@Dontis20] and recreation [@Chung18].
-Pervasive stressors can negatively impact the services provided by these habitats.
-Anthropogenic land development can remove or substantially modify existing habitats and can alter hydrology of nearby streams and rivers [@Theobald97; @Sim14].
+Anthropogenic stressors can negatively impact the services provided by coastal habitats and restoration practitioners must consider the anticipated effects of these stressors during planning [@Elliott07;@White17].
+The combined effects of land development and climate change are especially problematic for prioritizing habitat restoration activities in coastal environments.
 Habitat changes in response to climate change include landward migration of mangroves into salt marshes, upstream migration of salt marshes within tidal tributaries, and upland forest migration [@Brinson95; @Vogelmann12; @Cavanaugh19].
-Management of coastal habitats requires an understanding of how stressors influence these resources, while balancing competing societal uses.
-
-The combined effects of land development and climate change are especially problematic for prioritizing habitat restoration activities in coastal environments.
 Landward migration of critical habitats in response to sea-level rise may not be possible due to anthropogenic barriers in the watershed.
 Sea-level rise can occur quicker than landward migration of salt marshes and the upland slope may already be lost to urban development and hardening [@Titus09].
-Coastal development and climate change diminish the available space for future restoration in urban estuaries, thereby impacting ecosystem services provided by these habitats and the wildlife they support [@Enwright15].
 Given projected habitat losses and the limited resources available, appropriate and realistic sites for restoration need to be identified that account for future stressors and past trends.
@@
 Projections from the year 2000 to 2100 suggest sea levels can increase between 58 and 259 cm in the region [@Burke19].
+The coastal stratum within the 5-foot contour is used to better identify and prioritize coastal habitats at risk of landward migration and coastal development given that it includes the area of land within the sea level rise projections.
 

### 2022-12-30 revision/review_response.qmd
almost done with revisions for reviewer 3

@@
 The new approach also accounts for the future effects of sea-level rise, climate change, and watershed development by prioritizing native coastal habitats relative to subtidal or upland areas.
-The restoration potential for coastal habitats was 624 ha (0.1% of the watershed) for tidal wetlands and 128 ha (< 0.1% of the watershed) for coastal uplands, whereas the total restoration opportunity in the watershed was 17,205 ha (2.9% of the watershed).
+The restoration potential for coastal habitats on existing conservation lands was 624 ha for tidal wetlands and 128 ha for coastal uplands.
 Maps were created to identify the restoration opportunities where practitioners could focus efforts to achieve the defined targets and goals.
@@
 
-* __Response__: The expected effects of sea level rise are implicitly accounted for in the 5-foot contour or coastal stratum that was used to differentiate the opportunity areas in the coastal margin from those in the upland.
+* __Response__: The expected effects of sea level rise are implicitly accounted for in the 5-foot contour that was used to differentiate the opportunity areas in the coastal margin from those in the upland.
 By placing emphasis on these areas, it is anticipated that restoration practitioners can target these locations differently based on the diminishing opportunities in these areas.
 For example, figures 5 and 6 identify these areas differently from the other coastal strata.
+This area also accounts for the projected increases in sea level rise in @Burke19.
 The following text was added to line 132 to make this clear:
@@
 
-* __Response__:
+* __Response__: Thank you for the clarification and we agree that adopting a standardized and more widely accepted definition is more appropriate.
+However, the projects in the database have already been classified as restoration/enhancement and a distinction based on the previous definition is useful to local restoration practitioners that are familiar with the projects.
+For clarity, we have included the SER definition as a general description and amended the text to better justify the previous description.
+The addition starting on line 179 is as follows:
+
+     "Here and throughout, restoration describes the process of assisting the recovery of an ecosystem that has been degraded, damaged, or destroyed [@Gann19].
+Restoration projects in the database were also those that involved earthwork to reshape the land or the addition of structural elements (e.g., rock).
+This distinct categorization is useful for restoration practitioners familiar with the projects in the region."
 
@@
 
-* __Response__:
+* __Response__: Apologies for the confusion.
+The reservation areas are those in the coastal stratrum that are currently not in existing or proposed conservation areas.
+The sentence on line 211 was revised as follows:
+
+     "The coastal reservation native and coastal reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum and do not occur in existing or proposed conservation areas (described in the following paragraph)."
 
@@
 
-* __Response__:
+* __Response__: Yes, we agree that the effort to create and synthesize these data was one of the more useful products resulting from this work.
+Current discussions have focused on how to maintain the integrity of these data as additional work is completed in the future.
+The limitations discussed in the second paragraph of section 4.2 alludes to these challenges.
+A link to the online location of this database, that inclues a map, was added to line 188 in the methods (<https://www.tampabay.wateratlas.usf.edu/restoration/>).
 
@@
 
-* __Response__:
+* __Response__: We understand the general concern that have included explicit projections of how land use and cover may chagne with climate change and development.
+The methods we have used implicitly account for these factors and are our best estimate at what can be reasonably achieved by 2030 and 2050 based on the assessent.
+We hope that revisions to some of the earlier comments have also made this more clear (e.g., lines 130 - 132).
+We have added the following text to this section for clarify:
+
+     "These targets and goals do not consider an explicit projection of how habitats are expected to change as a result of climate change and anticipated development because no such estimates are available.
+However, the methods implicitly account for these anticipated changes by differentiating the watershed by strata and setting the targets and goals based on past trends that are affected both by climate change and development trajectories.
+The methods herein provide the best estimate of what restoration is likely to be achieved over the next few decades."
 
@@
 
-* __Response__:
+* __Response__: Correct, the current extent is the sum of existing and proposed conservation, plus those not in conservation.
+A sentence was added to the caption.
+The rounding issues for total restoration opportunity were also fixed.
 
@@
 
-* __Response__:
+* __Response__: Please see our response to your previous comment on lines 324-335.
+One could potentially understand how the land use and cover may change if the targets and goals were met by viewing the restoration potential in Figure 6.
+However, the restoration potential is general and does not provide specific recommendations on which native habitats to restore, i.e., tidal wetland is indicated and not mangrove forests, salt barrens, or salt marshes.
+Additional data would be needed to determine the appropriate type of tidal wetland and we leave that to the restoration practiioners to conduct site-level assessments when a restoration is planned.
+Also, our responses to the other reviewers address this concern regarding the true restoration potential of wetlands.
 
@@
 
-* __Response__:
+* __Response__: Please see our response to your previous comment about Table 5.
 

### 2022-12-30 manu-draft.qmd
almost done with revisions for reviewer 3

@@
 The new approach also accounts for the future effects of sea-level rise, climate change, and watershed development by prioritizing native coastal habitats relative to subtidal or upland areas.
-The restoration potential for coastal habitats was `r areas$potential[['Tidal Wetlands']]$areaha` (`r areas$potential[['Tidal Wetlands']]$areaper` of the watershed) for tidal wetlands and `r areas$potential[['Coastal Uplands']]$areaha` (`r areas$potential[['Coastal Uplands']]$areaper` of the watershed) for coastal uplands, whereas the total restoration opportunity in the watershed was `r areas$potentialall$areaha` (`r areas$potentialall$areaper` of the watershed).
+The restoration potential for coastal habitats on existing conservation lands was `r areas$potential[['Tidal Wetlands']]$areaha` for tidal wetlands and `r areas$potential[['Coastal Uplands']]$areaha` for coastal uplands.
 Maps were created to identify the restoration opportunities where practitioners could focus efforts to achieve the defined targets and goals.
@@
 
-Restoration projects conducted over the past 40 years were quantified for each of the major habitat types to inform expectations for setting short-term targets and long-term goals.
-Restoration was defined as any activity that involved earthwork to reshape the land or the addition of structural elements (e.g., rock).
+Restoration and enhancement projects conducted over the past 40 years were quantified for each of the major habitat types to inform expectations for setting short-term targets and long-term goals.
+Here and throughout, restoration describes the process of assisting the recovery of an ecosystem that has been degraded, damaged, or destroyed [@Gann19].
+Restoration projects in the database were also those that involved earthwork to reshape the land or the addition of structural elements (e.g., rock).
+This distinct categorization is useful for restoration practitioners familiar with the projects in the region.
 Enhancement was defined as any activity not including earthwork that improved the environment (e.g., planting native vegetation, invasive species or debris removal, prescribed burns, etc.).
@@
 Data gaps were supplemented by archival research, site visits, contacting entities, and expert knowledge from local professionals.
+The synthesized dataset is available at <https://www.tampabay.wateratlas.usf.edu/restoration/>.
 
@@
 Collectively, each of these unique products are considered the opportunity areas in the Tampa Bay watershed (Table 1).
-The coastal reservation native and coastal reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum.
+The coastal reservation native and coastal reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum and do not occur in existing or proposed conservation areas (described in the following paragraph).
 Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands or coastal uplands could be restored.
@@
 Combining the restorable lands on existing conservation areas and in the coastal stratum with soils data provided a spatial summary of the restoration potential grouped by habitat type (@fig-restmap).
-A total of `r areas$potentialall$areaha` (`r areas$potentialall$areaper` of the watershed) of potentially restorable lands were identified, further partitioned as coastal uplands (`r areas$potential[['Coastal Uplands']]$areaha`, `r areas$potential[['Coastal Uplands']]$areaper` of the watershed), freshwater wetlands (`r areas$potential[['Freshwater Wetlands']]$areaha`, `r areas$potential[['Freshwater Wetlands']]$areaper` of the watershed), native uplands (`r areas$potential[['Native Uplands']]$areaha`, `r areas$potential[['Native Uplands']]$areaper` of the watershed), or tidal wetlands (`r areas$potential[['Tidal Wetlands']]$areaha`, `r areas$potential[['Tidal Wetlands']]$areaper` of the watershed).
+A total of `r areas$potentialall$areaha` (`r areas$potentialall$areaper` of the watershed) of potentially restorable lands on existing conservation areas were identified, further partitioned as coastal uplands (`r areas$potential[['Coastal Uplands']]$areaha`, `r areas$potential[['Coastal Uplands']]$areaper` of the watershed), freshwater wetlands (`r areas$potential[['Freshwater Wetlands']]$areaha`, `r areas$potential[['Freshwater Wetlands']]$areaper` of the watershed), native uplands (`r areas$potential[['Native Uplands']]$areaha`, `r areas$potential[['Native Uplands']]$areaper` of the watershed), or tidal wetlands (`r areas$potential[['Tidal Wetlands']]$areaha`, `r areas$potential[['Tidal Wetlands']]$areaper` of the watershed).
 
@@
 Identifying short-term (2030) targets and long-term (2050) goals for the restoration extent of native habitats in Tampa Bay was informed by the assessment of current extents, past trends, and relative effort for past restoration and enhancement projects.
+These targets and goals do not consider an explicit projection of how habitats are expected to change as a result of climate change and anticipated development because no such estimates are available.
+However, the methods implicitly account for these anticipated changes by differentiating the watershed by strata and setting the targets and goals based on past trends that are affected both by climate change and development trajectories.
+The methods herein provide the best estimate of what restoration is likely to be achieved over the next few decades.
+
 Table 5 shows the targets and goals identified through this analysis and the associated rationale.
@@
 Proposed conservation lands are those identified for acquisition.
+Current extent is the sum of existing and proposed conservation lands, plus those not in conservation.
 Total restoration opportunity does not account for lands currently existing or proposed for conservation.

### 2022-12-30 R/funcs.R

@@ -330,7 +330,7 @@ curex_fun <- function(lulc, subt, hard, arti, tidt, livs, coastal, fluccs, strat
     ) %>%
     spread(typ, Hectares, fill = 0) %>%
     mutate(
-      `total restorable` = `restorable Existing` + `restorable Proposed`
+      `total restorable` = round(`restorable Existing`, 0) + round(`restorable Proposed`, 0)
     )
 
   # final table

### 2023-01-03 revision/review_response.qmd
finished first round of revisions and proofed

@@
 
-* __Response__: Thanks for this suggestion.  We have revised the abstracts to include more relevant content from the paper. The entire abstract from line 17 down was edited:
+* __Response__: Thanks for this suggestion.  We have revised the abstract to include more relevant content from the paper. The entire abstract from line 17 down was edited:
 
@@
 The new approach also accounts for the future effects of sea-level rise, climate change, and watershed development by prioritizing native coastal habitats relative to subtidal or upland areas.
-The restoration potential for coastal habitats on existing conservation lands was 624 ha for tidal wetlands and 128 ha for coastal uplands.
-Maps were created to identify the restoration opportunities where practitioners could focus efforts to achieve the defined targets and goals.
-Methods for repeatable analyses are also available using an open source workflow."
+Maps were created to identify the restoration opportunities where practitioners could focus efforts to achieve the targets and goals, with methods for repeatable analyses also available using an open source workflow."
 
@@
 By placing emphasis on these areas, it is anticipated that restoration practitioners can target these locations differently based on the diminishing opportunities in these areas.
-For example, figures 5 and 6 identify these areas differently from the other coastal strata.
+For example, figures 5 and 6 distinguish these areas from the other coastal strata.
 This area also accounts for the projected increases in sea level rise in @Burke19.
@@
 
-     For the subtidal layers, details were added starting on line 150: "These data include vector polygons coverages of the major subtidal habitat types in Tampa Bay, as interpreted from 1:24,000 scale natural color aerial photographs flown in winter 2018 under cloud free conditions.
-Accuracy assessments of the photo-interpreted map product included field verification by random sample points, with a requirement of 90% accuracy for the seagrass categories.
+     For the subtidal layers, details were added starting on line 150: "These data include vector polygon coverages of the major subtidal habitats in Tampa Bay, as interpreted from 1:24,000 scale natural color aerial photographs flown in winter 2018 under cloud free conditions.
+Accuracy assessments of the photo-interpreted map included field verification by random sample points, with a requirement of 90% accuracy for the seagrass categories.
 The minimum mapping unit for seagrass polygons is reported as 0.25 acres."
@@
 * __Response__: Additional detail was provided as follows: "Specifically, the spatial datasets were unioned and the total areal change of the polygons for each habitat type was quantified by taking the difference between the two terminal years.
-For example, the area that remained as native uplands between the 1990 and 2017 intertidal and suptratidal layers was quantified, whereas the area that changed from native uplands to another habitat category was also quantified.
+For example, the area that remained as native uplands between the 1990 and 2017 intertidal and supratidal layers was quantified, whereas the area that changed from native uplands to another habitat category was also quantified.
 This process was repeated for all native habitats, including developed and restorable lands (described below)."
@@
 * __Response__: Thank you for the clarification and we agree that adopting a standardized and more widely accepted definition is more appropriate.
-However, the projects in the database have already been classified as restoration/enhancement and a distinction based on the previous definition is useful to local restoration practitioners that are familiar with the projects.
+However, the projects in the database have already been classified as restoration/enhancement based on the previous definition and this distinction is useful to local restoration practitioners that are familiar with the projects.
 For clarity, we have included the SER definition as a general description and amended the text to better justify the previous description.
@@
 * __Response__: Apologies for the confusion.
-The reservation areas are those in the coastal stratrum that are currently not in existing or proposed conservation areas.
+The reservation areas are those in the coastal stratum that are currently not in existing or proposed conservation areas.
 The sentence on line 211 was revised as follows:
@@
 
-* __Response__: Details on these data were added as follows: "The FNAI data are Florida Managed Areas as vector-based polygons of public and some private lands identified as having natural resource value and that are being managed at least partially for conservation.
-The source data for this layer are provided to FNAI directly from the managing agency in digital format or as paper source maps that are digitized using appropriate topographic quadrangles, ortho-imagery, and property appraiser parcel data at a minimum spatial resolution of 1:5,000."
+* __Response__: Details on these data were added as follows: "The FNAI data are Florida Managed Areas as vector polygons of public and some private lands identified as having natural resource value and that are being managed at least partially for conservation.
+The source data for this layer are provided to FNAI directly from the managing agency in digital format or as paper maps that are digitized using appropriate topographic quadrangles, ortho-imagery, and property appraiser parcel data at a minimum spatial resolution of 1:5,000."
 
@@
 Current discussions have focused on how to maintain the integrity of these data as additional work is completed in the future.
-The limitations discussed in the second paragraph of section 4.2 alludes to these challenges.
-A link to the online location of this database, that inclues a map, was added to line 188 in the methods (<https://www.tampabay.wateratlas.usf.edu/restoration/>).
+The limitations discussed in the second paragraph of section 4.3 alludes to these challenges.
+A link to the online location of this database, that includes a map, was added to line 188 in the methods (<https://www.tampabay.wateratlas.usf.edu/restoration/>).
 
@@
 
-* __Response__: We understand the general concern that have included explicit projections of how land use and cover may chagne with climate change and development.
-The methods we have used implicitly account for these factors and are our best estimate at what can be reasonably achieved by 2030 and 2050 based on the assessent.
+* __Response__: We understand the general concern that we have not included explicit projections of how land use and cover may change with climate change and development.
+The methods we have used implicitly account for these factors and are our best estimate at what can be reasonably achieved by 2030 and 2050 based on the assessment.
 We hope that revisions to some of the earlier comments have also made this more clear (e.g., lines 130 - 132).
-We have added the following text to this section for clarify:
+We have added the following text to this section for clarity:
 
@@
 
-* __Response__: The column header was changed to "Landcover Category".  The table caption was also changed.
+* __Response__: The column header was changed to "Land cover Category".  The table caption was also changed.
 
@@
 However, the restoration potential is general and does not provide specific recommendations on which native habitats to restore, i.e., tidal wetland is indicated and not mangrove forests, salt barrens, or salt marshes.
-Additional data would be needed to determine the appropriate type of tidal wetland and we leave that to the restoration practiioners to conduct site-level assessments when a restoration is planned.
-Also, our responses to the other reviewers address this concern regarding the true restoration potential of wetlands.
+Additional data would be needed to determine the appropriate type of tidal wetland and we leave that to the restoration practitioners to conduct site-level assessments for planned restorations.
+Also, our responses to the other reviewer addresses this concern regarding the true restoration potential of wetlands.
 
@@
 
-* __Response__:
+* __Response__: The targets and goals are explained by the rationale in the final column.
+These values are not inclusive of the total restoration opportunity, rather they are set by the expectation of what can be restored during these time periods based on past restoration effort and trends.
+For example, the coastal uplands target for 2030 is the current extent plus 61 hectares (1446 ha + 61 ha = 1507 ha) and the goal for 2050 is 1507 ha plus 61 ha increased by 20 each out decade (1507 ha + 81 ha + 101 ha = 1689 ha).
+The paragraph on line 324 was edited to make this more clear.
 

### 2023-01-03 manu-draft.qmd
finished first round of revisions and proofed

@@
 The new approach also accounts for the future effects of sea-level rise, climate change, and watershed development by prioritizing native coastal habitats relative to subtidal or upland areas.
-The restoration potential for coastal habitats on existing conservation lands was `r areas$potential[['Tidal Wetlands']]$areaha` for tidal wetlands and `r areas$potential[['Coastal Uplands']]$areaha` for coastal uplands.
-Maps were created to identify the restoration opportunities where practitioners could focus efforts to achieve the defined targets and goals.
-Methods for repeatable analyses are also available using an open source workflow.
+Maps were created to identify the restoration opportunities where practitioners could focus efforts to achieve the targets and goals, with methods for repeatable analyses also available using an open source workflow.
 
@@
 For subtidal habitats, 2018 data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19].
-These data include vector polygons coverages of the major subtidal habitat types in Tampa Bay, as interpreted from 1:24,000 scale natural color aerial photographs flown in winter 2018 under cloud free conditions.
-Accuracy assessments of the photo-interpreted map product included field verification by random sample points, with a requirement of 90% accuracy for the seagrass categories.
+These data include vector polygon coverages of the major subtidal habitats in Tampa Bay, as interpreted from 1:24,000 scale natural color aerial photographs flown in winter 2018 under cloud free conditions.
+Accuracy assessments of the photo-interpreted map included field verification by random sample points, with a requirement of 90% accuracy for the seagrass categories.
 The minimum mapping unit for seagrass polygons is reported as 0.25 acres.
@@
 Specifically, the spatial datasets were unioned and the total areal change of the polygons for each habitat type was quantified by taking the difference between the two terminal years.
-For example, the area that remained as native uplands between the 1990 and 2017 intertidal and suptratidal layers was quantified, whereas the area that changed from native uplands to another habitat category was also quantified.
+For example, the area that remained as native uplands between the 1990 and 2017 intertidal and supratidal layers was quantified, whereas the area that changed from native uplands to another habitat category was also quantified.
 This process was repeated for all native habitats, including developed and restorable lands (described below).
@@
 To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory (FNAI) and permit databases of conservation and drainage easements [@FNAI20].
-The FNAI data are Florida Managed Areas as vector-based polygons of public and some private lands identified as having natural resource value and that are being managed at least partially for conservation.
-The source data for this layer are provided to FNAI directly from the managing agency in digital format or as paper source maps that are digitized using appropriate topographic quadrangles, ortho-imagery, and property appraiser parcel data at a minimum spatial resolution of 1:5,000.
+The FNAI data are Florida Managed Areas as vector polygons of public and some private lands identified as having natural resource value and that are being managed at least partially for conservation.
+The source data for this layer are provided to FNAI directly from the managing agency in digital format or as paper maps that are digitized using appropriate topographic quadrangles, ortho-imagery, and property appraiser parcel data at a minimum spatial resolution of 1:5,000.
 Intersecting the native and restorable lands in these areas produced four unique opportunity areas: existing conservation native, proposed conservation native, existing conservation restorable, and proposed conservation restorable.
@@
 Table 5 shows the targets and goals identified through this analysis and the associated rationale.
-For example, the targets and goals are established based on the current extent and the restoration potential.
-If restoration potential exists and coverage restored from past projects suggests a reasonable level of effort, the targets and goals reflect the current extent plus the restoration opportunity.
+For example, the targets and goals are established based on the current extent and informed by the restoration potential.
+If restoration potential exists and coverage restored from past projects suggests a reasonable level of effort, the targets and goals reflect the current extent relative to the restoration opportunity, past trends, and anticipated effort.
 Conversely, other habitats with no identified restoration opportunity, or with sufficient current extents (e.g., mangrove forests), were assigned targets and goals similar to the current extent, i.e., these habitats should be protected and further restoration will only increase resilience.
+The proposed targets and goals do not represent the current extent plus restoration potential for these reasons.
 Implicit in the targets and the goals is recurring re-assessment over time to evaluate progress and adjust expectations as appropriate.
@@
 ```{r}
-#| tbl-cap: "Table 3: Change over time in hectares for intertidal and supratidal landcover in Tampa Bay.
+#| tbl-cap: "Table 3: Change over time in hectares for intertidal and supratidal land cover in Tampa Bay.
 Columns show years with available data and the final two columns show the change and percent change from 1990 to 2017."

### 2023-01-03 R/tabs.R

@@ -89,31 +89,31 @@ gaincol <- 'green'
 # format for the table
 totab <- acres %>%
   rename(
-    `Landcover Category` = HMPU_TARGETS,
+    `Land Cover Category` = HMPU_TARGETS,
     yr = name,
     Hectares = Acres
   ) %>%
   filter(yr <= 2017) %>%
   mutate(
     Stratum = case_when(
-      `Landcover Category` %in% c('Mangrove Forests', 'Salt Barrens', 'Salt Marshes') ~ 'Intertidal',
+      `Land Cover Category` %in% c('Mangrove Forests', 'Salt Barrens', 'Salt Marshes') ~ 'Intertidal',
       T ~ 'Supratidal'
     ),
     Hectares = Hectares / 2.471
   ) %>%
   tidyr::pivot_wider(names_from = yr, values_from = Hectares) %>%
-  filter(`Landcover Category` != 'Open Water') %>%
+  filter(`Land Cover Category` != 'Open Water') %>%
   mutate(
     `1990 to 2017` = `2017` - `1990`,
     `% change` = 100 * (`2017` - `1990`) / `1990`,
     chgcol = case_when(
-      `% change` < 0 & `Landcover Category` != 'Developed' ~ 0,
-      `% change` >= 0 & `Landcover Category` != 'Developed' ~ 1,
-      `% change` < 0 & `Landcover Category` == 'Developed' ~ 1,
-      `% change` >= 0 & `Landcover Category` == 'Developed' ~ 0,
+      `% change` < 0 & `Land Cover Category` != 'Developed' ~ 0,
+      `% change` >= 0 & `Land Cover Category` != 'Developed' ~ 1,
+      `% change` < 0 & `Land Cover Category` == 'Developed' ~ 1,
+      `% change` >= 0 & `Land Cover Category` == 'Developed' ~ 0,
     )
   ) %>%
-  arrange(Stratum, `Landcover Category`) %>%
+  arrange(Stratum, `Land Cover Category`) %>%
   as_grouped_data(groups = 'Stratum')
 
 thm <- function(x){

### 2023-01-03 revision/review_response.qmd
resubmitted first revision

@@
      Second, the following was added to the methods on line 197: "The identification of these areas on a broad spatial scale serves as a planning tool for restoration practitioners, where follow-up assessments are expected to more fully quantify restoration potential at selected sites."
-"
+
 Figure 1. The classification criteria for land use types are confused. I suggest using the Ramsar Convention classification system.

### 2023-01-09 manu-draft.qmd
second revision done

@@
 
-*Key words*: Florida, land use change, restoration planning, sea-level rise, Tampa Bay, urbanization
+*Key words*: Florida, habitat loss, land use change, sea-level rise, Tampa Bay, urbanization
 
@@
 The general approach includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in coverage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories.
-These opportunity areas provide a first assessment of where restoration could occur, where on the ground assessments could be pursued to further quantify restoration potential.
+These opportunity areas provide a first assessment of where restoration could occur and where on the ground assessments could be pursued to further quantify restoration potential.
 The outcomes of the approach are also spatially specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.
@@
 Accuracy assessments of the photo-interpreted map included field verification by random sample points, with a requirement of 90% accuracy for the seagrass categories.
-The minimum mapping unit for seagrass polygons is reported as 0.25 acres.
+The minimum mapping unit for seagrass polygons is reported as 0.25 acres (~0.1 ha).
 Historical datasets for subtidal habitats using identical methods began in 1988 with updates occurring on an approximate biennial basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.
@@
 This required an intersection of the data layers to quantify if habitat types were unchanged or changed for any given location and identifying the type of change (e.g., seagrass to tidal flats).
-Specifically, the spatial datasets were unioned and the total areal change of the polygons for each habitat type was quantified by taking the difference between the two terminal years.
+Specifically, the spatial datasets were unioned and the total areal change of the polygons for each habitat type was quantified by calculating the difference between the two terminal years.
 For example, the area that remained as native uplands between the 1990 and 2017 intertidal and supratidal layers was quantified, whereas the area that changed from native uplands to another habitat category was also quantified.
@@
 All FLUCCS classification codes were placed into one of three categories.
-First, native habitats were those that included the full range of natural plant communities and other habitats that are endemic to the watershed Second, restorable habitats included existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through restoration.
+First, native habitats were those that included the full range of natural plant communities and other habitats that are endemic to the watershed.
+Second, restorable habitats included existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through restoration.
 Third, existing development included developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration.
@@
 Oyster bars were estimated at 67 ha in 2018 (Table 2), showing a 29% increase since mapping began in 2014.
-This increase may represent improved ground-truthing and photointerpretation, rather than a true increase in coverage.
+The increase in oyster bars may represent improved ground-truthing and photointerpretation.
 Tidal flats have generally increased from 1988 to the mid-2000s, followed by a decrease to present.
@@
 Mangrove expansion as a result of climate change has been observed throughout the Gulf of Mexico [@Comeaux12; @Osland22].
-Anthropogenic water withdrawals have also reduced freshwater flows reaching tidal marshes, contributing to reductions in coverage of key species (e.g., *Juncus roemerianus*) in place of mangroves [@Raabe12].
+Anthropogenic water withdrawals have also reduced freshwater flows reaching tidal marshes, contributing to reductions in coverage of key species (e.g., *Juncus roemerianus*) that have favored mangroves [@Raabe12].
 As such, the identified targets and goals for mangroves indicate protection of these habitats, without the need for additional restoration.

### 2023-01-09 R/dat_proc.R

@@ -26,17 +26,17 @@ trgsmetric <- structure(
       "Protect existing artificial reefs; enhance habitat complexity where feasible; expand reef area to promote fish and wildlife benefits",
       "Identify and protect existing tidal flats; assess restoration potential of other non-vegetated subtidal areas",
       "Protect existing seagrasses; assess restoration potential of non-vegetated subtidal areas",
-      "2030: Protect existing oysters and restore 20 hectares; increase target by 20 hectares each out-decade",
-      "2030: Construct 1.6 kilometers each year; better define opportunity areas; increase target to 2.4 and 3.2 kilometers per year for out decades",
-      "2030: Protect existing intertidal mosaic and restore 405 hectares; increase target by 61 hectares each out-decade; includes the mosaic of mangrove, salt barren, and salt marsh habitats",
+      "2030: Protect existing oysters and restore 20 hectares; increase target by 20 hectares each decade",
+      "2030: Construct 1.6 kilometers each year; better define opportunity areas; increase target to 2.4 and 3.2 kilometers per year for each decade",
+      "2030: Protect existing intertidal mosaic and restore 405 hectares; increase target by 61 hectares each decade; includes the mosaic of mangrove, salt barren, and salt marsh habitats",
       "Protect existing mangrove forests; restore opportunistically within the intertidal mosaic",
-      "2030: Protect existing salt barrens and restore 20 hectares; increase target by 20 hectares per out decade",
-      "2030: Protect existing low salinity salt marshes and restore 101 hectares; increase target by 20 hectares each out-decade",
-      "Inventory mapped tidal tributaries and identify restoration potential; restore 6.4 kilometers of urban tidal creek habitat where feasible; increase target by 3.2 kilometers per out decade",
-      "2030: Protect existing coastal uplands and restore 61 hectares; increase target by 20 hectares each out decade",
-      "2030: Protect existing non-forested freshwater wetlands and restore 546 hectares; increase target by 20 hectares each out decade",
-      "2030: Protect existing forested freshwater wetlands and restore 61 hectares; increase target by 20 hectares each out decade",
-      "2030: Protect existing native uplands and restore 182 hectares; increase target by 20 hectares each out decade; focus on pine flatwoods and protect current extent"
+      "2030: Protect existing salt barrens and restore 20 hectares; increase target by 20 hectares each decade",
+      "2030: Protect existing low salinity salt marshes and restore 101 hectares; increase target by 20 hectares each decade",
[…]

### 2023-01-09 R/tabs.R

@@ -21,8 +21,8 @@ totab <- tibble(
 )
 
 thm <- function(x){
-  flextable::width(x, width = 1.75, j = 1) %>%
-    flextable::width(width = 4.75, j = 2) %>%
+  flextable::width(x, width = 2.25, j = 1) %>%
+    flextable::width(width = 4.25, j = 2) %>%
     # flextable::width(width = 2.75, j = 7) %>%
     fontsize(size = 8, part = 'all') %>%
     flextable::padding(padding = 0, part = 'all')
